---
title: "Bifactor, higher-order, and testlet structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor, higher-order, and testlet structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette is a small tutorial paper on estimating structures that attach
an "extra" factor to a set of group factors: the orthogonal **bifactor**
model, the **higher-order** (HO) model, and the **testlet** model. It answers
four questions in order, with every claim demonstrated on simulated data:

1. How is an orthogonal bifactor fit with `vbfa()`, and how well does it
   recover a true bifactor structure? (Section 2)
2. Can a `pefa()` sweep determine the number of *group* factors in a
   bifactor design? (Sections 3 and 5: **no** -- and the reason is an
   identification property worth understanding, not a tuning problem.)
3. How are higher-order and testlet structures simulated, fit, checked, and
   effect-sized? (Section 4)
4. Given data of unknown structure, how should one *decide* between
   correlated factors, a bifactor, and a higher-order model? (Section 6: a
   two-step route that each of the earlier sections feeds into.)

Section 1 first lays out the three models and the small amount of covariance
geometry that explains everything the later sections observe.

## 1. Three structures, and where the information lives

Throughout, `J` items fall into `K` clusters (item `j`'s cluster is `k(j)`),
and all models are for continuous responses with uncorrelated residuals.

**Correlated factors (oblique).** Each item loads on its cluster factor only
(loading $a_j$); the factors correlate through $\Phi$. The model-implied
covariance between items in *different* clusters $k$ and $l$ is
$\phi_{kl}\, a_i a_j$; between items in the *same* cluster it is $a_i a_j$.

**Orthogonal bifactor.** Every item loads on one general factor ($b^{gen}_j$)
and on its cluster's group factor ($b^{grp}_j$); all factors are mutually
orthogonal. Cross-cluster covariance is $b^{gen}_i b^{gen}_j$; within-cluster
covariance is $b^{gen}_i b^{gen}_j + b^{grp}_i b^{grp}_j$.

**Higher-order (HO).** Items load on first-order factors ($\lambda_j$), which
load on a single second-order factor ($\gamma_k$, one per first-order
factor). The HO model is mathematically equivalent to an orthogonal bifactor
whose loadings obey the Schmid-Leiman (SL) proportionality constraint
(Schmid & Leiman, 1957; Yung, Thissen, & McLeod, 1999): for every item $j$
in cluster $k$,

$$b^{gen}_j = \lambda_j \gamma_k \qquad\qquad b^{grp}_j = \lambda_j \sqrt{1 - \gamma_k^2}.$$

So within each cluster the ratio $b^{grp}_j / b^{gen}_j =
\sqrt{1-\gamma_k^2}/\gamma_k$ is **constant** across items -- that constancy
is the testable signature of an HO model hiding inside a bifactor fit. The
**testlet model is mathematically equivalent to the higher-order model**:
the same SL-constrained bifactor, read with testlet or method-effect labels
in place of first-order-factor labels (Zhang & Chen, 2024).

Two geometric facts about these covariance structures drive all the
empirical behaviour below.

**Fact 1: cross-cluster blocks carry no information about a general
factor.** Both the oblique model ($\phi_{kl}\, a_i a_j$) and the bifactor
($b^{gen}_i b^{gen}_j$) imply *rank-one* cross-cluster blocks. Between-cluster
covariance therefore cannot distinguish "one general factor" from
"correlated group factors" -- ever, at any sample size.

**Fact 2: the general factor is identified only by within-cluster rank.**
The oblique model implies rank-*one* within-cluster common covariance
($a_i a_j$); the bifactor implies rank-*two*
($b^{gen}_i b^{gen}_j + b^{grp}_i b^{grp}_j$). A general factor is
empirically distinguishable from factor correlations exactly to the extent
that the within-cluster blocks are genuinely rank-two -- which requires the
general and group loadings to vary *independently* across the items of a
cluster. Two corners are degenerate: under exact SL proportionality the
rank-two block collapses to effective rank one (the HO model *equals* an
oblique model with $\Phi = \gamma\gamma'$ off the diagonal, so the bifactor
parameterization is redundant); and when loadings are near-homogeneous
within clusters, the two rank-one directions nearly coincide, so the models
are close to empirically equivalent at realistic sample sizes. Section 6
shows both corners in data.

## 2. Fitting an orthogonal bifactor

`orthogonal = TRUE` fixes the factor correlation matrix at the identity.
With a design whose first column is a general factor (every item loads) plus
group columns, this is an identified orthogonal bifactor model.

The population below is a *genuine* bifactor in the sense of Fact 2: the
general and group loadings are drawn independently and span a wide range, so
the general-to-group ratio varies haphazardly within clusters and no
higher-order representation exists (contrast Section 4, where the loadings
are built to satisfy exactly the constant-ratio constraint).

```{r bifactor-simulate}
K <- 3; ipf <- 6; J <- K * ipf
cl <- rep(1:K, each = ipf)
set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)                 # general
for (k in 1:K) mla[cl == k, k + 1] <- round(runif(ipf, .35, .70), 2)  # group
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)  # sane communalities
sim <- sim_fa(N = 1000, mla = mla, phi = 0, rseed = 1)   # phi = 0 -> orthogonal
Y <- sim$dat
```

The fit is partially confirmatory: the general column is fully specified
(every item loads), two items per group factor are anchored, and the
remaining group entries are left to the data.

```{r bifactor-fit}
Q <- matrix(-1L, J, K)
for (k in 1:K) {
  a <- which(cl == k)[1:2]
  Q[a, ] <- 0L
  Q[a, k] <- 1L
}
Qb <- cbind(1L, Q)
fb <- vbfa(Y, Qb, orthogonal = TRUE)
fb$Phi                 # identity by construction, not estimated
round(fit_stats(fb), 3)
```

`fit_stats()` reads `orthogonal` (and `ld`) from the fit object itself, so a
bifactor fit needs no extra argument at the `fit_stats()` call -- and
supplying a contradictory one is an error, not a silent miscount.

A check that the fit tracks the truth, on the hard part -- the
general/group split:

```{r bifactor-verify}
c(cor = cor(fb$Lam[, 1], sim$MLA[, 1]),
  MAE_general = mean(abs(fb$Lam[, 1] - sim$MLA[, 1])))
```

With heterogeneous loadings the split is well identified and the anchored
fit recovers it: the general loadings correlate near `.98` with the truth,
with a mean absolute error under `.05`. This anchored refit -- not any sweep
-- is the estimation workhorse that everything later relies on.

## 3. Why a sweep cannot count group factors

The gain-rule sweep from `pefa()` accepts a bifactor backbone: keep the
general column fixed and specified, anchor two items per group factor, and
let `pefa()` pad the rest of the window with exploratory group columns.

`Kmin` and `Kmax` count **every** column of the design, the general factor
included -- not just the group factors. The backbone below has a general
column plus two anchored group factors, so `K0 = 3` specified columns and
`Kmin = 3` means "general + 2 group factors." The `K` reported in the sweep,
and the `K` the selection rule picks, is that same total factor count
(general factor and all), not the number of group factors.

```{r pefa-sweep}
## backbone: general column (every item loads) + 2 anchored group columns
Q0 <- matrix(-1L, J, 3)
Q0[, 1] <- 1L                              # general factor
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0[a, 2:3] <- 0L
  Q0[a, k + 1] <- 1L
}

r <- pefa(Q0, Y, Kmin = 3, Kmax = 6, orthogonal = TRUE, v0 = .001,
          verbose = FALSE)
summary(r)$comparison[, c("K", "ELBO", "BIC", "RMSEA", "CFI")]
r
```

The true design has a general factor plus 3 group factors, i.e. `K = 4`
total columns -- and the sweep does **not** find it: the primary
(objective-gain) rule settles at `K = 3`, the *lower* edge of the window,
while the fit indices sit near their ceilings at every `K` and the BIC
differences are trivial relative to their scale. This outcome is worth
understanding rather than tuning away, because it follows directly from the
geometry of Section 1. By Fact 1, the omitted group factor's cross-cluster
signature does not exist; by Fact 2, its within-cluster rank-two block is
the only evidence -- and the unspecified (`-1`) entries of the columns that
*remain* supply enough free directions to reproduce it almost exactly. Look
at what the `K = 3` candidate did with the third cluster's items:

```{r pefa-absorb}
f3 <- r$fits[[1]]
round(f3$Lam[cl == 3, ], 2)   # columns: general, group 1, group 2
```

Their group variance has been folded into inflated general-factor loadings
(compare `mla[cl == 3, 1]`), leaving almost nothing for a fourth factor to
explain -- which is why the objective *falls* at every larger `K` (more
spike-and-slab bookkeeping, no new fit) and why the covariance indices
cannot tell the candidates apart. Under a different regularization or
anchoring pattern the same surplus reappears in a different disguise (split
across the two anchored group columns, say), but some absorbed,
covariance-equivalent representation is always available.

The blunt conclusion first: **an ordinary bifactor sweep -- general column
plus anchored group columns, everything else `-1` -- should not be used to
determine the number of group factors.** Its selection is not merely noisy
but systematically misleading: the absorbed candidate wins regardless of the
true `K`, and (Section 5) the truth-sized candidate inside the sweep is
itself rotationally scrambled, so its loadings cannot be inspected to detect
the problem. `print(r)` therefore flags an orthogonal sweep that lands on
the lower boundary as *weak evidence about K* rather than issuing the
generic advice to extend the window.

Two practical rules follow. First, a `boundary = "lower"` stall in a
bifactor sweep means "no *separable* extra group factor was found," never
"the backbone is the true structure." Second, the absorption route runs
through entries that are free but should not be: if an item is *known* not
to measure a group factor, encode that as a fixed `0` rather than an
exploratory `-1` -- that restores an interior selection for the factor
*count* (though still not interpretable loadings; see Section 5). And when
the real question is "how many factors are in these data at all," Section 6
shows the sweep that *does* answer it: the ordinary correlated-factor sweep,
with no general column.

(`v0 = .001` -- a single fixed spike rather than the default warm-started
path -- keeps a `Kmin:Kmax` sweep fast. The absorption behaviour above is
not an artifact of that choice; the default path lands in the same place.)

## 4. Higher-order and testlet structures

### (a) Simulate a true higher-order model directly

`sim_fa()`'s `gamma` argument generates directly from the SL equivalence of
Section 1: supply the second-order loadings and `sim_fa()` builds the SL
bifactor matrix internally and simulates from it, with the factor
correlation matrix `PHI` fixed at the identity.

```{r ho-simulate}
gamma_true <- c(.80, .70, .60)
sim_ho <- sim_fa(N = 3000, K = 3, ipf = 6, lam = .8, gamma = gamma_true, rseed = 11)
Yh <- sim_ho$dat
round(sim_ho$MLA, 2)    # the SL bifactor matrix sim_fa built internally
```

The sample is deliberately large. The quantity subsection (c) inspects is a
*ratio* of two fitted loadings, and -- by Fact 2 -- the general-versus-group
split behind that ratio is the softest part of an orthogonal bifactor
solution, so its sampling noise is large relative to the loadings
themselves. `N = 3000` keeps that noise small enough for the constant-ratio
signature to show through cleanly.

### (b) Fit a partially confirmatory orthogonal bifactor

The general column is fully specified; two items per group factor are
anchored, and the rest of the group loadings -- including the off-cluster
entries -- are left to the data.

```{r ho-fit}
Qg <- matrix(-1L, J, K)
for (k in 1:K) {
  a <- which(cl == k)[1:2]
  Qg[a, ] <- 0L
  Qg[a, k] <- 1L
}
Qh <- cbind(1L, Qg)

fit <- vbfa(Yh, Qh, orthogonal = TRUE)
fit
```

### (c) Check the SL constraint on the fitted loadings

A few lines of plain R -- deliberately not a package function, since this is
a descriptive diagnostic, not a fitted model. Per cluster, compute each
item's ratio of its own group loading to the general loading,
$r_j = |b^{grp}_j| / |b^{gen}_j|$, then that ratio's mean and coefficient of
variation ($CV = sd/mean$) within the cluster. A small CV means the ratios
are nearly constant across items in that cluster -- the SL-admissible
signature of a higher-order model hiding inside the bifactor fit. Averaging
the cluster CVs across all items (item-weighted, so the code generalizes to
unequal cluster sizes) gives one overall number, and the rule applied here
is:

**mean CV < .1 means the fitted bifactor is approximately a higher-order
model.**

```{r ho-sl-check}
b_gen <- fit$Lam[, 1]
cl_tab <- t(sapply(1:K, function(k) {
  idx <- which(cl == k)
  r <- abs(fit$Lam[idx, k + 1]) / abs(b_gen[idx])
  c(n = length(idx), mean = mean(r), sd = sd(r), cv = sd(r) / mean(r))
}))
rownames(cl_tab) <- paste0("cluster", 1:K)
round(cl_tab, 3)

mean_cv <- sum(cl_tab[, "n"] * cl_tab[, "cv"]) / sum(cl_tab[, "n"])
mean_cv
```

The mean CV comes out at about `.084` -- under the `.1` line, so the fitted
bifactor reads as approximately a higher-order model, exactly as it should:
the data were generated under one. The selection side of the fit is equally
clean here: every unanchored item's own-cluster loading is selected active
(posterior inclusion probability 1) and no off-cluster loading is. Note that
the pooled number passes while cluster 1 alone (CV about `.13`) sits
slightly above the line -- with only six items per cluster the per-cluster
CV is a noisy statistic, which is why the rule is stated on the
item-weighted mean. The per-cluster values are still worth a glance: a
*single* cluster far above the line with the others near zero would tell a
different story (one non-HO-like cluster) than the same mean produced
evenly.

### (d) Transform back to higher-order parameters

Given the SL relations, $\gamma_k$ and $\lambda_j$ can be recovered from
each cluster's mean ratio $\bar r_k$: $\hat\gamma_k = 1/\sqrt{1 + \bar
r_k^2}$ and $\hat\lambda_j = |b^{gen}_j| / \hat\gamma_{k(j)}$.

```{r ho-transform}
gamma_hat <- 1 / sqrt(1 + cl_tab[, "mean"]^2)
names(gamma_hat) <- paste0("cluster", 1:K)
rbind(true = gamma_true, hat = round(gamma_hat, 3))

lambda_true <- sim_ho$MLA1[cbind(seq_len(J), cl)]   # true first-order loading of each item
lambda_hat <- numeric(J)
for (k in 1:K) {
  idx <- which(cl == k)
  lambda_hat[idx] <- abs(b_gen[idx]) / gamma_hat[k]
}
c(worst = max(abs(lambda_hat - lambda_true)),   # first-order loading errors
  mean  = mean(abs(lambda_hat - lambda_true)))
```

The second-order loadings come back at about `.81`, `.69`, and `.59`
against the true `.80`, `.70`, `.60` -- each within about `.013` of the
truth. The first-order loadings (`.8` throughout, since `lam = .8` is
constant here) are recovered with a worst-case error of about `.06` and a
mean error near `.02`. When the SL constraint genuinely holds and the CV
check passes, the transform-back step is this accurate: the mean ratio per
cluster is all the information the second-order structure needs.

### (e) Group-factor effect size (Zhang & Chen, 2024, Eq. 16)

Whether the model is read as a bifactor, an HO model, or a testlet model,
the group/testlet factor's effect size quantifies how strongly its items
load on it. Following Eq. 16 of Zhang and Chen (2024, p. 11), the effect
size of special (group/testlet) factor $k$ is the **average squared
loading** on that factor,

$$D_k = \frac{\sum (\text{squared specified loadings}) + \sum (\text{squared active unspecified loadings})}{\text{number of loadings included}},$$

where an unspecified loading counts as active only if it passes a cutoff.
The package is a spike-and-slab framework rather than a Lasso one, so the
natural cutoff here is the posterior inclusion probability: an unspecified
entry (`Q == -1`) counts as active when `fit$pi > tau` with `tau = .5`;
specified entries (`Q == 1`) always count.

```{r ho-D}
tau <- .5
Dk <- sapply(1:K, function(k) {
  col <- k + 1
  included <- (Qh[, col] == 1) | (Qh[, col] == -1 & fit$pi[, col] > tau)
  sum(fit$Lam[included, col]^2) / sum(included)
})
names(Dk) <- paste0("cluster", 1:K)
round(Dk, 3)
```

The effect sizes come out at about `.22`, `.34`, and `.42`, against
population values of $(\lambda\sqrt{1-\gamma_k^2})^2 = .23, .33, .41$ -- an
essentially exact, and correctly ordered, recovery. The ordering is the
substantive reading: the *weaker* a cluster's second-order loading, the
*larger* its group/testlet effect, because whatever the general factor does
not explain of an item's common variance is exactly what its group factor
carries. Cluster 3 (true $\gamma = .60$) therefore shows the largest special
effect and cluster 1 (true $\gamma = .80$) the smallest.

## 5. Can a sweep recover the higher-order structure? No

Section 3 showed the absorption problem on genuine bifactor data; the same
question is sharper here, where the population is *exactly* a higher-order
model: hand the ordinary bifactor backbone (general column, two anchored
group factors, all else `-1`) to `pefa()` on `Yh` and ask whether it finds
the true `K = 4`.

```{r ho-sweep}
Q0h <- matrix(-1L, J, 3)
Q0h[, 1] <- 1L
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0h[a, 2:3] <- 0L
  Q0h[a, k + 1] <- 1L
}
rh <- pefa(Q0h, Yh, Kmin = 3, Kmax = 6, orthogonal = TRUE, v0 = .001,
           verbose = FALSE)
rh
```

It does not: the selection stalls at the lower boundary `K = 3`, with the
fit indices at their ceilings for every candidate, for exactly the
absorption reason -- SL-proportionality buys nothing, because the omitted
group factor's rank-one block is reproducible by the free entries either
way. Repeating this at several seeds and at `N = 500` gives the same lower
boundary selection every time; this is an identification property, not bad
luck. Worse, the truth-sized `K = 4` candidate *inside* the sweep is not
interpretable either. Its padded column is never anchored, so nothing pins
the general/group allocation: here the extra column simply stays empty and
the absorbed representation persists (cluster 3's group variance sits in
inflated general loadings plus off-cluster spill), while under other
settings and seeds the allocation comes out rotated across columns instead.
Either way, these loadings must not be read, checked for SL
proportionality, or transformed back:

```{r ho-sweep-k4}
f4 <- rh$fits[[which(rh$sweep$K == 4)]]
round(f4$Lam[cl == 3, ], 2)     # cluster-3 rows: compare sim_ho$MLA[cl == 3, ]
```

All interpretation must instead happen on an *anchored refit* at the chosen
`K` -- the design of Section 4(b), whose fitted columns match the true SL
matrix with congruence above `.99`. Anchors pin the rotation; a sweep's
padded columns, by construction, never have them. That leaves the question
Section 6 answers: if sweeps cannot decide the bifactor question, what can?

## 6. Which structure? A two-step decision route

Sections 3 and 5 established what sweeps over *bifactor* candidates cannot
do. This section assembles what the tools *can* do into a defensible
procedure for the practical situation: data in hand, clusters suspected,
structure -- correlated factors, bifactor, or higher-order -- unknown.

### Step 1: an ordinary sweep for the first-order K

Run the plain correlated-factor sweep of the `pefa` vignette -- no general
column, no `orthogonal = TRUE`. This asks only: *how many cluster-level
factors are there?* For that question an omitted factor produces honest
misfit (its within-cluster block is not reproducible by other clusters'
columns), so interior selections are trustworthy. Run it on both datasets
of this vignette -- the genuine bifactor `Y` of Section 2 and a fresh
higher-order sample at the same `N`:

```{r step1}
sim_ho2 <- sim_fa(N = 1000, K = 3, ipf = 6, lam = .8, gamma = gamma_true,
                  rseed = 11)
Yh2 <- sim_ho2$dat

Q0_ob <- matrix(-1L, J, 2)
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0_ob[a, ] <- 0L
  Q0_ob[a, k] <- 1L
}
s_bif <- pefa(Q0_ob, Y,   Kmin = 2, Kmax = 6, verbose = FALSE)
s_ho  <- pefa(Q0_ob, Yh2, Kmin = 2, Kmax = 6, verbose = FALSE)
rbind(bifactor_data = c(K = s_bif$selected_K, boundary = s_bif$boundary),
      HO_data       = c(K = s_ho$selected_K,  boundary = s_ho$boundary))
```

Both land at `K = 3`, *interior*, and both are correct in first-order
terms: the HO data genuinely are three correlated factors (Section 1), and
the bifactor data have three clusters whose general factor the oblique
model absorbs into the factor correlations (Fact 1 -- cross-cluster blocks
cannot tell the difference). Note what step 1 does **not** claim: it says
nothing about whether a general factor should be carved out. That is step
2's question.

### Step 2: anchored oblique-K versus anchored bifactor-(K+1), by BIC

At the selected `K`, fit two *anchored, confirmatory-leaning* models with
the same anchors: the oblique `K`-factor model, and the orthogonal bifactor
with one more column -- the same design plus a fully specified general
factor. Compare their `BIC` from `fit_stats()`:

```{r step2}
Q3 <- matrix(-1L, J, 3)
for (k in 1:3) {
  a <- which(cl == k)[1:2]
  Q3[a, ] <- 0L
  Q3[a, k] <- 1L
}
Q4 <- cbind(1L, Q3)

step2 <- function(Y) {
  b3 <- fit_stats(vbfa(Y, Q3))["BIC"]
  b4 <- fit_stats(vbfa(Y, Q4, orthogonal = TRUE))["BIC"]
  c(oblique_K = unname(b3), bifactor_K1 = unname(b4), diff = unname(b4 - b3))
}
round(rbind(bifactor_data = step2(Y), HO_data = step2(Yh2)), 1)
```

The comparison resolves both datasets correctly at `N = 1000`: on the
genuine bifactor data the bifactor wins (`diff` about `-59`), and on the
higher-order data the oblique model wins (`diff` about `+103`) -- which is
the *right* verdict there, since by the SL equivalence the bifactor
parameterization of an HO model is redundant, and BIC correctly refuses to
pay 18 extra loadings for it.

Use **BIC, not the ELBO, as the arbiter in this comparison.** The two
designs place spike-and-slab prior mass on different numbers of entries, so
their ELBOs differ by prior bookkeeping as well as by fit; in simulations
the ELBO kept preferring the oblique model in regimes where BIC had already
(correctly) flipped to the bifactor. The ELBO is the right currency
*within* a sweep of same-shaped designs; BIC is the right currency between
these two differently-shaped ones.

### Step 3: resolve the winner into a final structure

**If the bifactor won**, run the SL check of Section 4(c) on the bifactor
fit. A mean CV under `.1` says the bifactor is SL-degenerate -- collapse it
to the more parsimonious HO model via the transform of Section 4(d). A
large mean CV says the general factor is genuinely its own thing:

```{r step3-bifactor}
f4b <- vbfa(Y, Q4, orthogonal = TRUE)
rt  <- sapply(1:J, function(j) abs(f4b$Lam[j, cl[j] + 1]) / abs(f4b$Lam[j, 1]))
cvs <- sapply(1:K, function(k) { r <- rt[cl == k]; sd(r) / mean(r) })
sum(table(cl) * cvs) / J          # item-weighted mean CV
```

Here it is about `.39` -- far above the line, so these data keep their full
bifactor: correctly so, since their generating loadings were drawn with no
proportionality constraint.

**If the oblique model won**, the general factor question is settled
(no separable general factor) -- but a higher-order *layer* may still be
available for free. Under an HO population, $\Phi$ has the rank-one
structure $\gamma\gamma'$ off the diagonal, so the second-order loadings
can be read out of the fitted factor correlations by the classical triad
(Spearman) relations, e.g. $\hat\gamma_1 =
\sqrt{\phi_{12}\phi_{13}/\phi_{23}}$:

```{r step3-oblique}
f3h <- vbfa(Yh2, Q3)
P <- f3h$Phi
round(P, 2)
gamma_triad <- c(sqrt(P[1, 2] * P[1, 3] / P[2, 3]),
                 sqrt(P[1, 2] * P[2, 3] / P[1, 3]),
                 sqrt(P[1, 3] * P[2, 3] / P[1, 2]))
round(gamma_triad, 2)             # true: .80 .70 .60
```

The triads recover the second-order loadings to within a few hundredths --
a second, independent route to the HO structure that never fits an
orthogonal model at all, and that can cross-validate the SL route of
Section 4 when both are applicable. (With more than three first-order
factors, replace the triads by a one-factor model fit to $\hat\Phi$; with
exactly three, the triads are that model's exact solution. Coherent triads
-- all inside the unit interval, stable across item subsets -- support the
HO reading; incoherent ones say $\Phi$ is just a correlation matrix, and
the structure is plain correlated factors.)

### The route, and its limits

In summary:

1. **Ordinary oblique sweep** (`pefa()` without a general column) for the
   first-order `K`. Trust interior selections; treat boundaries as
   instructions to extend the window.
2. **Anchored oblique-K vs anchored bifactor-(K+1)**, decided by `BIC`.
3. **Within the winner:** bifactor -> SL mean-CV check (< .1: collapse to
   HO; otherwise keep the bifactor). Oblique -> triad check of $\hat\Phi$
   (coherent: an HO layer is available; otherwise correlated factors).

Every population in this vignette routes correctly: the heterogeneous
bifactor of Section 2 goes sweep -> `K = 3` -> bifactor wins -> CV `.39`
-> **bifactor**; the higher-order data go sweep -> `K = 3` -> oblique wins
-> coherent triads (`.76/.72/.62`) -> **higher-order**.

The honest limits, from a 3-populations x 3-sample-sizes x 3-seeds
simulation with this design (18 items, 3 clusters):

| population | N = 500 | N = 1000 | N = 3000 |
|---|---|---|---|
| bifactor, heterogeneous loadings | bifactor wins 3/3 | bifactor 3/3 | bifactor 3/3 (margin grows) |
| bifactor, near-homogeneous loadings | oblique 3/3 | oblique 3/3 | bifactor 3/3 |
| higher-order | oblique 3/3 | oblique 3/3 | oblique 3/3 (margin flat) |

The middle row is Fact 2's degenerate corner: when general and group
loadings barely vary within clusters, a true bifactor is nearly empirically
equivalent to correlated factors, and step 2 only resolves it with a large
sample. Two mitigations: a *small* `|diff|` in step 2 is itself the warning
sign (the decisive wins above come with margins that grow with `N`, the
redundant HO comparison stays flat at roughly the parameter penalty); and
in that regime the parsimonious miss is cheap, because the two descriptions
of the data are nearly interchangeable by construction. What the route
never does is what Sections 3 and 5 warned against: asking a bifactor
*sweep* to make any of these calls.

## 7. Caveats

The SL check is descriptive, not a hypothesis test: there is no p-value
that certifies "this bifactor is really an HO model," and the
`mean CV < .1` rule is a working convention, not a distributional result.
Sampling error alone makes the CV strictly positive even when the data are
generated exactly under an HO model, as they are in Section 4 -- and the
loading *ratio* it is built on inherits the softness of the
general-versus-group split (Fact 2), which is why that section runs at a
generous `N`; at a few hundred observations the same population can land
above the line on noise alone, and an occasional item whose posterior
settles on the wrong group factor will inflate the pooled number further.
So read the mean CV together with the per-cluster (or per-item) ratios
rather than as a verdict on its own. The same softness bounds step 2 of
Section 6, as its table shows. And the diagnostics run both ways -- a large
mean CV, when it reflects genuinely non-constant ratios rather than noise
or an isolated item, is informative in its own right: it means the bifactor
solution is richer than any higher-order representation of it, and
collapsing it to a single second-order factor per cluster would discard
real structure.

## 8. References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical
  factor solutions. *Psychometrika*, 22(1), 53-61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113-128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models
  for special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5),
  208-229. \doi{10.1177/01466216241261704}
* Chen, J. (2021). A generalized partially confirmatory factor analysis
  framework with mixed Bayesian Lasso methods. *Multivariate Behavioral
  Research*, 57(6), 879-894. \doi{10.1080/00273171.2021.1925520}
* Jin, Y., & Chen, J. (2025). Regularized variational approximation for
  partially confirmatory factor analysis. *Structural Equation Modeling*,
  32(3), 437-449. \doi{10.1080/10705511.2024.2432612}
