---
title: "Bifactor, higher-order, and testlet structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor, higher-order, and testlet structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette collects three related structures that all attach "extra"
factors to a set of primary or group factors: the orthogonal **bifactor**
model, sweeping a **factor-number window** over bifactor candidates with
`pefa()`, and the connection between bifactor, **higher-order**, and
**testlet** parameterizations via the Schmid-Leiman transformation. All data
here are simulated; nothing is loaded from disk.

## 1. Orthogonal bifactor basics

`orthogonal = TRUE` fixes the factor correlation matrix at the identity. With
a design whose first column is a general factor (every item loads) plus group
columns, this is an identified orthogonal bifactor model.

The data below are built from a true bifactor loading matrix -- a general
factor plus three group factors, with the general and group loadings drawn
independently of one another -- so the general-to-group loading ratio varies
haphazardly across items within a cluster instead of holding constant. That
makes it a genuine bifactor structure, not a disguised higher-order model
(contrast with Section 3, where the loadings are built to satisfy exactly
that constant-ratio constraint).

```{r bifactor-simulate}
K <- 3; ipf <- 6; J <- K * ipf
cl <- rep(1:K, each = ipf)
set.seed(2)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .55, .70), 2)                 # general
for (k in 1:K) mla[cl == k, k + 1] <- round(runif(ipf, .45, .60), 2)  # group
sim <- sim_fa(N = 1000, mla = mla, phi = 0, rseed = 1)   # phi = 0 -> orthogonal
Y <- sim$dat
```

Fit is partially confirmatory: the general column is fully specified (every
item loads), two items per group factor are anchored, and the remaining group
entries are left to the data.

```{r bifactor-fit}
Q <- matrix(-1L, J, K)
for (k in 1:K) {
  a <- which(cl == k)[1:2]
  Q[a, ] <- 0L
  Q[a, k] <- 1L
}
Qb <- cbind(1L, Q)
fb <- vbfa(Y, Qb, orthogonal = TRUE)
fb$Phi                 # identity by construction, not estimated
round(fit_stats(fb), 3)
```

`fit_stats()` reads `orthogonal` (and `ld`) from the fit object itself, so a
bifactor fit needs no extra argument at the `fit_stats()` call -- and
supplying a contradictory one is an error, not a silent miscount.

A quick check that the fit tracks the truth: the correlation and mean
absolute error between the fitted and true general-factor loadings.

```{r bifactor-verify}
cor(fb$Lam[, 1], sim$MLA[, 1])
mean(abs(fb$Lam[, 1] - sim$MLA[, 1]))
```

The mean absolute error is about `.04`. The correlation (about `.85`) is the
less informative of the two numbers here: the true general loadings span only
`.55` to `.70`, and a correlation over so narrow a range is mostly measuring
noise ordering, not recovery. Splitting each item's common variance between a
general and a group factor is the genuinely hard part of bifactor estimation,
and an average error of `.04` per loading on that split is a tight result for
`N = 1000`.

## 2. A pefa sweep over bifactor candidates

The same gain-rule sweep from `pefa()` works with a bifactor backbone: keep the
general column fixed and specified, anchor two items per group factor, and let
`pefa()` pad the rest of the window with exploratory group columns.

`Kmin` and `Kmax` count **every** column of the design, the general factor
included -- not just the group factors. The backbone below has a general
column plus two anchored group factors, so `K0 = 3` specified columns and
`Kmin = 3` means "general + 2 group factors." The `K` reported in the sweep,
and the `K` the selection rule picks, is that same total factor count
(general factor and all), not the number of group factors.

```{r pefa-sweep}
## backbone: general column (every item loads) + 2 anchored group columns
Q0 <- matrix(-1L, J, 3)
Q0[, 1] <- 1L                              # general factor
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0[a, 2:3] <- 0L
  Q0[a, k + 1] <- 1L
}

r <- pefa(Q0, Y, Kmin = 3, Kmax = 6, orthogonal = TRUE, v0 = .001,
          verbose = FALSE)
summary(r)$selection
summary(r)$comparison[, c("K", "ELBO", "ELBO_gain", "BIC", "BIC_gain", "CFI")]
```

The true design has a general factor plus 3 group factors, i.e. `K = 4` total
columns -- and the sweep does **not** find it: the primary (objective-gain)
rule settles at `K = 3`, the *lower* edge of the window, while the fit
indices sit near their ceilings at every `K`. This outcome is worth
understanding rather than tuning away, because it is not sampling noise and
not an optimizer fluke. It is an identification fact about orthogonal
bifactor sweeps: **an omitted group factor can be absorbed, almost exactly,
by the unspecified entries of the columns that remain.** A cluster's extra
within-cluster covariance is a rank-one block, and free `-1` entries on the
general column and the other group columns give the model more than enough
directions to reproduce it without any new factor. Look at what the `K = 3`
candidate did with the third cluster's items:

```{r pefa-absorb}
f3 <- r$fits[[1]]
round(f3$Lam[cl == 3, ], 2)   # columns: general, group 1, group 2
```

Their group variance has been folded into inflated general-factor loadings
(compare `mla[cl == 3, 1]`, at most `.70`, with what is fitted here), leaving
almost nothing for a fourth factor to explain -- which is why the objective
*falls* at every larger `K` (more spike-and-slab bookkeeping, no new fit) and
why the covariance indices cannot tell the candidates apart. (The BIC-based
selections wander for the same reason -- with all candidates nearly
covariance-equivalent, they end up ranking bookkeeping differences rather
than structure.) Under a different regularization or anchoring pattern the
same surplus reappears in a different disguise (split across the two anchored
group columns, say), but some absorbed, covariance-equivalent representation
is always available.

The blunt conclusion first: **an ordinary bifactor sweep -- general column
plus anchored group columns, everything else `-1` -- should not be used to
determine the number of group factors.** Its selection is not merely noisy
but systematically misleading: the absorbed candidate wins regardless of the
true `K`, and (Section 3f below) the truth-sized candidate inside the sweep
is itself rotationally scrambled, so its loadings cannot be inspected to
detect the problem. `print(r)` therefore flags an orthogonal sweep that
lands on the lower boundary as *weak evidence about K* rather than issuing
the generic advice to extend the window.

Two practical rules follow. First, a `boundary = "lower"` stall in a
bifactor sweep means "no *separable* extra group factor was found," never
"the backbone is the true structure." Second, the absorption route runs
through entries that are free but should not be: if an item is *known* not
to measure a group factor, encode that as a fixed `0` rather than an
exploratory `-1` -- that restores an interior selection for the factor
*count* (though still not interpretable loadings; see Section 3f). The
clean, interior selections shown in the `pefa` vignette come from
correlated-factor structure, where an omitted factor produces honest misfit;
orthogonal bifactor structure is intrinsically the harder case for
factor-number discovery, and this section is deliberately an example of the
sweep telling you so.

(`v0 = .001` -- a single fixed spike rather than the default warm-started
path -- keeps a `Kmin:Kmax` sweep fast. The absorption behaviour above is not
an artifact of that choice; the default path lands in the same place.)

## 3. Higher-order CFA, Schmid-Leiman, and the bifactor connection

### (a) Simulate a true higher-order model directly

A higher-order (HO) model has items loading on first-order (group) factors
with loadings $\lambda_j$, which in turn load on a single second-order general
factor with loadings $\gamma_k$ (one per first-order factor $k$). This is
mathematically equivalent to an **orthogonal bifactor** model whose loadings
satisfy the Schmid-Leiman (SL) proportionality constraint: for every item $j$
in cluster $k$,

$$b^{general}_j = \lambda_j \gamma_k \qquad\qquad b^{group}_j = \lambda_j \sqrt{1 - \gamma_k^2}.$$

`sim_fa()`'s `gamma` argument generates directly from this equivalence:
supply the second-order loadings and `sim_fa()` builds the SL bifactor matrix
internally and simulates from it, with the factor correlation matrix `PHI`
fixed at the identity.

It is also worth being explicit that the **testlet model is mathematically
equivalent to the higher-order model** -- the same SL-constrained bifactor,
read with testlet or method-effect labels in place of first-order-factor
labels (Zhang & Chen, 2024).

```{r ho-simulate}
gamma_true <- c(.80, .70, .60)
sim_ho <- sim_fa(N = 3000, K = 3, ipf = 6, lam = .8, gamma = gamma_true, rseed = 11)
Yh <- sim_ho$dat
round(sim_ho$MLA, 2)    # the SL bifactor matrix sim_fa built internally
```

The sample is deliberately large. The quantity the next subsection inspects
is a *ratio* of two fitted loadings, and -- as Section 2 just showed -- the
general-versus-group split behind that ratio is the softest part of an
orthogonal bifactor solution, so its sampling noise is large relative to the
loadings themselves. `N = 3000` keeps that noise small enough for the
constant-ratio signature to show through cleanly.

### (b) Fit a partially confirmatory orthogonal bifactor

The general column is fully specified; two items per group factor are
anchored, and the rest of the group loadings -- including the off-cluster
entries -- are left to the data.

```{r ho-fit}
Qg <- matrix(-1L, J, K)
for (k in 1:K) {
  a <- which(cl == k)[1:2]
  Qg[a, ] <- 0L
  Qg[a, k] <- 1L
}
Qh <- cbind(1L, Qg)

fit <- vbfa(Yh, Qh, orthogonal = TRUE)
fit
```

### (c) Check the SL constraint on the fitted loadings

A few lines of plain R -- deliberately not a package function, since this is
a descriptive diagnostic, not a fitted model. Per cluster, compute each item's
ratio of its own group loading to the general loading,
$r_j = |b^{group}_j| / |b^{general}_j|$, then that ratio's mean and
coefficient of variation ($CV = sd/mean$) within the cluster. A small CV means
the ratios are nearly constant across items in that cluster -- the
SL-admissible signature of a higher-order model hiding inside the bifactor
fit. Averaging the cluster CVs across all items (item-weighted, so the code
generalizes to unequal cluster sizes) gives one overall number, and the rule
applied here is:

**mean CV < .1 means the fitted bifactor is approximately a higher-order
model.**

```{r ho-sl-check}
b_gen <- fit$Lam[, 1]
cl_tab <- t(sapply(1:K, function(k) {
  idx <- which(cl == k)
  r <- abs(fit$Lam[idx, k + 1]) / abs(b_gen[idx])
  c(n = length(idx), mean = mean(r), sd = sd(r), cv = sd(r) / mean(r))
}))
rownames(cl_tab) <- paste0("cluster", 1:K)
round(cl_tab, 3)

mean_cv <- sum(cl_tab[, "n"] * cl_tab[, "cv"]) / sum(cl_tab[, "n"])
mean_cv
```

The mean CV comes out at about `.084` -- under the `.1` line, so the fitted
bifactor reads as approximately a higher-order model, exactly as it should:
the data were generated under one. The selection side of the fit is equally
clean here: every unanchored item's own-cluster loading is selected active
(posterior inclusion probability 1) and no off-cluster loading is. Note that
the pooled number passes while cluster 1 alone (CV about `.13`) sits slightly
above the line -- with only six items per cluster the per-cluster CV is a
noisy statistic, which is why the rule is stated on the item-weighted mean.
The per-cluster values are still worth a glance: a *single* cluster far above
the line with the others near zero would tell a different story (one
non-HO-like cluster) than the same mean produced evenly.

### (d) Transform back to higher-order parameters

Given the SL relations, $\gamma_k$ and $\lambda_j$ can be recovered from each
cluster's mean ratio $\bar r_k$: $\hat\gamma_k = 1/\sqrt{1 + \bar r_k^2}$ and
$\hat\lambda_j = |b^{general}_j| / \hat\gamma_{k(j)}$.

```{r ho-transform}
gamma_hat <- 1 / sqrt(1 + cl_tab[, "mean"]^2)
names(gamma_hat) <- paste0("cluster", 1:K)
rbind(true = gamma_true, hat = round(gamma_hat, 3))

lambda_true <- sim_ho$MLA1[cbind(seq_len(J), cl)]   # true first-order loading of each item
lambda_hat <- numeric(J)
for (k in 1:K) {
  idx <- which(cl == k)
  lambda_hat[idx] <- abs(b_gen[idx]) / gamma_hat[k]
}
c(worst = max(abs(lambda_hat - lambda_true)),   # first-order loading errors
  mean  = mean(abs(lambda_hat - lambda_true)))
```

The second-order loadings come back at about `.81`, `.69`, and `.59` against
the true `.80`, `.70`, `.60` -- each within about `.013` of the truth. The
first-order loadings (`.8` throughout, since `lam = .8` is constant here) are
recovered with a worst-case error of about `.06` and a mean error near `.02`.
When the SL constraint genuinely holds and the CV check passes, the
transform-back step is this accurate: the mean ratio per cluster is all the
information the second-order structure needs.

### (e) Group-factor effect size (Zhang & Chen, 2024, Eq. 16)

Whether the model is read as a bifactor, an HO model, or a testlet model, the
group/testlet factor's effect size quantifies how strongly its items load on
it. Following Eq. 16 of Zhang and Chen (2024, p. 11), the effect size of
special (group/testlet) factor $k$ is the **average squared loading** on that
factor,

$$D_k = \frac{\sum (\text{squared specified loadings}) + \sum (\text{squared active unspecified loadings})}{\text{number of loadings included}},$$

where an unspecified loading counts as active only if it passes a cutoff. The
package is a spike-and-slab framework rather than a Lasso one, so the natural
cutoff here is the posterior inclusion probability: an unspecified entry
(`Q == -1`) counts as active when `fit$pi > tau` with `tau = .5`; specified
entries (`Q == 1`) always count.

```{r ho-D}
tau <- .5
Dk <- sapply(1:K, function(k) {
  col <- k + 1
  included <- (Qh[, col] == 1) | (Qh[, col] == -1 & fit$pi[, col] > tau)
  sum(fit$Lam[included, col]^2) / sum(included)
})
names(Dk) <- paste0("cluster", 1:K)
round(Dk, 3)
```

The effect sizes come out at about `.22`, `.34`, and `.42`, against
population values of $(\lambda\sqrt{1-\gamma_k^2})^2 = .23, .33, .41$ -- an
essentially exact, and correctly ordered, recovery. The ordering is the
substantive reading: the *weaker* a cluster's second-order loading, the
*larger* its group/testlet effect, because whatever the general factor does
not explain of an item's common variance is exactly what its group factor
carries. Cluster 3 (true $\gamma = .60$) therefore shows the largest special
effect and cluster 1 (true $\gamma = .80$) the smallest.

### (f) Can a sweep recover the higher-order structure? No -- and here is the fix

Section 2 showed the absorption problem on generic bifactor data; the same
question is sharper here, where the population is *exactly* a higher-order
model: hand the ordinary bifactor backbone (general column, two anchored
group factors, all else `-1`) to `pefa()` on `Yh` and ask whether it finds
the true `K = 4`.

```{r ho-sweep}
Q0h <- matrix(-1L, J, 3)
Q0h[, 1] <- 1L
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0h[a, 2:3] <- 0L
  Q0h[a, k + 1] <- 1L
}
rh <- pefa(Q0h, Yh, Kmin = 3, Kmax = 6, orthogonal = TRUE, v0 = .001,
           verbose = FALSE)
rh
```

It does not: the selection stalls at the lower boundary `K = 3`, with the
fit indices at their ceilings for every candidate, for exactly the
absorption reason -- SL-proportionality buys nothing, because the omitted
group factor's rank-one block is reproducible by the free entries either
way. Repeating this at several seeds and at `N = 500` gives the same lower
boundary selection every time; this is an identification property, not bad
luck. Worse, the truth-sized `K = 4` candidate *inside* the sweep is not
interpretable either. Its padded column is never anchored, so nothing pins
the general/group allocation: here the extra column simply stays empty and
the absorbed representation persists (cluster 3's group variance sits in
inflated general loadings plus off-cluster spill), while under other
settings and seeds the allocation comes out rotated across columns instead.
Either way, these loadings must not be read, checked for SL proportionality,
or transformed back:

```{r ho-sweep-k4}
f4 <- rh$fits[[which(rh$sweep$K == 4)]]
round(f4$Lam[cl == 3, ], 2)     # cluster-3 rows: compare sim_ho$MLA[cl == 3, ]
```

The working recipe is two steps, each doing only its own job. If the number
of group factors is genuinely unknown, run the sweep with every *known*
absence encoded as a fixed `0` (Section 2) -- that recovers the *count* with
an interior selection, though its fits stay uninterpretable. Then **refit at
the selected `K` with anchors on every group factor** -- the partially
confirmatory design of part (b) -- and do all interpretation there: on that
refit the fitted columns match the true SL matrix nearly perfectly
(congruence > .99), which is what parts (c)-(e) relied on. Anchors pin the
rotation; a sweep's padded columns, by construction, never have them.

### Caveats

This check is descriptive, not a hypothesis test: there is no p-value that
certifies "this bifactor is really an HO model," and the `mean CV < .1` rule
used above is a working convention, not a distributional result. Sampling
error alone makes the CV strictly positive even when the data are generated
exactly under an HO model, as they are here -- and the loading *ratio* it is
built on inherits the softness of the general-versus-group split, which is
why this section runs at a generous `N`; at a few hundred observations the
same population can land above the line on noise alone, and an occasional
item whose posterior settles on the wrong group factor will inflate the
pooled number further. So read the mean CV together with the per-cluster
(or per-item) ratios rather than as a verdict on its own. And the diagnostic
runs both ways -- a large mean CV, when it reflects genuinely non-constant
ratios rather than noise or an isolated item, is informative in its own
right: it means the bifactor solution is richer than any higher-order
representation of it, and collapsing it to a single second-order factor per
cluster would discard real structure.

## 4. References

* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5), 208-229.
  \doi{10.1177/01466216241261704}
* Chen, J. (2021). A generalized partially confirmatory factor analysis
  framework with mixed Bayesian Lasso methods. *Multivariate Behavioral
  Research*, 57(6), 879-894. \doi{10.1080/00273171.2021.1925520}
* Jin, Y., & Chen, J. (2025). Regularized variational approximation for
  partially confirmatory factor analysis. *Structural Equation Modeling*,
  32(3), 437-449. \doi{10.1080/10705511.2024.2432612}
