---
title: "Bifactor, higher-order, and testlet structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor, higher-order, and testlet structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette introduces the structures, fits an anchored bifactor model,
summarizes higher-order and testlet effects, and ends with a two-step
application. Factor-number selection and the bifactor comparison answer
different questions and are kept separate.

## 1. Three related structures

Suppose `J` items fall into `K` clusters.

**Correlated factors.** Each item loads on its cluster factor, and the cluster
factors correlate through $\Phi$.

**Orthogonal bifactor.** Every item loads on a general factor and on one group
factor; the general and group factors are mutually orthogonal.

**Higher-order (HO).** Items load on first-order factors, which in turn load on
a second-order factor. The HO model has an equivalent Schmid--Leiman (SL)
bifactor representation. For item $j$ in cluster $k$,

$$b^{gen}_j = \lambda_j\gamma_k, \qquad
  b^{grp}_j = \lambda_j\sqrt{1-\gamma_k^2}.$$

Thus $|b^{grp}_j|/|b^{gen}_j|$ is constant within each cluster when the SL
constraint holds. A bifactor model with visibly non-constant within-cluster
ratios is richer than that HO representation. Testlet or method factors use
the same general-plus-special-factor loading geometry; the group-factor effect
size below can therefore be read as a testlet effect when that interpretation
fits the design.

Cross-cluster covariance alone cannot distinguish a general factor from
correlated group factors: both produce rank-one cross-cluster blocks. The
distinguishing information is within clusters, so fitted comparisons remain
sensitive to anchors, residual structure, and how strongly the general and
group loading patterns differ.

## 2. Fit an anchored bifactor directly

`vbfa(Y, Q, bifactor = TRUE)` treats the columns of `Q` as group factors and
adds a fully specified general column internally. Consequently `K` denotes
the number of group factors, while the fitted loading matrix has `K + 1`
columns. The factor correlation matrix is fixed at the identity.

The following population is a genuine bifactor: general and group loadings
vary independently within each cluster rather than satisfying SL
proportionality.

```{r bifactor-simulate}
K <- 3
ipf <- 6
J <- K * ipf
cl <- rep(seq_len(K), each = ipf)

set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)
for (k in seq_len(K)) {
  mla[cl == k, k + 1] <- round(runif(ipf, .35, .70), 2)
}
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)

sim_bi <- sim_fa(N = 1000, mla = mla, phi = 0, rseed = 1)
Y <- sim_bi$dat
```

Anchor two items per group. For those rows, the own-group entry is one and
the other group entries are fixed at zero; the remaining entries stay
partially exploratory.

```{r bifactor-fit}
Q <- matrix(-1L, J, K)
for (k in seq_len(K)) {
  anchor <- which(cl == k)[1:2]
  Q[anchor, ] <- 0L
  Q[anchor, k] <- 1L
}

fit_bi <- vbfa(Y, Q, bifactor = TRUE)
fit_bi$Phi
round(fit_stats(fit_bi), 3)
```

`fit_bi$Q` contains the augmented design with the general column first. The
short form above is equivalent to constructing that all-one column manually
and fitting an orthogonal model, but it keeps the group-factor count explicit.
A quick recovery check can focus on the general/group split:

```{r bifactor-check}
c(
  congruence_general = abs(sum(fit_bi$Lam[, 1] * sim_bi$MLA[, 1])) /
    sqrt(sum(fit_bi$Lam[, 1]^2) * sum(sim_bi$MLA[, 1]^2)),
  MAE_general = mean(abs(abs(fit_bi$Lam[, 1]) - abs(sim_bi$MLA[, 1])))
)
```

## 3. Higher-order and testlet summaries

`sim_fa(..., gamma = )` generates the SL representation of an HO model. Fit
it as an anchored bifactor, then use `special_effects()` to summarize the
within-group ratios, the implied HO parameters, and group/testlet effect
sizes.

```{r higher-order}
gamma_true <- c(.80, .70, .60)
sim_ho <- sim_fa(
  N = 3000, K = K, ipf = ipf, lam = .8,
  gamma = gamma_true, rseed = 11
)
fit_ho <- vbfa(sim_ho$dat, Q, bifactor = TRUE)

effects <- special_effects(fit_ho)
effects$mean_cv
effects$groups[, c("group", "mean_ratio", "cv", "gamma", "D")]
```

For group $k$, `cv` describes how constant the fitted
$|b^{grp}_j|/|b^{gen}_j|$ ratios are, `gamma` is the corresponding approximate
second-order loading, and `D` is the average squared loading on the selected
group or testlet factor (Zhang & Chen, 2024, Eq. 16). `$items` contains the
item-level ratios and implied first-order loadings.

These summaries are descriptive. Small CVs support an approximate SL/HO
re-expression, but `special_effects()` does not perform a hypothesis test or
choose a structure. When ratios are not close to constant, `gamma` and the
implied first-order loadings are intentionally only summaries of the richer
bifactor fit; `D` remains a direct special-factor effect-size summary.

## 4. A two-step application

When the first-order count and the presence of a separable general factor are
both uncertain, use two steps. The first is the PEFA workflow of Chen and Jin
(2026a); the second is the research comparison developed in Chen and Jin
(2026b), not part of the PEFA gain rule itself.

### Step 1: select the first-order count with ordinary PEFA

Run an oblique, diagonal-residual sweep without a general factor. This selects
the number of cluster-level factors; it does not decide whether their
correlations should later be represented by a general factor.

```{r step1}
Q0 <- matrix(-1L, J, 2)
for (k in 1:2) {
  anchor <- which(cl == k)[1:2]
  Q0[anchor, ] <- 0L
  Q0[anchor, k] <- 1L
}

count_sweep <- pefa(Q0, Y, Kmin = 2, Kmax = 6, verbose = FALSE)
subset(
  count_sweep$selection,
  criterion == "elbo" & form == "gain" & cut_name == "primary"
)
```

Check that the selected count is not merely a requested-window boundary, and
read the fit trajectories descriptively. The PEFA vignette shows named-cut,
backbone, and adjacent-count sensitivity summaries. A boundary result requires
a complete wider sweep before proceeding.

### Step 2: compare matched anchored models by BIC

At the selected count, build one anchored group design and use it unchanged in
two direct fits: an oblique model and a bifactor model with an added general
factor. In this simulated example the selected count is three, so the `Q`
constructed in Section 2 is the matched design.

```{r step2}
fit_oblique <- vbfa(Y, Q)
fit_bifactor <- vbfa(Y, Q, bifactor = TRUE)

bic <- c(
  oblique = unname(fit_stats(fit_oblique)["BIC"]),
  bifactor = unname(fit_stats(fit_bifactor)["BIC"])
)
data.frame(model = names(bic), BIC = unname(bic),
           difference_from_oblique = unname(bic - bic["oblique"]))
```

Lower BIC is preferred. BIC is the operational comparison here because the
two competitors contain different loading designs and therefore different
prior bookkeeping; ELBO remains appropriate for comparing the same-shaped
candidates within the PEFA sweep.

This comparison is evidence, not an automatic structural verdict. Use the
same data, anchors, residual assumptions, and estimation controls in both
arms. Thin margins, sensitivity to defensible anchor choices, or conspicuous
residual dependence should be reported and investigated before interpreting a
bifactor preference. If the bifactor is retained, `special_effects()` then
describes whether it is close to an HO representation and quantifies its
group/testlet effects.

## References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical factor
  solutions. *Psychometrika*, 22(1), 53--61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113--128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4--5),
  208--229. \doi{10.1177/01466216241261704}
* Chen, J., & Jin, Y. (2026a). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
* Chen, J., & Jin, Y. (2026b). *When factor count is uncertain: Stable
  structure before the bifactor decision.* Manuscript in preparation.
