---
title: "Bifactor Structures and PEFA Sweeps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor Structures and PEFA Sweeps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette covers the bifactor side of `vbpm`: what a bifactor model is,
how a `pefa()` sweep is read when the factor count is uncertain, and how a
later oblique-versus-bifactor comparison is built out of the sweep that
preceded it. The sweep object itself is walked in `vignette("pefa")`; nothing
here re-explains its tables.

## Three related structures

Suppose `J` items fall into `K` clusters.

**Correlated factors.** Each item loads on its cluster factor, and the cluster
factors correlate through $\Phi$.

**Orthogonal bifactor.** Every item loads on a general factor and on one group
factor; general and group factors are mutually orthogonal.

**Higher-order.** Items load on first-order factors, which load in turn on a
second-order factor. Its Schmid-Leiman representation is the orthogonal
bifactor with

$$
b^{gen}_j=\lambda_j\gamma_k,
\qquad
b^{grp}_j=\lambda_j\sqrt{1-\gamma_k^2}
$$

for item $j$ in cluster $k$, so $|b^{grp}_j|/|b^{gen}_j|$ is constant within a
cluster under exact higher-order proportionality. Testlet and method factors
use the same general-plus-special geometry under a different substantive
reading.

In `vbfa()` and `pefa()`, `bifactor = TRUE` makes `K` the number of **group**
factors: the design matrix supplies only the group block and the general column
is added internally, so a fitted matrix has `K + 1` columns. Every
between-candidate comparison a sweep performs removes that general column
first and works on the `K` group columns, which is what makes an ordinary and a
bifactor sweep of the same data structurally comparable.

## A bifactor population and two backbones

Three clusters of six items, each item loading on a general factor and on its
own group factor.

```{r population}
K <- 3
ipf <- 6
J <- K * ipf
cluster <- rep(seq_len(K), each = ipf)

set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)
for (k in seq_len(K)) {
  mla[cluster == k, k + 1] <- round(runif(ipf, .35, .70), 2)
}
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)

sim <- sim_fa(N = 800, mla = mla, phi = 0, rseed = 1)
Y <- sim$dat
```

Two anchors per backbone factor, expressed as the two designs `vbpm`
documentation uses throughout.

* **AO (anchor-only)** specifies each anchor's intended cell as `1` and leaves
  that anchor's cells on the *other* backbone columns unspecified (`-1`).
* **AZ (anchor-zero)** additionally fixes those cross-backbone anchor cells to
  `0`.

```{r backbones}
Q0_AO <- matrix(-1L, J, 2)
for (k in 1:2) Q0_AO[which(cluster == k)[1:2], k] <- 1L

Q0_AZ <- Q0_AO
for (k in 1:2) {
  anchors <- which(cluster == k)[1:2]
  Q0_AZ[anchors, ] <- 0L
  Q0_AZ[anchors, k] <- 1L
}

Q0_show <- cbind(Q0_AO[1:8, ], Q0_AZ[1:8, ])
colnames(Q0_show) <- c("AO_1", "AO_2", "AZ_1", "AZ_2")
Q0_show
```

AZ is the stronger claim: it asserts that an anchor of cluster 1 has no loading
on backbone factor 2. AO only asserts where the anchor *does* load. Both
backbones cover two of the three clusters, so the third has to be found by the
sweep.

## Step 1: ordinary oblique sweeps

Every `pefa()` and `vbfa()` call below omits `v0` and `max_it` and therefore
runs at its own function's documented defaults. Those are not one shared cap:
`pefa()` allows `max_it = 10000` per candidate and `vbfa()` allows 5000.

```{r step1}
AO <- pefa(Q0_AO, Y, Kmin = 2, Kmax = 5, verbose = FALSE)
AZ <- pefa(Q0_AZ, Y, Kmin = 2, Kmax = 5, verbose = FALSE)

AO$sweep[, c("K", "ELBO", "AIC", "BIC", "SRMR", "CFI")]
AZ$sweep[, c("K", "ELBO", "AIC", "BIC", "SRMR", "CFI")]
```

```{r step1-transitions}
AO$transitions[, c("K_from", "K_to", "ELBO_gain_pct", "BIC_gain_pct",
                   "phi_min", "rmsd_max", "unmatched_ssl", "collision")]
AZ$transitions[, c("K_from", "K_to", "ELBO_gain_pct", "BIC_gain_pct",
                   "phi_min", "rmsd_max", "unmatched_ssl", "collision")]
```

Both backbones put their only positive gain on the `2 -> 3` step and both put
`unmatched_ssl` near 2.6 there, which is a full column's worth against the
`ssl()` values below. Both minimize BIC at `K = 3` --- 35212.89 for AO and
35209.52 for AZ --- but only just: the `3 -> 4` BIC step gives back 0.5% (AO)
and 1.0% (AZ) of that first gain, so the criterion is nearly flat between three
and four factors. AIC dissents in both backbones, taking its minimum at `K = 4`
(35015.32 and 35015.94, against 35025.55 and 35022.18 at three).

```{r step1-persistence}
round(AO$persistence$phi, 3)
round(AZ$persistence$phi, 3)
ssl(AO)[c("3", "4")]
ssl(AZ)[c("3", "4")]
```

Persistence separates the two backbones a little more than the criteria do.
AO's three-factor columns survive into `K = 4` and `K = 5` at 0.989; AZ's
survive at 0.985, and AZ's `2 -> 4` and `2 -> 5` cells read 0.905 against AO's
0.925. In both backbones the fourth column is small --- `ssl()` puts it at 0.37
(AO) and 0.45 (AZ) against three columns above 2.4 --- so the four-factor
candidate is the three-factor one plus a thin extra column, not a different
structure.

```{r step1-reading}
## Analysis-side reading, not a package rule. Variant: full look-ahead with no
## fallback -- the smallest K_from whose own and every later adjacent ELBO gain
## is under 20% of the largest positive gain, NA_integer_ when none qualifies.
reading <- vapply(list(AO = AO, AZ = AZ), function(x) {
  g <- x$transitions$ELBO_gain_pct
  below <- vapply(seq_along(g), function(i) all(g[i:length(g)] < 20),
                  logical(1))
  if (any(below)) x$transitions$K_from[which(below)[1]] else NA_integer_
}, integer(1))
reading
```

Both deliver three. That agreement is a fact about this sample and window; the
AIC minimum at four is in the same evidence and disagrees, and no part of
`pefa()` arbitrates between them.

## Step 2: a comparison derived from Step 1

Step 1 delivers a count and a loading matrix. Step 2 is built out of *that*
result, not out of a design written down independently, so `Kp` is the count
the reading above returned and the group Q is derived from the AO candidate at
that count. All of this is ordinary analysis-side code: `pefa()` neither
performs it nor stores it.

Let `K0 = ncol(Q0)` and let the **specification depth** `d` be the uniform
number of backbone markers per factor, `colSums(Q0 == 1L)`. The automatic route
needs one common depth; had the backbone columns been specified to different
depths, the analyst would write down a depth rule of their own or supply a
complete substantive `J x Kp` group-loading Q instead.

At `Kp == K0` the Step-2 design is `Q0` itself. Otherwise `Q0` is carried over
cell for cell --- every anchor, every AO or AZ zero, every unspecified cell ---
and `Kp - K0` all-`-1` columns are appended. Each added column is then anchored
independently: rank all `J` items by decreasing absolute loading in that column
of the delivered candidate, break exact ties by item index, and turn its top `d`
cells from `-1` into `1`. Nothing else is fixed --- no PIP cutoff, no magnitude
cutoff, no dominant-factor assignment, no uniqueness restriction --- so the same
item may be a top marker for more than one added factor.

```{r step2-design}
Kp    <- reading[["AO"]]
K0    <- ncol(AO$Q0)
Lam   <- AO$loadings[[as.character(Kp)]]
d     <- unique(colSums(AO$Q0 == 1L))
added <- if (Kp > K0) seq(K0 + 1L, Kp) else integer(0)
stopifnot(length(d) == 1L, d > 0L)        # one inherited specification depth

Q2 <- AO$Q0
if (length(added)) {
  Q2 <- cbind(Q2, matrix(-1L, J, Kp - K0,
                         dimnames = list(rownames(AO$Q0),
                                         paste0("F", added))))
  for (k in added) {
    top <- order(abs(Lam[, k]), seq_len(J),
                 decreasing = c(TRUE, FALSE), method = "radix")[seq_len(d)]
    Q2[top, k] <- 1L
  }
}

c(Kp = Kp, K0 = K0, depth = d)
Q2[rowSums(Q2 == 1L) > 0L, , drop = FALSE]
```

The added column inherits the backbone's depth of two and anchors itself on
items 13 and 18, both from the cluster the two-column backbone did not cover.
The checks below run at render time, so the vignette fails rather than
misreports if the construction drifts from the recipe.

```{r step2-check}
stopifnot(
  ## group columns only, no general column
  nrow(Q2) == J, ncol(Q2) == Kp,
  ## the first K0 columns are Q0 itself, cell for cell
  identical(Q2[, seq_len(K0), drop = FALSE], AO$Q0),
  ## no new fixed zero anywhere, in an added column or outside one
  all(Q2[, added, drop = FALSE] %in% c(-1L, 1L)),
  sum(Q2 == 0L) == sum(AO$Q0 == 0L),
  ## each added column inherits the backbone depth exactly
  all(colSums(Q2[, added, drop = FALSE] == 1L) == d),
  ## and its markers are the top d absolute loadings, exact ties by item
  ## index, with no PIP or magnitude cutoff standing between them
  all(vapply(added, function(k) {
    a   <- abs(Lam[, k])
    sel <- which(Q2[, k] == 1L)
    oth <- setdiff(seq_len(J), sel)
    all(vapply(sel, function(i)
      all(a[i] > a[oth] | (a[i] == a[oth] & i < oth)), logical(1)))
  }, logical(1)))
)
```

The same group Q now goes to both models. The bifactor general column is added
inside `vbfa()` and never participated in marker selection, which the second
check below confirms.

```{r step2-fit}
f_obl <- vbfa(Y, Q2)
f_bif <- vbfa(Y, Q2, bifactor = TRUE)
stopifnot(identical(f_obl$Q, Q2),
          identical(f_bif$Q[, -1L, drop = FALSE], Q2))

idx <- c("RMSEA", "SRMR", "CFI", "TLI", "AIC", "BIC")
round(rbind(
  oblique  = fit_stats(f_obl, Y = Y, Q = f_obl$Q, tau = .50)[idx],
  bifactor = fit_stats(f_bif, Y = Y, Q = f_bif$Q, tau = .50)[idx]
), 3)
```

The bifactor arm is ahead on every index, which is the right answer for this
population, but by very unequal margins: 71.2 AIC points and only 5.6 BIC
points. It is an operational comparison of two fitted models on one design, not
a test against an unrestricted alternative, and it is conditional on the count
and the markers that Step 1 delivered. A thin margin, or a margin that flips
when the Step-1 count changes, should be reported rather than resolved.

An analyst may instead supply a substantive `J x Kp` group-loading Q of their
own, never one containing the bifactor general column. That is a different
analysis design: it should be labelled as such, recorded, validated for codes
and dimnames, and used identically in both arms. It is not a silent default and
it is not a `pefa()` decision.

`special_effects()` then describes the retained bifactor fit: within-cluster
Schmid-Leiman ratios and their CVs, the implied second-order loadings, and
group effect sizes.

```{r step2-effects}
effects <- special_effects(f_bif)
effects$mean_cv
effects$groups[, c("group", "mean_ratio", "cv", "gamma", "D")]
```

## The same window as a bifactor sweep

`pefa(..., bifactor = TRUE)` sweeps the number of group factors. The backbone
supplies group columns only, the general column is added internally, and the
comparator strips it before matching, so the correspondence and the seven pair
facts have exactly the definitions used above.

```{r bifactor-sweep}
BI <- pefa(Q0_AO, Y, Kmin = 2, Kmax = 5, bifactor = TRUE, verbose = FALSE)
BI$sweep[, c("K", "ELBO", "AIC", "BIC", "SRMR", "CFI")]
BI$transitions[, c("K_from", "K_to", "ELBO_gain_pct", "BIC_gain_pct",
                   "phi_min", "rmsd_max", "unmatched_ssl", "collision")]
round(BI$persistence$phi, 3)
round(BI$persistence$rmsd, 3)
ssl(BI)
```

In this example the bifactor sweep is much less sharp than the ordinary one.

* `ELBO_gain_pct` is `NA` for every row. ELBO is highest at the window's lower
  edge and falls at every step, so that path contains no positive gain and its
  whole percentage column is `NA` by construction. The Step-1 reading has no
  defined path to consume; a rule that reads that column must decide for itself
  what an all-`NA` path means, which is exactly the kind of decision the
  package leaves outside. `BIC_gain_pct` does have a positive step, at
  `3 -> 4`, and normalizes to it.
* `phi_min` is 0.998 or better in every cell that involves `K = 2` and 0.990 at
  `4 -> 5`, then drops to 0.444 and 0.519 in the two cells that start at
  `K = 3`. No `rmsd_max` cell exceeds 0.122, so no column moved far in absolute
  terms; those two low congruences are the weakest-link statistic scoring a
  group column that has been emptied.
* `ssl()` says why. The general column holds 4.29-4.37 and the two backbone
  group columns hold about 2.09-2.25 at every candidate, while the third group
  column reaches only 0.27 at `K = 3` and collapses to the order of 1e-9 to
  1e-7 at `K = 4` and `K = 5`. With the general factor already carrying the
  shared variance, the third cluster's specific variance never accumulates into
  a group column here. The four- and five-factor candidates are the two-group
  solution again --- `$persistence$phi` reads 1.000 and 0.999 against `K = 2`
  --- which is also why the four-factor candidate's two empty columns pick the
  same target and raise `collision` at `4 -> 5`.

Read that as a property of this example --- this population, this two-column
backbone, this window, this sample size --- and not as a general ranking. A
bifactor sweep measures the same quantities as an ordinary one, and whether it
separates candidates as well depends on how much group-specific variance is
left once a general factor is in the model. A study that needs both readings
should run both on the same data, backbone, and window rather than assume they
must agree.

## References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical factor
  solutions. *Psychometrika*, 22(1), 53-61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113-128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5), 208-229.
  \doi{10.1177/01466216241261704}
* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
