---
title: "Structural Persistence Before the Bifactor Decision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural Persistence Before the Bifactor Decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette introduces bifactor, higher-order, and testlet structures, then
shows how structural evidence from an ordinary PEFA sweep can be converted into
different, explicit decision profiles before a bifactor comparison. The
package returns the evidence object; it does not select a profile, threshold,
horizon, collision policy, factor count, or layer. Fit assessment and count-only
readings are treated in `vignette("pefa")`.

## 1. Three related structures

Suppose `J` items fall into `K` clusters.

**Correlated factors.** Each item loads on its cluster factor, and the cluster
factors correlate through $\Phi$.

**Orthogonal bifactor.** Every item loads on a general factor and on one group
factor; the general and group factors are mutually orthogonal.

**Higher-order.** Items load on first-order factors, which in turn load on a
second-order factor. The Schmid-Leiman representation is the orthogonal
bifactor with

$$
b^{gen}_j=\lambda_j\gamma_k,
\qquad
b^{grp}_j=\lambda_j\sqrt{1-\gamma_k^2}
$$

for item $j$ in cluster $k$. Consequently
$|b^{grp}_j|/|b^{gen}_j|$ is constant within a cluster under exact
higher-order proportionality. Testlet or method factors use the same
general-plus-special-factor geometry under a different substantive reading.

Each individual cross-cluster covariance block is rank one under both a general
factor and correlated group factors, so one block alone cannot distinguish
them. Across four or more clusters, however, the joint block scales obey
multiplicative (tetrad-type) restrictions under a single general or higher-order
dimension and can differ from unrestricted factor correlations. Within-cluster
evidence, anchors, residual assumptions, and persistence of the group-loading
structure therefore remain consequential rather than uniquely decisive.

## 2. Fit and summarize a bifactor directly

`vbfa(Y, Q, bifactor = TRUE)` treats the columns of `Q` as group factors and
adds a fully specified general column internally. Thus `K` counts group
factors, while the fitted loading matrix has `K + 1` columns.

```{r bifactor-simulate}
K <- 3
ipf <- 6
J <- K * ipf
cluster <- rep(seq_len(K), each = ipf)

set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)
for (k in seq_len(K)) {
  mla[cluster == k, k + 1] <- round(runif(ipf, .35, .70), 2)
}
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)

sim_bi <- sim_fa(N = 1000, mla = mla, phi = 0, rseed = 1)
Y <- scale(sim_bi$dat, center = TRUE, scale = TRUE)
```

```{r bifactor-fit}
Q_direct <- matrix(-1L, J, K)
for (k in seq_len(K)) {
  anchors <- which(cluster == k)[1:2]
  Q_direct[anchors, ] <- 0L
  Q_direct[anchors, k] <- 1L
}

fit_bi <- vbfa(
  Y, Q_direct, bifactor = TRUE,
  v0 = c(.01, .005, .002, .001), max_it = 5000,
  convChk = FALSE, tolVal = 1e-4
)
fit_bi$Phi
round(fit_stats(
  fit_bi, Y = Y, Q = fit_bi$Q, tau = .50,
  rank_adjust = TRUE, rank_max_J = 100
), 3)
```

`special_effects()` summarizes fitted within-cluster ratios, approximate
higher-order parameters, and group/testlet effect sizes. These are descriptive
summaries rather than a structure-selection test.

```{r higher-order}
gamma_true <- c(.80, .70, .60)
sim_ho <- sim_fa(
  N = 3000, K = K, ipf = ipf, lam = .8,
  gamma = gamma_true, rseed = 11
)
fit_ho <- vbfa(sim_ho$dat, Q_direct, bifactor = TRUE)
effects <- special_effects(fit_ho)
effects$mean_cv
effects$groups[, c("group", "mean_ratio", "cv", "gamma", "D")]
```

## 3. Construct one Step-1 evidence object

When both the first-order count and an added general dimension are uncertain,
Step 1 uses ordinary oblique PEFA. It asks whether cluster-level loading
patterns persist across candidate counts; it does not fit a general factor.

Use the same anchors in two backbones. Anchor-only (AO) leaves each anchor's
cross-backbone cells unspecified. Anchor-zero (AZ) fixes those cells to zero.
Both branches use the same sample and the same fixed window. With source counts
`2:3` and maximum illustrative horizon three, candidates through six provide all
required direct targets.

```{r step1-evidence}
Q0_AO <- matrix(-1L, J, 2)
for (k in 1:2) {
  anchors <- which(cluster == k)[1:2]
  Q0_AO[anchors, k] <- 1L
}

Q0_AZ <- Q0_AO
for (k in 1:2) {
  anchors <- which(cluster == k)[1:2]
  Q0_AZ[anchors, ] <- 0L
  Q0_AZ[anchors, k] <- 1L
}

sweep_AO <- pefa(
  Q0_AO, Y, Kmin = 2, Kmax = 6,
  v0 = .001, max_it = 10000, convChk = FALSE, tolVal = 1e-4,
  tau = .50, stability_eps = .10,
  rank_adjust = TRUE, rank_max_J = 100, verbose = FALSE
)
sweep_AZ <- pefa(
  Q0_AZ, Y, Kmin = 2, Kmax = 6,
  v0 = .001, max_it = 10000, convChk = FALSE, tolVal = 1e-4,
  tau = .50, stability_eps = .10,
  rank_adjust = TRUE, rank_max_J = 100, verbose = FALSE
)

evidence_object <- list(AO = sweep_AO, AZ = sweep_AZ)
```

Each branch also carries immutable `$provenance` identifying the evidence
schema/engine and hashing its standardized data, Q0, settings, scientific
payload, and lineage declaration. This engine-data fingerprint is distinct from any external hash of the
matrix passed to `pefa()`; retain both when the outer analysis defines one.
Decision profiles are deliberately absent from that package record.

The all-pairs table contains one direct comparison for every `K_from < K_to`.
Changing horizon or threshold therefore requires no refit and no chained match.

```{r persistence-evidence}
sweep_AO$persistence
persistence(sweep_AO, "phi_min")
persistence(sweep_AO, "rmsd_max")
persistence(sweep_AO, "unmatched_ssl")
persistence(sweep_AO, "n_collisions")
ssl(sweep_AO)
```

A collision marker `*` says that two source columns selected the same target;
`~` marks a reduced scored-pair set, `!` a nonconverged endpoint, and `?` an
unavailable selected metric. `n_pairs` gives the exact scored-pair count behind
the matched-pair summaries. `unmatched_ssl` is on the same column-SSL scale as
`ssl()`, so it can be read against the retained candidate columns. None of these
facts is a package verdict. The decision profile below states how they enter an
edge.

## 4. Apply several decision profiles

A decision profile records every judgement needed to turn the same evidence
object into a count reading and a profile-indexed L1/L2/L3 result. The following
values are pedagogical. They are not `vbpm` defaults or universal R8 cutoffs.
The source hash is left `NA` only because this installed-vignette illustration
is not a sealed research registry; a production registry must populate it.

```{r profile-table}
decision_profiles <- data.frame(
  profile_id = c(
    "illustrative_phi", "allow_completed_endpoints", "shorter_horizon",
    "longer_horizon", "pefa_fit_v1_style", "bic_gain_phi", "ao_rmsd",
    "az_phi", "flag_collision"
  ),
  profile_role = "illustration",
  profile_source_id = "vbpm_bifactor_vignette",
  profile_source_version = "0.9.0",
  profile_source_sha256 = NA_character_,
  specification = c("AO", "AO", "AO", "AO", "AO", "AO", "AO", "AZ", "AO"),
  criterion = c("elbo", "elbo", "elbo", "elbo", "elbo", "bic", "elbo", "elbo", "elbo"),
  count_rule = "gain",
  count_rule_semantics_id = "vbpm_0.9_gain_full_lookahead_na",
  candidate_path_policy_id = "complete_consecutive_all_ok",
  endpoint_fit_policy = c(
    "require_converged", "allow_completed", "require_converged",
    "require_converged", "require_converged", "require_converged",
    "require_converged", "require_converged", "require_converged"
  ),
  delta = c(20, 20, 20, 20, 10, 20, 20, 20, 20),
  sustain = c(1, 1, 1, 1, 2, 1, 1, 1, 1),
  metric = c(
    "phi_min", "phi_min", "phi_min", "phi_min", "phi_min", "phi_min",
    "rmsd_max", "phi_min", "phi_min"
  ),
  metric_direction = c(
    "at_least", "at_least", "at_least", "at_least", "at_least",
    "at_least", "at_most", "at_least", "at_least"
  ),
  threshold = c(.85, .85, .85, .85, .85, .85, .20, .85, .85),
  horizon_r = c(2L, 2L, 1L, 3L, 2L, 2L, 2L, 2L, 2L),
  source_set_id = "demo_2_3",
  collision_policy = c("veto", "veto", "veto", "veto", "veto", "veto", "veto", "veto", "flag"),
  source_eligibility_policy = "require_all",
  target_capacity_policy = "require_distinct_capacity",
  layer_rule_id = "richest_persistent",
  assessment_step2 = c(TRUE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE),
  stringsAsFactors = FALSE
)
decision_profiles
```

The helper below is vignette code, not a package classifier. It makes the
direction of every decision visible: congruence must be at least its threshold,
RMSD must be no greater than its threshold, and a collision is rejected only
under a `"veto"` profile.

```{r profile-helper}
assess_profile <- function(evidence_object, profile,
                           source_sets = list(demo_2_3 = 2:3)) {
  stopifnot(
    profile$count_rule == "gain",
    profile$count_rule_semantics_id == "vbpm_0.9_gain_full_lookahead_na",
    profile$candidate_path_policy_id == "complete_consecutive_all_ok",
    profile$endpoint_fit_policy %in% c("require_converged", "allow_completed"),
    profile$source_eligibility_policy == "require_all",
    profile$target_capacity_policy == "require_distinct_capacity",
    profile$layer_rule_id == "richest_persistent"
  )
  source_set <- source_sets[[profile$source_set_id]]
  if (is.null(source_set)) stop("Unknown source_set_id.")
  x <- evidence_object[[profile$specification]]
  K_path <- as.integer(x$sweep$K)
  score <- switch(
    profile$criterion,
    elbo = x$sweep$ELBO,
    aic = -x$sweep$AIC,
    bic = -x$sweep$BIC
  )
  component_ok <- switch(
    profile$criterion,
    elbo = x$sweep$elbo_status == "ok",
    aic = x$sweep$stats_status == "ok",
    bic = x$sweep$stats_status == "ok"
  )
  path_ok <- identical(K_path, seq.int(min(K_path), max(K_path))) &&
    all(x$sweep$fit_status == "ok" & component_ok & is.finite(score))
  Khat <- if (path_ok) {
    select_K_elbow(
      K_path, score,
      delta = profile$delta,
      sustain = profile$sustain
    )
  } else {
    NA_integer_
  }
  count_status <- if (!path_ok) "unusable" else if (is.na(Khat)) "absent" else "usable"

  z <- x$persistence
  metric_value <- z[[profile$metric]]
  z$metric_pass <- switch(
    profile$metric_direction,
    at_least = metric_value >= profile$threshold,
    at_most = metric_value <= profile$threshold,
    stop("Unsupported metric_direction.")
  )
  endpoint_admissible <- switch(
    profile$endpoint_fit_policy,
    require_converged =
      (z$from_fit_ok %in% TRUE) & (z$to_fit_ok %in% TRUE),
    allow_completed = !is.na(z$from_fit_ok) & !is.na(z$to_fit_ok),
    stop("Unsupported endpoint_fit_policy.")
  )
  edge_known <-
    z$pair_status %in% "available" &
    is.finite(metric_value) &
    endpoint_admissible &
    !is.na(z$source_ineligible_n) &
    !is.na(z$target_eligible_n) &
    !is.na(z$collision)
  z$edge_pass <- rep(NA, nrow(z))
  z$edge_pass[edge_known] <-
    z$source_ineligible_n[edge_known] == 0L &
    z$target_eligible_n[edge_known] >=
      (z$K_from[edge_known] - ncol(x$Q0)) &
    z$metric_pass[edge_known] &
    (profile$collision_policy == "flag" | !z$collision[edge_known])

  P <- setNames(rep(NA, length(source_set)), as.character(source_set))
  required_rows <- integer()
  for (k in source_set) {
    needed_idx <- which(z$K_from == k &
                          z$K_to %in% (k + seq_len(profile$horizon_r)))
    needed <- z[needed_idx, , drop = FALSE]
    if (nrow(needed) != profile$horizon_r) next
    required_rows <- c(required_rows, needed_idx)
    if (any(needed$edge_pass %in% FALSE)) P[as.character(k)] <- FALSE
    else if (all(needed$edge_pass %in% TRUE)) P[as.character(k)] <- TRUE
  }

  stable <- as.integer(names(P)[P %in% TRUE])
  Kp <- if (length(stable)) max(stable) else NA_integer_
  higher_unknown <- if (is.na(Kp)) any(is.na(P)) else
    any(is.na(P[names(P) %in% as.character(source_set[source_set > Kp])]))

  layer <- if (all(P %in% FALSE)) {
    "L3"
  } else if (length(stable) && !higher_unknown) {
    if (!path_ok) NA_character_ else if (!is.na(Khat) && Khat == Kp) "L1" else "L2"
  } else {
    NA_character_
  }

  req <- sort(unique(required_rows))
  required_edge_state <- if (length(req)) paste0(
    z$K_from[req], "->", z$K_to[req], ":",
    ifelse(is.na(z$edge_pass[req]), "?", as.integer(z$edge_pass[req])),
    collapse = ","
  ) else ""

  data.frame(
    profile_id = profile$profile_id,
    specification = profile$specification,
    Khat = Khat,
    count_status = count_status,
    Kp = Kp,
    layer = layer,
    required_edge_state = required_edge_state,
    persistence_state = paste0(
      names(P), ":", ifelse(is.na(P), "?", as.integer(P)), collapse = ","
    ),
    stable_set = if (length(stable)) paste(stable, collapse = ",") else "",
    required_collided_edge_n = if (!length(req) || anyNA(z$collision[req]))
      NA_integer_ else as.integer(sum(z$collision[req])),
    required_collision_target_n = if (!length(req) || anyNA(z$n_collisions[req]))
      NA_integer_ else as.integer(sum(z$n_collisions[req])),
    analysis_status = if (is.na(layer)) "technically_unresolved" else "complete"
  )
}
```

```{r profile-results}
profile_results <- do.call(
  rbind,
  lapply(seq_len(nrow(decision_profiles)), function(i) {
    assess_profile(evidence_object, decision_profiles[i, , drop = FALSE])
  })
)
cbind(decision_profiles, profile_results[, c(
  "Khat", "count_status", "Kp", "persistence_state", "stable_set", "layer",
  "required_edge_state", "required_collided_edge_n",
  "required_collision_target_n", "analysis_status"
)])
```

Differences in this table have interpretable sources. Each row changes one
decision family relative to `illustrative_phi`: endpoint-fit policy, horizon,
PEFA-fit-style count settings, count criterion, structural metric/gate, AO
versus AZ, or collision policy. Under `require_converged`, a nonconverged
endpoint yields `?`; `allow_completed` may evaluate the retained finite metric
only when both endpoints have known completion facts. A fit error or malformed
convergence flag remains `?` under either policy.
Changing `delta` or `sustain` changes only the count reading. Changing the metric, threshold, horizon,
or collision policy can change the persistent count. L1 means the two readings
agree under that profile; L2 means a persistent structure exists but count
evidence disagrees or is absent; L3 means no source count persists under that
profile. The layer is therefore `profile_id`-indexed, not an intrinsic label of
the dataset.

The horizons are deliberately local. At a five-column major target, horizons
`1:3`
asks whether the same core survives after one, two, or three added columns. This
is a bounded local-to-moderate overextraction band for that design, not a
universal complexity scale: mild overextraction may preserve the major core
beside a surplus column, but added columns can also split or destabilize it.
Underextraction cannot contain every major direction. The profile table shows
all three horizons rather than letting the package choose among them.

For confirmatory work, declare the profile before interpreting its layer. For
exploratory work, show the relevant profile comparison rather than selecting
the row that produces the preferred conclusion. A profile outside the operating
conditions studied in a supporting simulation is possible, but its error rate
is not thereby validated.

## 5. Step 2 is conditional on a declared delivery

Only an L1/L2 profile supplies a count for the later oblique-versus-bifactor
comparison. L3 and technically unresolved profiles deliver no structure. The
comparison must use the selected branch's candidate at its actual delivered
count; it must not substitute another branch, count, or known population design.

The R8 workflow uses two paired design families at a declared delivered count:

1. a loading-derived disjoint-support `Q_hard`, fit as strict oblique and strict
   bifactor models; and
2. the original branch-preserving `Q_flex`, fit as a flexible oblique model and
   the **general-plus-sparse-groups sensitivity**.

The four-arm code is shown conditionally because this illustrative dataset may
legitimately receive L3 under the prospectively marked assessment profile.

```{r step2, eval=FALSE}
assessment_id <- decision_profiles$profile_id[decision_profiles$assessment_step2]
stopifnot(length(assessment_id) == 1L)
chosen <- profile_results[
  profile_results$profile_id == assessment_id &
    profile_results$layer %in% c("L1", "L2"),
]
stopifnot(nrow(chosen) == 1L)

x <- evidence_object[[chosen$specification]]
Kp <- chosen$Kp
Lam <- x$loadings[[as.character(Kp)]]
cand <- x$sweep[x$sweep$K == Kp, , drop = FALSE]
stopifnot(
  nrow(cand) == 1L,
  cand$fit_status == "ok",
  cand$loading_status == "ok",
  is.matrix(Lam), nrow(Lam) == nrow(x$Q0), ncol(Lam) == Kp,
  all(is.finite(Lam))
)
K0 <- ncol(x$Q0)
Q_flex <- if (Kp == K0) x$Q0 else
  cbind(x$Q0, matrix(-1L, nrow(x$Q0), Kp - K0))

# Deterministic disjoint support: each item is assigned to its largest absolute
# loading, with the smaller factor index resolving exact ties. Preserve the Q0
# anchors; choose two distinct markers for every exploratory factor from its
# assigned items. A profile is Step-2 unavailable if this is impossible.
anchor_index <- which(x$Q0 == 1L, arr.ind = TRUE)
stopifnot(all(rowSums(x$Q0 == 1L) <= 1L))
anchor_rows <- unique(anchor_index[, "row"])
nonanchor_rows <- setdiff(seq_len(nrow(Lam)), anchor_rows)

Q_hard <- matrix(0L, nrow(Lam), Kp)
Q_hard[anchor_index] <- 1L
support <- max.col(abs(Lam[nonanchor_rows, , drop = FALSE]), ties.method = "first")
Q_hard[cbind(nonanchor_rows, support)] <- -1L

marker_rows <- vector("list", Kp)
for (h in seq_len(Kp)) {
  if (h <= K0) {
    marker_rows[[h]] <- anchor_index[anchor_index[, "col"] == h, "row"]
  } else {
    assigned <- which(Q_hard[, h] == -1L)
    if (length(assigned) < 2L)
      stop("Step 2 unavailable: an exploratory factor lacks two markers.")
    ord <- order(-abs(Lam[assigned, h]), assigned)
    marker_rows[[h]] <- assigned[ord[1:2]]
    Q_hard[marker_rows[[h]], h] <- 1L
  }
}
stopifnot(all(lengths(marker_rows) == 2L))
stopifnot(all(rowSums(Q_hard != 0L) == 1L))

arm_controls <- list(
  v0 = c(.01, .005, .002, .001), max_it = 5000,
  ld = FALSE, Qe = NULL, ld_control = list(),
  convChk = FALSE, tolVal = 1e-4
)
fits <- list(
  strict_oblique = do.call(
    vbfa, c(list(Y = Y, Q = Q_hard, bifactor = FALSE, orthogonal = FALSE),
            arm_controls)
  ),
  strict_bifactor = do.call(
    vbfa, c(list(Y = Y, Q = Q_hard, bifactor = TRUE, orthogonal = TRUE,
                 general = 1), arm_controls)
  ),
  flexible_oblique = do.call(
    vbfa, c(list(Y = Y, Q = Q_flex, bifactor = FALSE, orthogonal = FALSE),
            arm_controls)
  ),
  general_plus_sparse_groups = do.call(
    vbfa, c(list(Y = Y, Q = Q_flex, bifactor = TRUE, orthogonal = TRUE,
                 general = 1), arm_controls)
  )
)

arm_stats <- lapply(fits, function(f) {
  if (!isTRUE(f$flag == 1L))
    return(list(status = "unavailable", reason = "nonconverged", BIC = NA_real_))
  fs <- tryCatch(
    fit_stats(
      f, Y = Y, Q = f$Q, tau = .50,
      rank_adjust = TRUE, rank_max_J = 100
    ),
    error = function(e) e
  )
  if (inherits(fs, "error"))
    return(list(status = "unavailable", reason = "fit_stats_error", BIC = NA_real_))
  if (!is.finite(fs["BIC"]))
    return(list(status = "unavailable", reason = "nonfinite_hard_bic", BIC = NA_real_))
  list(status = "ok", reason = NA_character_, BIC = unname(fs["BIC"]))
})
stopifnot(all(vapply(arm_stats, `[[`, character(1), "status") == "ok"))
bic <- vapply(arm_stats, `[[`, numeric(1), "BIC")
c(
  delta_BIC_strict = bic["strict_oblique"] - bic["strict_bifactor"],
  delta_BIC_flexible =
    bic["flexible_oblique"] - bic["general_plus_sparse_groups"]
)
```

Positive `delta_BIC` favors the corresponding bifactor arm. The contrast is an
operational, profile-conditional comparison, not a formal test of an unrestricted
population distance. Thin margins, disagreement between design families, or
conspicuous residual dependence should be reported before a substantive general
factor is interpreted. If a bifactor is retained, `special_effects()` can then
describe its higher-order proportionality and group/testlet effects.

## References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical factor
  solutions. *Psychometrika*, 22(1), 53-61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113-128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5), 208-229.
  \doi{10.1177/01466216241261704}
* Chen, J., & Jin, Y. (2026a). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
* Chen, J., & Jin, Y. (2026b). *When factor count is uncertain: Stable
  structure before the bifactor decision.* Manuscript in preparation.
