---
title: "Bifactor, higher-order, and testlet structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor, higher-order, and testlet structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette is a small tutorial paper on estimating structures that attach
an "extra" factor to a set of group factors: the orthogonal **bifactor**
model, the **higher-order** (HO) model, and the **testlet** model. It answers
four questions in order, with every claim demonstrated on simulated data:

1. How is an orthogonal bifactor fit with `vbfa()`, and how well does it
   recover a true bifactor structure? (Section 2)
2. Can a `pefa()` sweep determine the number of *group* factors in a
   bifactor design? (Sections 3 and 5: **no** -- and the reason is an
   identification property worth understanding, not a tuning problem.)
3. How are higher-order and testlet structures simulated, fit, checked, and
   effect-sized? (Section 4)
4. Given data of unknown structure, how should one *decide* between
   correlated factors, a bifactor, and a higher-order model? (Section 6: a
   two-step process that each of the earlier sections feeds into.)

Section 1 first lays out the three models and the small amount of covariance
geometry that explains everything the later sections observe.

## 1. Three structures, and where the information lives

Throughout, `J` items fall into `K` clusters (item `j`'s cluster is `k(j)`),
and all models are for continuous responses with uncorrelated residuals.

**Correlated factors (oblique).** Each item loads on its cluster factor only
(loading $a_j$); the factors correlate through $\Phi$. The model-implied
covariance between items in *different* clusters $k$ and $l$ is
$\phi_{kl}\, a_i a_j$; between items in the *same* cluster it is $a_i a_j$.

**Orthogonal bifactor.** Every item loads on one general factor ($b^{gen}_j$)
and on its cluster's group factor ($b^{grp}_j$); all factors are mutually
orthogonal. Cross-cluster covariance is $b^{gen}_i b^{gen}_j$; within-cluster
covariance is $b^{gen}_i b^{gen}_j + b^{grp}_i b^{grp}_j$.

**Higher-order (HO).** Items load on first-order factors ($\lambda_j$), which
load on a single second-order factor ($\gamma_k$, one per first-order
factor). The HO model is mathematically equivalent to an orthogonal bifactor
whose loadings obey the Schmid-Leiman (SL) proportionality constraint
(Schmid & Leiman, 1957; Yung, Thissen, & McLeod, 1999): for every item $j$
in cluster $k$,

$$b^{gen}_j = \lambda_j \gamma_k \qquad\qquad b^{grp}_j = \lambda_j \sqrt{1 - \gamma_k^2}.$$

So within each cluster the ratio $b^{grp}_j / b^{gen}_j =
\sqrt{1-\gamma_k^2}/\gamma_k$ is **constant** across items -- that constancy
is the testable signature of an HO model hiding inside a bifactor fit. The
**testlet model is mathematically equivalent to the higher-order model**:
the same SL-constrained bifactor, read with testlet or method-effect labels
in place of first-order-factor labels (Zhang & Chen, 2024).

Two geometric facts about these covariance structures drive all the
empirical behaviour below.

**Fact 1: cross-cluster blocks carry no information about a general
factor.** Both the oblique model ($\phi_{kl}\, a_i a_j$) and the bifactor
($b^{gen}_i b^{gen}_j$) imply *rank-one* cross-cluster blocks. Between-cluster
covariance therefore cannot distinguish "one general factor" from
"correlated group factors" -- ever, at any sample size.

**Fact 2: the general factor is identified only by within-cluster rank.**
The oblique model implies rank-*one* within-cluster common covariance
($a_i a_j$); the bifactor implies rank-*two*
($b^{gen}_i b^{gen}_j + b^{grp}_i b^{grp}_j$). A general factor is
empirically distinguishable from factor correlations exactly to the extent
that the within-cluster blocks are genuinely rank-two -- which requires the
general and group loadings to vary *independently* across the items of a
cluster. Two corners are degenerate: under exact SL proportionality the
rank-two block collapses to effective rank one (the HO model *equals* an
oblique model with $\Phi = \gamma\gamma'$ off the diagonal, so the bifactor
parameterization is redundant); and when loadings are near-homogeneous
within clusters, the two rank-one directions nearly coincide, so the models
are close to empirically equivalent at realistic sample sizes. Section 6
shows both corners in data.

## 2. Fitting an orthogonal bifactor

`vbfa(Y, Q, bifactor = TRUE)` fits an orthogonal bifactor model directly:
`Q` specifies the *group* design only (one column per group factor, so `K`
here counts group factors, comparable to an oblique-K model), and a general
column (every item loads) is added internally, ahead of the group columns.
`bifactor = TRUE` fixes the factor correlation matrix at the identity, and
overrides `orthogonal = FALSE`. It is bit-for-bit equivalent to the longer
form `vbfa(Y, cbind(1L, Q), orthogonal = TRUE)` -- building the general
column by hand and setting `orthogonal = TRUE` -- but `fit$Q` from the
`bifactor = TRUE` call already carries that augmented design (general column
first), so there is no need to build or store it separately.

The population below is a *genuine* bifactor in the sense of Fact 2: the
general and group loadings are drawn independently and span a wide range, so
the general-to-group ratio varies haphazardly within clusters and no
higher-order representation exists (contrast Section 4, where the loadings
are built to satisfy exactly the constant-ratio constraint).

```{r bifactor-simulate}
K <- 3; ipf <- 6; J <- K * ipf
cl <- rep(1:K, each = ipf)
set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)                 # general
for (k in 1:K) mla[cl == k, k + 1] <- round(runif(ipf, .35, .70), 2)  # group
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)  # sane communalities
sim <- sim_fa(N = 1000, mla = mla, phi = 0, rseed = 1)   # phi = 0 -> orthogonal
Y <- sim$dat
```

The fit is partially confirmatory: the general column is fully specified
(every item loads), two items per group factor are anchored, and the
remaining group entries are left to the data.

```{r bifactor-fit}
Q <- matrix(-1L, J, K)
for (k in 1:K) {
  a <- which(cl == k)[1:2]
  Q[a, ] <- 0L
  Q[a, k] <- 1L
}
fb <- vbfa(Y, Q, bifactor = TRUE)
fb$Phi                 # identity by construction, not estimated
round(fit_stats(fb), 3)
```

`fit_stats()` reads `orthogonal` (and `ld`) from the fit object itself, so a
bifactor fit needs no extra argument at the `fit_stats()` call -- and
supplying a contradictory one is an error, not a silent miscount.

A check that the fit tracks the truth, on the hard part -- the
general/group split:

```{r bifactor-verify}
c(cor = cor(fb$Lam[, 1], sim$MLA[, 1]),
  MAE_general = mean(abs(fb$Lam[, 1] - sim$MLA[, 1])))
```

With heterogeneous loadings the split is well identified and the anchored
fit recovers it: the general loadings correlate near `.98` with the truth,
with a mean absolute error under `.05`. This anchored refit -- not any sweep
-- is the estimation workhorse that everything later relies on.

## 3. Why a sweep cannot count group factors

The gain-rule sweep from `pefa()` accepts a bifactor backbone: anchor two
items per group factor and let `pefa()` pad the rest of the window with
exploratory group columns, adding the general column internally throughout.

With `bifactor = TRUE`, `pefa()`'s `Kmin` and `Kmax` count **group** factors
only; the general column is added internally and is not part of the window.
The backbone below anchors 2 of the 3 group factors (2 columns), so
`Kmin = 2` means "2 anchored group factors, general factor implicit," and
`pefa()` pads the window up with exploratory group columns. The sweep's
output gains a `K_total` column (`= K + 1`, the group count plus the
general factor); the `K` reported and selected is the group-factor count,
and `K_total` is what should be compared to a total column count from a
hand-built design.

```{r pefa-sweep}
## backbone: 2 anchored group columns (general column added internally)
Q0 <- matrix(-1L, J, 2)
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0[a, ] <- 0L
  Q0[a, k] <- 1L
}

r <- pefa(Q0, Y, Kmin = 2, Kmax = 5, bifactor = TRUE, v0 = .001,
          verbose = FALSE)
summary(r)$sweep[, c("K", "K_total", "ELBO", "BIC", "RMSEA", "CFI")]
r
```

The true design has 3 group factors plus the general factor, i.e.
`K_total = 4` -- and the sweep does **not** find it: the primary
(objective-gain) rule settles at `K = 2` group factors (`K_total = 3`), the
*lower* edge of the window, while the fit indices sit near their ceilings at
every `K` and the BIC differences are trivial relative to their scale. This
outcome is worth
understanding rather than tuning away, because it follows directly from the
geometry of Section 1. By Fact 1, the omitted group factor's cross-cluster
signature does not exist; by Fact 2, its within-cluster rank-two block is
the only evidence -- and the unspecified (`-1`) entries of the columns that
*remain* supply enough free directions to reproduce it almost exactly. Look
at what the `K = 2` candidate did with the third cluster's items:

```{r pefa-absorb}
f3 <- r$fits[[1]]
round(f3$Lam[cl == 3, ], 2)   # columns: general, group 1, group 2
```

Their group variance has been folded into inflated general-factor loadings
(compare `mla[cl == 3, 1]`), leaving almost nothing for a third group factor
to explain -- which is why the objective *falls* at every larger `K` (more
spike-and-slab bookkeeping, no new fit) and why the covariance indices
cannot tell the candidates apart. Under a different regularization or
anchoring pattern the same surplus reappears in a different disguise (split
across the two anchored group columns, say), but some absorbed,
covariance-equivalent representation is always available.

The blunt conclusion first: **an ordinary bifactor sweep -- general column
plus anchored group columns, everything else `-1` -- should not be used to
determine the number of group factors.** Its selection is not merely noisy
but systematically misleading: the absorbed candidate wins regardless of the
true `K`, and (Section 5) the truth-sized candidate inside the sweep is
itself rotationally scrambled, so its loadings cannot be inspected to detect
the problem. `print(r)` therefore flags an orthogonal sweep that lands on
the lower boundary as *weak evidence about K* rather than issuing the
generic advice to extend the window.

Two practical rules follow. First, a `boundary = "lower"` stall in a
bifactor sweep means "no *separable* extra group factor was found," never
"the backbone is the true structure." Second, the absorption route runs
through entries that are free but should not be: if an item is *known* not
to measure a group factor, encode that as a fixed `0` rather than an
exploratory `-1` -- that restores an interior selection for the factor
*count* (though still not interpretable loadings; see Section 5). And when
the real question is "how many factors are in these data at all," Section 6
shows the sweep that *does* answer it: the ordinary correlated-factor sweep,
with no general column.

(`v0 = .001` -- a single fixed spike rather than the default warm-started
path -- keeps a `Kmin:Kmax` sweep fast. The absorption behaviour above is
not an artifact of that choice; the default path lands in the same place.)

## 4. Higher-order and testlet structures

### (a) Simulate a true higher-order model directly

`sim_fa()`'s `gamma` argument generates directly from the SL equivalence of
Section 1: supply the second-order loadings and `sim_fa()` builds the SL
bifactor matrix internally and simulates from it, with the factor
correlation matrix `PHI` fixed at the identity.

```{r ho-simulate}
gamma_true <- c(.80, .70, .60)
sim_ho <- sim_fa(N = 3000, K = 3, ipf = 6, lam = .8, gamma = gamma_true, rseed = 11)
Yh <- sim_ho$dat
round(sim_ho$MLA, 2)    # the SL bifactor matrix sim_fa built internally
```

The sample is deliberately large. The quantity subsection (c) inspects is a
*ratio* of two fitted loadings, and -- by Fact 2 -- the general-versus-group
split behind that ratio is the softest part of an orthogonal bifactor
solution, so its sampling noise is large relative to the loadings
themselves. `N = 3000` keeps that noise small enough for the constant-ratio
signature to show through cleanly.

### (b) Fit a partially confirmatory orthogonal bifactor

The general column is fully specified; two items per group factor are
anchored, and the rest of the group loadings -- including the off-cluster
entries -- are left to the data.

```{r ho-fit}
Qg <- matrix(-1L, J, K)
for (k in 1:K) {
  a <- which(cl == k)[1:2]
  Qg[a, ] <- 0L
  Qg[a, k] <- 1L
}

fit <- vbfa(Yh, Qg, bifactor = TRUE)
fit
```

### (c) Check the SL constraint on the fitted loadings

A few lines of plain R -- deliberately not a package function, since this is
a descriptive diagnostic, not a fitted model. Per cluster, compute each
item's ratio of its own group loading to the general loading,
$r_j = |b^{grp}_j| / |b^{gen}_j|$, then that ratio's mean and coefficient of
variation ($CV = sd/mean$) within the cluster. A small CV means the ratios
are nearly constant across items in that cluster -- the SL-admissible
signature of a higher-order model hiding inside the bifactor fit. Averaging
the cluster CVs across all items (item-weighted, so the code generalizes to
unequal cluster sizes) gives one overall number, and the rule applied here
is:

**mean CV < .1 means the fitted bifactor is approximately a higher-order
model.**

```{r ho-sl-check}
b_gen <- fit$Lam[, 1]
cl_tab <- t(sapply(1:K, function(k) {
  idx <- which(cl == k)
  r <- abs(fit$Lam[idx, k + 1]) / abs(b_gen[idx])
  c(n = length(idx), mean = mean(r), sd = sd(r), cv = sd(r) / mean(r))
}))
rownames(cl_tab) <- paste0("cluster", 1:K)
round(cl_tab, 3)

mean_cv <- sum(cl_tab[, "n"] * cl_tab[, "cv"]) / sum(cl_tab[, "n"])
mean_cv
```

The mean CV comes out at about `.084` -- under the `.1` line, so the fitted
bifactor reads as approximately a higher-order model, exactly as it should:
the data were generated under one. The selection side of the fit is equally
clean here: every unanchored item's own-cluster loading is selected active
(posterior inclusion probability 1) and no off-cluster loading is. Note that
the pooled number passes while cluster 1 alone (CV about `.13`) sits
slightly above the line -- with only six items per cluster the per-cluster
CV is a noisy statistic, which is why the rule is stated on the
item-weighted mean. The per-cluster values are still worth a glance: a
*single* cluster far above the line with the others near zero would tell a
different story (one non-HO-like cluster) than the same mean produced
evenly.

### (d) Transform back to higher-order parameters

Given the SL relations, $\gamma_k$ and $\lambda_j$ can be recovered from
each cluster's mean ratio $\bar r_k$: $\hat\gamma_k = 1/\sqrt{1 + \bar
r_k^2}$ and $\hat\lambda_j = |b^{gen}_j| / \hat\gamma_{k(j)}$.

```{r ho-transform}
gamma_hat <- 1 / sqrt(1 + cl_tab[, "mean"]^2)
names(gamma_hat) <- paste0("cluster", 1:K)
rbind(true = gamma_true, hat = round(gamma_hat, 3))

lambda_true <- sim_ho$MLA1[cbind(seq_len(J), cl)]   # true first-order loading of each item
lambda_hat <- numeric(J)
for (k in 1:K) {
  idx <- which(cl == k)
  lambda_hat[idx] <- abs(b_gen[idx]) / gamma_hat[k]
}
c(worst = max(abs(lambda_hat - lambda_true)),   # first-order loading errors
  mean  = mean(abs(lambda_hat - lambda_true)))
```

The second-order loadings come back at about `.81`, `.69`, and `.59`
against the true `.80`, `.70`, `.60` -- each within about `.013` of the
truth. The first-order loadings (`.8` throughout, since `lam = .8` is
constant here) are recovered with a worst-case error of about `.06` and a
mean error near `.02`. When the SL constraint genuinely holds and the CV
check passes, the transform-back step is this accurate: the mean ratio per
cluster is all the information the second-order structure needs.

### (e) Group-factor effect size (Zhang & Chen, 2024, Eq. 16)

Whether the model is read as a bifactor, an HO model, or a testlet model,
the group/testlet factor's effect size quantifies how strongly its items
load on it. Following Eq. 16 of Zhang and Chen (2024, p. 11), the effect
size of special (group/testlet) factor $k$ is the **average squared
loading** on that factor,

$$D_k = \frac{\sum (\text{squared specified loadings}) + \sum (\text{squared active unspecified loadings})}{\text{number of loadings included}},$$

where an unspecified loading counts as active only if it passes a cutoff.
The package is a spike-and-slab framework rather than a Lasso one, so the
natural cutoff here is the posterior inclusion probability: an unspecified
entry (`Q == -1`) counts as active when `fit$pi > tau` with `tau = .5`;
specified entries (`Q == 1`) always count.

```{r ho-D}
tau <- .5
Dk <- sapply(1:K, function(k) {
  col <- k + 1
  included <- (fit$Q[, col] == 1) | (fit$Q[, col] == -1 & fit$pi[, col] > tau)
  sum(fit$Lam[included, col]^2) / sum(included)
})
names(Dk) <- paste0("cluster", 1:K)
round(Dk, 3)
```

The effect sizes come out at about `.22`, `.34`, and `.42`, against
population values of $(\lambda\sqrt{1-\gamma_k^2})^2 = .23, .33, .41$ -- an
essentially exact, and correctly ordered, recovery. The ordering is the
substantive reading: the *weaker* a cluster's second-order loading, the
*larger* its group/testlet effect, because whatever the general factor does
not explain of an item's common variance is exactly what its group factor
carries. Cluster 3 (true $\gamma = .60$) therefore shows the largest special
effect and cluster 1 (true $\gamma = .80$) the smallest.

### (f) All of the above in one call: `special_effects()`

Parts (c)-(e) were written out in plain R because the mathematics deserves
to be seen once. In practice, `special_effects()` computes all three
summaries from the fit directly -- per-group ratio CVs and the
item-weighted mean CV, the higher-order parameters $\hat\gamma_k$ (with
$\hat\lambda_j$ in `$items`), and the Eq.-16 effect sizes -- inferring
group membership from the design and the posterior inclusion
probabilities, and flagging any items it had to set aside:

```{r ho-special-effects}
special_effects(fit)
```

The numbers match parts (c)-(e) exactly. Note what the function does
*not* do: it takes no cutoff argument and makes no decision. The `.1`
convention for the mean CV is suggestive, and the higher-order parameters
and effect sizes are reported regardless -- reading them faithfully (a
re-expression) or loosely (a lossy summary of a genuine bifactor) is the
analyst's call, informed by the CVs.

## 5. Can a sweep recover the higher-order structure? No

Section 3 showed the absorption problem on genuine bifactor data; the same
question is sharper here, where the population is *exactly* a higher-order
model: hand the ordinary bifactor backbone (two anchored group factors, all
else `-1`) to `pefa()` on `Yh` and ask whether it finds the true 3 group
factors (`K_total = 4`).

```{r ho-sweep}
Q0h <- matrix(-1L, J, 2)
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0h[a, ] <- 0L
  Q0h[a, k] <- 1L
}
rh <- pefa(Q0h, Yh, Kmin = 2, Kmax = 5, bifactor = TRUE, v0 = .001,
           verbose = FALSE)
rh
```

It does not: the selection stalls at the lower boundary `K = 2` group
factors (`K_total = 3`), with the fit indices at their ceilings for every
candidate, for exactly the absorption reason -- SL-proportionality buys
nothing, because the omitted group factor's rank-one block is reproducible
by the free entries either way. Repeating this at several seeds and at
`N = 500` gives the same lower boundary selection every time; this is an
identification property, not bad luck. Worse, the truth-sized `K_total = 4`
candidate *inside* the sweep is not interpretable either. Its padded column
is never anchored, so nothing pins the general/group allocation: here the
extra column simply stays empty and the absorbed representation persists
(cluster 3's group variance sits in inflated general loadings plus
off-cluster spill), while under other settings and seeds the allocation
comes out rotated across columns instead. Either way, these loadings must
not be read, checked for SL proportionality, or transformed back:

```{r ho-sweep-k4}
f4 <- rh$fits[[which(rh$sweep$K_total == 4)]]
round(f4$Lam[cl == 3, ], 2)     # cluster-3 rows: compare sim_ho$MLA[cl == 3, ]
```

All interpretation must instead happen on an *anchored refit* at the chosen
`K` -- the design of Section 4(b), whose fitted columns match the true SL
matrix with congruence above `.99`. Anchors pin the rotation; a sweep's
padded columns, by construction, never have them. That leaves the question
Section 6 answers: if sweeps cannot decide the bifactor question, what can?

## 6. Which structure? The two-step process

Sections 3 and 5 established what sweeps over *bifactor* candidates cannot
do. This section assembles what the tools *can* do into the two-step
process of Chen and Jin (2026b; the "accompanying manuscript" throughout
this section), run in miniature on the practical situation: data in hand,
clusters suspected, structure -- correlated factors, bifactor, or
higher-order -- unknown. Everything below is one replication per
population at `N = 1000` and is illustrative only: operating
characteristics come from the manuscript's simulations, and this
section postdates the analyses reported there (which were generated with
vbpm 0.7.1; this vignette was built with version
`r as.character(packageVersion("vbpm"))`).

### Step 1: anchored oblique sweeps from both backbones

Run the plain correlated-factor sweep of the `pefa` vignette -- no general
column, no `orthogonal = TRUE` (the sweep, its gain cuts, and the
reproduction statistic below are those of Chen and Jin, 2026a). This asks
only: *how many cluster-level
factors are there?* For that question an omitted factor produces honest
misfit (its within-cluster block is not reproducible by other clusters'
columns), so interior selections are trustworthy.

One selection is not yet a result, though. The sweep's backbone -- how the
anchor rows are treated -- is an algorithmic convention with no
substantive content, so run the sweep from both conventions and ask
whether they agree: BPC-type anchors (the anchor rows' off-column entries
fixed to zero, `Q0_ob` below) and LPC-type anchors (the 1s alone, every
other entry free). Do the same with the selection cut: the 20% ELBO-gain
cut is the primary rule, the laxer 10% cut the sensitivity. Run all of it
on both datasets of this vignette -- the genuine bifactor `Y` of Section 2
and a fresh higher-order sample at the same `N`:

```{r step1}
sim_ho2 <- sim_fa(N = 1000, K = 3, ipf = 6, lam = .8, gamma = gamma_true,
                  rseed = 11)
Yh2 <- sim_ho2$dat

Q0_ob <- matrix(-1L, J, 2)                       # BPC-type backbone
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0_ob[a, ] <- 0L
  Q0_ob[a, k] <- 1L
}
Q0_lp <- matrix(-1L, J, 2)                       # LPC-type backbone
for (k in 1:2) Q0_lp[which(cl == k)[1:2], k] <- 1L

## both cuts in one call; pefa() applies its own eligibility and gap rules,
## which a direct select_K_elbow() call on the sweep table would bypass
sweep_at <- function(Q0, Y)
  pefa(Q0, Y, Kmin = 2, Kmax = 6, verbose = FALSE,
       cuts = c(primary = 20, sensitivity = 10))
cut_picks <- function(s) {
  sel <- s$selection
  keep <- sel$criterion == "elbo" & sel$form == "gain" &
          sel$eligibility_scope == "converged+fit"
  stats::setNames(sel$selected_K[keep], paste0("K_", sel$cut_value[keep]))
}
sweeps <- list(bif_BPC = sweep_at(Q0_ob, Y),   bif_LPC = sweep_at(Q0_lp, Y),
               HO_BPC  = sweep_at(Q0_ob, Yh2), HO_LPC  = sweep_at(Q0_lp, Yh2))
t(sapply(sweeps, cut_picks))
```

All four sweeps land at `K = 3` -- interior to the window -- under the
primary cut, the sensitivity cut agrees everywhere, and both selections
are correct in first-order terms: the HO data genuinely are three
correlated factors (Section 1), and the bifactor data have three clusters
whose general factor the oblique model absorbs into the factor
correlations (Fact 1 -- cross-cluster blocks cannot tell the difference).
Note what step 1 does **not** claim: it says nothing about whether a
general factor should be carved out. That is step 2's question.

### The stability analysis: adjacent-count reproduction

Backbone and cut agreement are cheap consistency checks. The direct
evidence that a selected solution is worth delivering is *reproduction*:
the `K` columns just selected should reappear, nearly unchanged, inside
the `(K+1)`-factor solution the sweep has already fitted. If they do, the
extra candidate factor is surplus sitting beside a stable structure; if
instead the columns reorganize, the selection was a score that moved, not
a structure that persisted. Match each column of the `K`-factor fit to its
best counterpart in the `(K+1)`-factor fit by Tucker congruence, without
reuse, and take the minimum:

```{r stability}
tucker <- function(a, b) abs(sum(a * b) / sqrt(sum(a^2) * sum(b^2) + 1e-12))
reproduce_up <- function(s, k) {
  fA <- s$fits[[as.character(k)]]
  fB <- s$fits[[as.character(k + 1)]]
  used <- integer(0)
  phis <- numeric(ncol(fA$Lam))
  for (i in seq_len(ncol(fA$Lam))) {
    cong <- sapply(seq_len(ncol(fB$Lam)), function(j)
      if (j %in% used) -1 else tucker(fA$Lam[, i], fB$Lam[, j]))
    used <- c(used, which.max(cong))
    phis[i] <- max(cong)
  }
  min(phis)
}
round(sapply(sweeps, reproduce_up, k = 3), 3)
```

All four values sit at or near `1`, far above the `.85` threshold the
manuscript calibrates for this check (a provisional convention, like every
cutoff in this vignette). Both datasets therefore *deliver* a stable
three-factor structure, and the process moves on to step 2.

The other branches matter as much as this one. When the two backbones
select adjacent counts and the smaller solution reproduces inside the
larger, the delivery is a stable core with count uncertainty, and step 2
runs at both counts. When neither holds -- selections apart by more than
one, or columns that fail to reproduce -- no stable structure is
delivered, and the honest report is exactly that: non-delivery, with the
sweep evidence shown. Imposing a structure anyway, because theory demands
one, is a substantively specified sensitivity analysis rather than a
delivery, and should be labeled as such.

### Step 2: anchored oblique-K versus anchored bifactor-(K+1), by BIC

At the delivered `K`, fit two *anchored, confirmatory-leaning* models with
the same anchors: the oblique `K`-factor model, and the orthogonal bifactor
with the same group design plus a fully specified general factor. Compare
their `BIC` from `fit_stats()`. This is the payoff of the `bifactor = TRUE`
argument: the oblique-K and bifactor-K models are now literally the same
design matrix with a switch, rather than two hand-built designs that have
to be kept in sync. And as in step 1, the anchoring depth is an
algorithmic convention, so run the comparison at both depths -- BPC-type
(`Q3`) and LPC-type (`Q3_lp`) -- and ask whether the two `diff` signs
agree:

```{r step2}
Q3 <- matrix(-1L, J, 3)
for (k in 1:3) {
  a <- which(cl == k)[1:2]
  Q3[a, ] <- 0L
  Q3[a, k] <- 1L
}
Q3_lp <- matrix(-1L, J, 3)
for (k in 1:3) Q3_lp[which(cl == k)[1:2], k] <- 1L

step2 <- function(Y, Q, ...) {                 # `...` reaches BOTH arms alike
  b3 <- fit_stats(vbfa(Y, Q, ...))["BIC"]
  b4 <- fit_stats(vbfa(Y, Q, bifactor = TRUE, ...))["BIC"]
  c(oblique_K = unname(b3), bifactor_K1 = unname(b4), diff = unname(b4 - b3))
}
round(rbind(bifactor_BPC = step2(Y, Q3),   bifactor_LPC = step2(Y, Q3_lp),
            HO_BPC       = step2(Yh2, Q3), HO_LPC       = step2(Yh2, Q3_lp)), 1)
```

The comparison resolves both datasets correctly at `N = 1000`, with the
depths agreeing in sign. On the genuine bifactor data the bifactor wins at
both depths (`diff` about `-59` and `-9`; the LPC margin is the smaller
one because the freer depth lets the oblique model absorb more of the
general variance). Read that `-9` against the small-margin warning that
closes this section: a margin of single-digit BIC points is thin, and
the LPC arm *by itself* would license nothing better than a provisional
call. What carries the conclusion here is the agreement of the two depths
in sign together with the substantial BPC margin -- and, in Step 3 below,
a mean CV far from the HO boundary. Had the depths disagreed, or had both
margins been this thin, the honest report would be that the comparison did
not resolve. On the higher-order data the oblique model wins at both
depths (`diff` about `+103` and `+114`) -- correctly so, since by the SL
equivalence the bifactor parameterization of an HO model is redundant, and
BIC refuses to pay 18 extra loadings for it. When the depths *disagree* in
sign, the comparison has not resolved anything: withhold the structural
conclusion and look for the reason (the absorbed-dependence subsection
below shows the usual one).

Use **BIC, not the ELBO, as the operational criterion in this
comparison**; the ELBO serves as a sensitivity check. The two designs
place spike-and-slab prior mass on different numbers of entries, so their
ELBOs differ by prior bookkeeping as well as by fit; in the manuscript's
simulations the ELBO kept preferring the oblique model in regimes where
BIC had already (correctly) flipped to the bifactor. Here the two agree in
direction on both datasets, which is worth one sentence in a report and no
more. The ELBO is the right currency *within* a sweep of same-shaped
designs; BIC is the right currency between these two differently-shaped
ones.

### Step 3: resolve the winner into a final structure

**If the bifactor won**, run the SL check of Section 4(c) on the bifactor
fit. A mean CV under `.1` says the bifactor is SL-degenerate -- collapse it
to the more parsimonious HO model via the transform of Section 4(d). A
large mean CV says the general factor is genuinely its own thing:

```{r step3-bifactor}
f4b <- vbfa(Y, Q3, bifactor = TRUE)
se  <- special_effects(f4b)
se$mean_cv
se$groups[, c("group", "mean_ratio", "cv", "gamma", "D")]
```

Here it is about `.39` -- far above the line, so these data keep their full
bifactor: correctly so, since their generating loadings were drawn with no
proportionality constraint.

**If the oblique model won**, the general factor question is settled
(no separable general factor) -- but a second-order description may still
be available for free. Under an HO population, $\Phi$ has the rank-one
structure $\gamma\gamma'$ off the diagonal, so the second-order loadings
can be read out of the fitted factor correlations by the classical triad
(Spearman) relations, e.g. $\hat\gamma_1 =
\sqrt{\phi_{12}\phi_{13}/\phi_{23}}$:

```{r step3-oblique}
f3h <- vbfa(Yh2, Q3)
P <- f3h$Phi
round(P, 2)
gamma_triad <- c(sqrt(P[1, 2] * P[1, 3] / P[2, 3]),
                 sqrt(P[1, 2] * P[2, 3] / P[1, 3]),
                 sqrt(P[1, 3] * P[2, 3] / P[1, 2]))
round(gamma_triad, 2)             # true: .80 .70 .60
```

The triads recover the second-order loadings to within a few hundredths --
a second, independent route to the HO structure that never fits an
orthogonal model at all, and that can cross-validate the SL route of
Section 4 when both are applicable. (With more than three first-order
factors, replace the triads by a one-factor model fit to $\hat\Phi$; with
exactly three, the triads are that model's exact solution. Coherent triads
-- all inside the unit interval, stable across item subsets -- support the
HO reading; incoherent ones say $\Phi$ is just a correlation matrix, and
the structure is plain correlated factors.)

### The hazard step 2 cannot see: absorbed local dependence

The manuscript's sharpest warning deserves its own demonstration. Take the
higher-order population -- where the oblique model just won by a hundred
BIC points -- and contaminate it with a single residual correlation of
`.25` between two non-anchor items of the first cluster, the kind of extra
dependence a pair of same-stem items produces:

```{r doublet-simulate}
bg <- .8 * gamma_true[cl]
bs <- .8 * sqrt(1 - gamma_true^2)[cl]
L_sl <- cbind(bg, matrix(0, J, K))
for (k in 1:K) L_sl[cl == k, k + 1] <- bs[cl == k]
S0 <- tcrossprod(L_sl) + diag(1 - rowSums(L_sl^2))  # HO population, SL form
Sd <- S0
Sd[3, 4] <- Sd[3, 4] + .25
Sd[4, 3] <- Sd[4, 3] + .25                          # one residual pair
set.seed(12)
Yd <- scale(matrix(rnorm(1000 * J), 1000, J) %*% chol(Sd))
```

Step 1 looks almost clean -- both backbones still select `K = 3` under the
primary cut, and the three factors still reproduce -- but the sensitivity
cut now points one count higher on both backbones:

```{r doublet-step1}
sd_b <- sweep_at(Q0_ob, Yd)
sd_l <- sweep_at(Q0_lp, Yd)
t(sapply(list(BPC = sd_b, LPC = sd_l), cut_picks))
round(c(BPC = reproduce_up(sd_b, 3), LPC = reproduce_up(sd_l, 3)), 3)
```

And step 2, run mechanically at `K = 3`, is now decisively wrong at *both*
depths -- the bifactor is the only model on the menu that can reach the
extra within-cluster covariance, so BIC buys it:

```{r doublet-step2}
round(rbind(BPC = step2(Yd, Q3), LPC = step2(Yd, Q3_lp)), 1)
```

A bifactor win of hundreds of BIC points, on data whose general structure
is exactly the higher-order model the oblique fit represents. The
manuscript's adversarial population shows this error *grows* with sample
size; more data do not fix it, because the bifactor genuinely fits better
-- it is just not the right explanation.

This example *happens* to leave a trace at the count level -- the
sensitivity cut moved while the primary cut did not -- and that trace is
worth following. But do not generalize it into a safeguard. Cut
disagreement is not a local-dependence detector: it fires for any thin
extra dimension, of which absorbed dependence is only one kind, and it
stays silent whenever the extra covariance is too small to bend a gain
curve. The manuscript's adversarial populations are precisely the harder
case, in which the absorbed dependence sits at the *selected* count and
every count-level indicator -- both cuts, the reproduction statistic, the
convergence flags -- reads clean while step 2 still hands the win to the
bifactor. Nothing in Step 1 is designed to catch that; what catches it is
looking, which is why residual inspection belongs in the workflow
unconditionally rather than as a response to a warning.

Wherever the suspicion comes from, the cheapest place to look is the
`(K+1)`-factor solution the sweep already retained -- no refitting
required:

```{r doublet-inspect}
round(sd_b$fits[["4"]]$Lam[cl == 1, ], 2)   # cluster-1 rows at K = 4
```

The four-factor solution splits the first cluster: the residual pair
(items 3 and 4) keeps the cluster's column nearly to itself with loadings
around `.95`, while the cluster's other four items move onto the new
column. Two items dominating a factor is a doublet signature, not a
substantive dimension.

The repair is to model the pair as what it is -- a residual edge -- and
rerun *the same* comparison. Two requirements make this informative rather
than decorative. It has to be a residual, not a factor: an extra two-item
factor is a different model, and it is a *differently* different model in
each arm, because the oblique arm would estimate that factor's
correlations with the others while `bifactor = TRUE` forces the new column
orthogonal. And it has to be applied identically to both competitors,
which is what passing the same arguments through `step2()` guarantees:

```{r doublet-model}
Qe <- matrix(0L, J, J)
Qe[3, 4] <- Qe[4, 3] <- 1L                 # exactly one free residual edge
round(rbind(BPC = step2(Yd, Q3,    ld = TRUE, Qe = Qe, tolVal = 1e-3),
            LPC = step2(Yd, Q3_lp, ld = TRUE, Qe = Qe, tolVal = 1e-3)), 1)
```

With the pair modeled, the oblique model wins at both depths (`diff` about
`+82` and `+80`) -- a swing of several hundred BIC points from the
contaminated comparison, landing near the margins the *uncontaminated*
higher-order data produced in Step 2 (`+103` and `+114`). The entire
bifactor preference was the absorbed residual pair. (The loosened
`tolVal` is bookkeeping: the LD branch's tail converges slowly on these
data, and at the default tolerance the BIC values are the same to a tenth
of a point but the fits report unmet convergence.)

Read this as a sensitivity analysis, not a test. It answers one narrowly
posed question -- does the bifactor preference survive when this specific
dependence is modeled in both arms? -- and a `Qe` aimed at the wrong pair
answers nothing. The moral is the manuscript's, and it is unconditional: a
step-2 bifactor win is credible only after local dependence has been
looked for, whether or not anything at the count level asked you to look.

### The process in six steps, and its limits

The process this vignette has just walked, in the order the manuscript
prescribes:

1. **Sweep**: anchored oblique sweeps from both backbones (`pefa()`
   without a general column); ELBO gain at the 20% cut primary, the 10%
   cut as sensitivity. Trust interior selections; treat boundaries as
   instructions to extend the window.
2. **Check structure**: adjacent-count reproduction on the sweep's own
   retained fits (matched congruence of the selected solution inside its
   neighbor).
3. **Deliver**: a stable structure, a stable core with count uncertainty,
   or no stable structure (non-delivery) -- reported as such, with step 2
   running at both counts when the count is uncertain and not at all
   after non-delivery.
4. **Compare**: at every delivered count, anchored oblique-`K` versus
   anchored bifactor-`(K+1)` by `BIC`, at both depths.
5. **Diagnose**: withhold the conclusion if the comparison reverses across
   depth or count, or if both margins are thin; treat the ELBO as a
   sensitivity. Before accepting *any* bifactor win, look for local
   dependence -- inspect the adjacent solution for doublet signatures and
   refit both competitors with the suspect residual edges freed
   (`ld = TRUE`, same `Qe` in both arms). A clean count-level picture is
   not evidence that there is nothing to find.
6. **Interpret**: within the winner, the SL mean-CV check (under `.1`:
   collapse to HO; otherwise keep the bifactor) or the triad check of
   $\hat\Phi$ (coherent: a second-order description is available;
   otherwise correlated factors).

Every population in this vignette ends at the right structure: the
heterogeneous bifactor of Section 2 goes sweep -> `K = 3` -> delivered ->
bifactor wins at both depths -> CV `.39` -> **bifactor**; the higher-order
data go sweep -> `K = 3` -> delivered -> oblique wins at both depths ->
coherent triads (`.76/.72/.62`) -> **higher-order**; and the contaminated
variant is caught at the diagnosis step rather than swallowed at the
comparison step -- because step 5 was actually carried out, not because
step 2 protested.

The honest limits, from a 3-populations x 3-sample-sizes x 3-seeds
simulation with this design (18 items, 3 clusters):

| population | N = 500 | N = 1000 | N = 3000 |
|---|---|---|---|
| bifactor, heterogeneous loadings | bifactor wins 3/3 | bifactor 3/3 | bifactor 3/3 (margin grows) |
| bifactor, near-homogeneous loadings | oblique 3/3 | oblique 3/3 | bifactor 3/3 |
| higher-order | oblique 3/3 | oblique 3/3 | oblique 3/3 (margin flat) |

The middle row is Fact 2's degenerate corner: when general and group
loadings barely vary within clusters, a true bifactor is nearly empirically
equivalent to correlated factors, and step 2 only resolves it with a large
sample. Two mitigations: a *small* `|diff|` in step 2 is itself the warning
sign (the decisive wins above come with margins that grow with `N`, the
redundant HO comparison stays flat at roughly the parameter penalty); and
in that regime the parsimonious miss is cheap, because the two descriptions
of the data are nearly interchangeable by construction. What the process
never does is what Sections 3 and 5 warned against: asking a bifactor
*sweep* to make any of these calls.

## 7. Caveats

The SL check is descriptive, not a hypothesis test: there is no p-value
that certifies "this bifactor is really an HO model," and the
`mean CV < .1` rule is a working convention, not a distributional result.
Sampling error alone makes the CV strictly positive even when the data are
generated exactly under an HO model, as they are in Section 4 -- and the
loading *ratio* it is built on inherits the softness of the
general-versus-group split (Fact 2), which is why that section runs at a
generous `N`; at a few hundred observations the same population can land
above the line on noise alone, and an occasional item whose posterior
settles on the wrong group factor will inflate the pooled number further.
So read the mean CV together with the per-cluster (or per-item) ratios
rather than as a decision on its own. The same softness bounds step 2 of
Section 6, as its table shows, and the gain cuts, the `.85` reproduction
threshold, and the `.1` CV line are all working conventions of the same
kind: the accompanying manuscript calibrates them, but none is a
distributional guarantee. The local-dependence check of Section 6 is
softer still -- an exploratory sensitivity whose reach is limited to the
residual edges you thought to free, with no accompanying claim that a
bifactor win surviving it is therefore dependence-free. And the
diagnostics run both ways -- a large mean CV, when it reflects genuinely
non-constant ratios rather than noise or an isolated item, is informative
in its own right: it means the bifactor
solution is richer than any higher-order representation of it, and
collapsing it to a single second-order factor per cluster would discard
real structure.

## 8. References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical
  factor solutions. *Psychometrika*, 22(1), 53-61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113-128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models
  for special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5),
  208-229. \doi{10.1177/01466216241261704}
* Chen, J. (2021). A generalized partially confirmatory factor analysis
  framework with mixed Bayesian Lasso methods. *Multivariate Behavioral
  Research*, 57(6), 879-894. \doi{10.1080/00273171.2021.1925520}
* Jin, Y., & Chen, J. (2025). Regularized variational approximation for
  partially confirmatory factor analysis. *Structural Equation Modeling*,
  32(3), 437-449. \doi{10.1080/10705511.2024.2432612}
* Chen, J., & Jin, Y. (2026a). Recovering latent structures after
  variational Bayesian variable selection: Fit assessment and
  factor-number selection in partially exploratory factor analysis.
  *arXiv preprint* arXiv:2607.07159. -- the sweep, gain-cut, and
  reproduction machinery of Step 1.
* Chen, J., & Jin, Y. (2026b). *When factor count is uncertain: Stable
  structure before the bifactor decision.* Manuscript in preparation. --
  the two-step process of Section 6, its simulation-based operating
  characteristics, the `.85` reproduction convention, and the
  absorbed-dependence hazard it warns about.
