---
title: "Bifactor, higher-order, and testlet structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor, higher-order, and testlet structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette collects three related structures that all attach "extra"
factors to a set of primary or group factors: the orthogonal **bifactor**
model, sweeping a **factor-number window** over bifactor candidates with
`pefa()`, and the connection between bifactor, **higher-order**, and
**testlet** parameterizations via the Schmid-Leiman transformation. All data
here are simulated; nothing is loaded from disk.

## 1. Orthogonal bifactor basics

`orthogonal = TRUE` fixes the factor correlation matrix at the identity. With
a design whose first column is a general factor (every item loads) plus group
columns, this is an identified orthogonal bifactor model.

```{r bifactor-simulate}
sim <- sim_fa(N = 500, K = 3, ipf = 6, lam = .7, lac = .3, rseed = 1)
Y <- sim$dat

## two anchors per factor; everything else left to the data
Q <- matrix(-1L, ncol(Y), 3)
for (k in 1:3) {
  a <- which(rep(1:3, each = 6) == k)[1:2]
  Q[a, ] <- 0L
  Q[a, k] <- 1L
}
```

Prepend a general column that every item loads on -- `Qb <- cbind(1L, Q)` --
and fit with `orthogonal = TRUE`:

```{r bifactor-fit}
Qb <- cbind(1L, Q)
fb <- vbfa(Y, Qb, orthogonal = TRUE)
fb$Phi                 # identity by construction, not estimated
round(fit_stats(fb), 3)
```

`fit_stats()` reads `orthogonal` (and `ld`) from the fit object itself, so a
bifactor fit needs no extra argument at the `fit_stats()` call -- and
supplying a contradictory one is an error, not a silent miscount.

## 2. A pefa sweep over bifactor candidates

The same gain-rule sweep from `pefa()` works with a bifactor backbone: keep the
general column fixed and specified, anchor two items per group factor, and let
`pefa()` pad the rest of the window with exploratory group columns.

```{r pefa-sweep}
cl <- rep(1:3, each = 6)

## backbone: general column (every item loads) + 2 anchored group columns
Q0 <- matrix(-1L, ncol(Y), 3)
Q0[, 1] <- 1L                              # general factor
for (k in 1:2) {
  a <- which(cl == k)[1:2]
  Q0[a, 2:3] <- 0L
  Q0[a, k + 1] <- 1L
}

r <- pefa(Q0, Y, Kmin = 3, Kmax = 5, orthogonal = TRUE, v0 = .001,
          verbose = FALSE)
summary(r)$selection
summary(r)$comparison[, c("K", "ELBO", "ELBO_gain", "BIC", "BIC_gain", "CFI")]
```

`Kmin = 3` is the backbone as specified (general + 2 group factors); `Kmax = 5`
lets `pefa()` try one and two additional exploratory group factors. Here the
gain rule keeps the backbone at `K = 3`, matching the three-cluster design the
data were simulated from.

## 3. Higher-order CFA, Schmid-Leiman, and the bifactor connection

A higher-order (HO) model has items loading on first-order factors (loadings
$\lambda_j$), which in turn load on a single second-order factor (loadings
$\gamma_k$, one per first-order factor $k$). This is mathematically equivalent
to an **orthogonal bifactor** model whose loadings satisfy the Schmid-Leiman
(SL) proportionality constraint: for every item $j$ in cluster $k$,

$$b^{general}_j = \lambda_j \gamma_k \qquad\qquad b^{group}_j = \lambda_j \sqrt{1 - \gamma_k^2}.$$

So within each cluster the ratio $b^{group}_j / b^{general}_j =
\sqrt{1-\gamma_k^2}/\gamma_k$ is **constant** across items -- that constancy
is the testable signature of an HO model hiding inside a bifactor fit. The
testlet model is the same structure read as method/testlet effects (Zhang &
Chen, 2024).

### (a) Simulate from a true HO model

Rather than simulate an HO model and hope some estimator recovers it, we build
the SL-equivalent bifactor loading matrix directly from $\lambda$ and $\gamma$
and hand it to `sim_lvm()`. Passing `phi = 0` together with a supplied `mla`
makes the factor correlation matrix the identity -- an orthogonal bifactor,
exactly the SL parameterization -- so the general and group factors are
generated as uncorrelated from the start.

```{r ho-simulate}
K <- 3; ipf <- 6; J <- K * ipf
gamma_true <- c(.80, .70, .60)          # second-order (general) loadings

set.seed(1)
lambda_true <- round(runif(J, .6, .8), 2)   # first-order loadings, varied a bit
cl <- rep(1:K, each = ipf)                  # cluster (first-order factor) of each item

## column 1: general factor, b_general_j = lambda_j * gamma_k(j)
## columns 2:4: group factors, blockwise b_group_j = lambda_j * sqrt(1 - gamma_k^2)
mla <- matrix(0, J, K + 1)
mla[, 1] <- lambda_true * gamma_true[cl]
for (k in 1:K) {
  idx <- which(cl == k)
  mla[idx, k + 1] <- lambda_true[idx] * sqrt(1 - gamma_true[k]^2)
}

sim <- sim_lvm(N = 800, mla = mla, phi = 0, rseed = 11)
Y <- sim$dat
round(sim$PHI, 2)      # identity: phi = 0 with a supplied mla
```

### (b) Fit a confirmatory orthogonal bifactor

The design is fully confirmatory: column 1 (general) is `1` for every item;
columns 2:4 (group) are `1` within their own cluster and `0` elsewhere.

```{r ho-fit}
Q <- matrix(0L, J, K + 1)
Q[, 1] <- 1L
for (k in 1:K) Q[cl == k, k + 1] <- 1L

fit <- vbfa(Y, Q, orthogonal = TRUE)
fit
```

### (c) Check the SL constraint on the fitted loadings

A few lines of plain R -- deliberately not a package function, since this is a
descriptive diagnostic, not a fitted model. Per cluster, compute the ratio of
the group to the general loading for every item, then its mean and coefficient
of variation (CV = sd/mean). A near-zero CV means the bifactor solution is
SL-admissible, i.e. approximately a higher-order model.

```{r ho-sl-check}
b_gen <- fit$Lam[, 1]
ratio_tab <- t(sapply(1:K, function(k) {
  idx <- which(cl == k)
  r <- abs(fit$Lam[idx, k + 1]) / abs(b_gen[idx])  # abs(): guards an arbitrary sign flip
  c(mean = mean(r), sd = sd(r), cv = sd(r) / mean(r))
}))
rownames(ratio_tab) <- paste0("cluster", 1:K)
round(ratio_tab, 3)
```

The CVs are small (well under 0.2) relative to the means, so the fitted
bifactor loadings are close to SL-proportional within each cluster -- as they
should be, since the data were generated that way.

### (d) Transform back to HO parameters

Given the SL relations, $\gamma_k$ and $\lambda_j$ can be recovered from the
mean ratio in each cluster: $\hat\gamma_k = 1/\sqrt{1 + \bar r_k^2}$ and
$\hat\lambda_j = b^{general}_j / \hat\gamma_{k(j)}$.

```{r ho-transform}
gamma_hat <- 1 / sqrt(1 + ratio_tab[, "mean"]^2)
names(gamma_hat) <- paste0("cluster", 1:K)
rbind(true = gamma_true, hat = round(gamma_hat, 3))

lambda_hat <- numeric(J)
for (k in 1:K) {
  idx <- which(cl == k)
  lambda_hat[idx] <- abs(b_gen[idx]) / gamma_hat[k]
}
max(abs(lambda_hat - lambda_true))   # worst-case first-order loading error
```

$\hat\gamma$ recovers the true second-order loadings `c(.80, .70, .60)` to
within about 0.08, and the worst first-order loading is off by less than 0.1 --
both well within the range expected from a single sample of `N = 800`.

### (e) Testlet effect size (Zhang & Chen, 2024)

Whether the model is read as a bifactor, an HO model, or a testlet model, the
group/testlet factor's share of an item's variance quantifies how much local
structure the general factor does *not* absorb. Per cluster, define

$$\omega_k = \frac{\sum_j (b^{group}_j)^2}{\sum_j \left[(b^{general}_j)^2 + (b^{group}_j)^2 + \psi_j\right]}$$

with $\psi_j = 1/\text{PsiInv}_j$ the residual (unique) variance.

```{r ho-omega}
psi <- 1 / fit$PsiInv
omega <- sapply(1:K, function(k) {
  idx <- which(cl == k)
  bg <- fit$Lam[idx, 1]; bs <- fit$Lam[idx, k + 1]
  sum(bs^2) / sum(bg^2 + bs^2 + psi[idx])
})
names(omega) <- paste0("cluster", 1:K)
round(omega, 3)
```

Cluster 3 (true $\gamma = .60$, the weakest second-order loading) has the
largest testlet effect and cluster 1 (true $\gamma = .80$) the smallest --
the less a cluster's items are explained by the general factor, the more
residual local structure is left for the group factor to carry.

### Caveats

This check is descriptive, not a hypothesis test: there is no p-value or
threshold that certifies "this bifactor is really an HO model." Sampling error
alone makes the CV strictly positive even when the data are generated exactly
under an HO model, as they are here. And the diagnostic runs both ways -- a
clearly non-constant ratio (a large CV, items in the same cluster with wildly
different group-to-general ratios) is informative in its own right: it means
the bifactor solution is genuinely richer than any higher-order representation
of it, and collapsing it to a single second-order factor per cluster would
discard real structure.

## 4. References

* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5), 208-229.
  \doi{10.1177/01466216241261704}
* Chen, J. (2021). A generalized partially confirmatory factor analysis
  framework with mixed Bayesian Lasso methods. *Multivariate Behavioral
  Research*, 57(6), 879-894. \doi{10.1080/00273171.2021.1925520}
* Jin, Y., & Chen, J. (2025). Regularized variational approximation for
  partially confirmatory factor analysis. *Structural Equation Modeling*,
  32(3), 437-449. \doi{10.1080/10705511.2024.2432612}
