---
title: "Bifactor Structures and PEFA Sweeps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor Structures and PEFA Sweeps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette covers the bifactor side of `vbpm`: what a bifactor model is,
how a `pefa()` sweep is read when the factor count is uncertain, and how a
later oblique-versus-bifactor comparison is built out of the sweep that
preceded it. The sweep object itself is walked in `vignette("pefa")`; nothing
here re-explains its tables.

## Three related structures

Suppose `J` items fall into `K` clusters.

**Correlated factors.** Each item loads on its cluster factor, and the cluster
factors correlate through $\Phi$.

**Orthogonal bifactor.** Every item loads on a general factor and on one group
factor; general and group factors are mutually orthogonal.

**Higher-order.** Items load on first-order factors, which load in turn on a
second-order factor. Its Schmid-Leiman representation is the orthogonal
bifactor with

$$
b^{gen}_j=\lambda_j\gamma_k,
\qquad
b^{grp}_j=\lambda_j\sqrt{1-\gamma_k^2}
$$

for item $j$ in cluster $k$, so $|b^{grp}_j|/|b^{gen}_j|$ is constant within a
cluster under exact higher-order proportionality. Testlet and method factors
use the same general-plus-special geometry under a different substantive
reading.

In `vbfa()` and `pefa()`, `bifactor = TRUE` makes `K` the number of **group**
factors: the design matrix supplies only the group block and the general column
is added internally, so a fitted matrix has `K + 1` columns. Every
between-candidate comparison a sweep performs removes that general column
first and works on the `K` group columns, which is what makes an ordinary and a
bifactor sweep of the same data structurally comparable.

## A bifactor population and two backbones

Three clusters of six items, each item loading on a general factor and on its
own group factor.

```{r population}
K <- 3
ipf <- 6
J <- K * ipf
cluster <- rep(seq_len(K), each = ipf)

set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)
for (k in seq_len(K)) {
  mla[cluster == k, k + 1] <- round(runif(ipf, .35, .70), 2)
}
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)

sim <- sim_fa(N = 800, mla = mla, phi = 0, rseed = 1)
Y <- sim$dat
```

Two anchors per backbone factor, expressed as the two designs `vbpm`
documentation uses throughout.

* **AO (anchor-only)** specifies each anchor's intended cell as `1` and leaves
  that anchor's cells on the *other* backbone columns unspecified (`-1`).
* **AZ (anchor-zero)** additionally fixes those cross-backbone anchor cells to
  `0`.

```{r backbones}
Q0_AO <- matrix(-1L, J, 2)
for (k in 1:2) Q0_AO[which(cluster == k)[1:2], k] <- 1L

Q0_AZ <- Q0_AO
for (k in 1:2) {
  anchors <- which(cluster == k)[1:2]
  Q0_AZ[anchors, ] <- 0L
  Q0_AZ[anchors, k] <- 1L
}

Q0_show <- cbind(Q0_AO[1:8, ], Q0_AZ[1:8, ])
colnames(Q0_show) <- c("AO_1", "AO_2", "AZ_1", "AZ_2")
Q0_show
```

AZ is the stronger claim: it asserts that an anchor of cluster 1 has no loading
on backbone factor 2. AO only asserts where the anchor *does* load. Both
backbones cover two of the three clusters, so the third has to be found by the
sweep.

## Step 1: ordinary oblique sweeps

```{r step1}
AO <- pefa(Q0_AO, Y, Kmin = 2, Kmax = 5, v0 = .001, verbose = FALSE)
AZ <- pefa(Q0_AZ, Y, Kmin = 2, Kmax = 5, v0 = .001, verbose = FALSE)

AO$sweep[, c("K", "ELBO", "AIC", "BIC", "SRMR", "CFI")]
AZ$sweep[, c("K", "ELBO", "AIC", "BIC", "SRMR", "CFI")]
```

```{r step1-transitions}
AO$transitions[, c("K_from", "K_to", "ELBO_gain_pct", "BIC_gain_pct",
                   "phi_min", "rmsd_max", "unmatched_ssl", "collision")]
AZ$transitions[, c("K_from", "K_to", "ELBO_gain_pct", "BIC_gain_pct",
                   "phi_min", "rmsd_max", "unmatched_ssl", "collision")]
```

Both backbones put their only positive gain on the `2 -> 3` step and both put
`unmatched_ssl` near 2.6-2.9 there, which is a full column's worth against the
`ssl()` values below. After that they part company slightly. AO's BIC is
minimized at `K = 3` (35196.8) and its later ELBO gains are -6.9% and -8.8%.
AZ's BIC is minimized at `K = 4` (35209.3, against 35227.0 at `K = 3`), and its
`3 -> 4` ELBO gain is -0.5%, close enough to zero that a stricter reading of
the same path would hesitate there.

```{r step1-persistence}
round(AO$persistence$phi, 3)
round(AZ$persistence$phi, 3)
ssl(AO)[c("3", "4")]
ssl(AZ)[c("3", "4")]
```

Persistence separates the two backbones more than the criteria do. AO's
three-factor columns survive into `K = 4` and `K = 5` at 0.988 and 0.987; AZ's
survive at 0.935 and 0.922, and its `4 -> 5` cell falls to 0.834. In both
backbones the fourth column is small --- `ssl()` puts it at 0.36 (AO) and 0.38
(AZ) against three columns above 2.2 --- so the four-factor candidate is the
three-factor one plus a thin extra column, not a different structure.

```{r step1-reading}
## Analysis-side reading, not a package rule. Variant: full look-ahead with no
## fallback -- the smallest K_from whose own and every later adjacent ELBO gain
## is under 20% of the largest positive gain, NA_integer_ when none qualifies.
vapply(list(AO = AO, AZ = AZ), function(x) {
  g <- x$transitions$ELBO_gain_pct
  below <- vapply(seq_along(g), function(i) all(g[i:length(g)] < 20),
                  logical(1))
  if (any(below)) x$transitions$K_from[which(below)[1]] else NA_integer_
}, integer(1))
```

Both deliver three. That agreement is a fact about this sample and window; the
AZ BIC minimum at four is in the same evidence and disagrees, and no part of
`pefa()` arbitrates between them.

## Step 2: a comparison derived from Step 1

Step 1 delivers a count and a loading matrix. Step 2 has to be built out of
*that* result, not out of a design written down independently. Take the AO
candidate at the delivered count, keep the backbone anchors as anchors, and
assign every remaining item to the column carrying its largest absolute
loading as a regularized cell.

```{r step2-design}
Kp <- 3
Lam <- AO$loadings[[as.character(Kp)]]
anchor <- which(AO$Q0 == 1L, arr.ind = TRUE)

Q2 <- matrix(0L, J, Kp)
Q2[anchor] <- 1L
free <- setdiff(seq_len(J), anchor[, "row"])
Q2[cbind(free, max.col(abs(Lam[free, , drop = FALSE]),
                       ties.method = "first"))] <- -1L

table(cluster, support = max.col(Q2 != 0L, ties.method = "first"))
```

The derived support reproduces the three generating clusters exactly. Now fit
the same design twice --- once oblique, once bifactor --- and compare.

```{r step2-fit}
f_obl <- vbfa(Y, Q2, v0 = .001)
f_bif <- vbfa(Y, Q2, bifactor = TRUE, v0 = .001)

idx <- c("RMSEA", "SRMR", "CFI", "TLI", "AIC", "BIC")
round(rbind(
  oblique  = fit_stats(f_obl, Y = Y, Q = f_obl$Q, tau = .50)[idx],
  bifactor = fit_stats(f_bif, Y = Y, Q = f_bif$Q, tau = .50)[idx]
), 3)
```

The bifactor arm is ahead by 57.3 BIC points, which is the right answer for
this population. It is an operational comparison of two fitted models on one
design, not a test against an unrestricted alternative, and it is conditional
on the count and support that Step 1 delivered. A thin margin, or a margin that
flips when the Step-1 count changes, should be reported rather than resolved.

`special_effects()` then describes the retained bifactor fit: within-cluster
Schmid-Leiman ratios and their CVs, the implied second-order loadings, and
group effect sizes.

```{r step2-effects}
effects <- special_effects(f_bif)
effects$mean_cv
effects$groups[, c("group", "mean_ratio", "cv", "gamma", "D")]
```

## The same window as a bifactor sweep

`pefa(..., bifactor = TRUE)` sweeps the number of group factors. The backbone
supplies group columns only, the general column is added internally, and the
comparator strips it before matching, so the correspondence and the seven pair
facts have exactly the definitions used above.

```{r bifactor-sweep}
BI <- pefa(Q0_AO, Y, Kmin = 2, Kmax = 5, bifactor = TRUE, v0 = .001,
           verbose = FALSE)
BI$sweep[, c("K", "ELBO", "AIC", "BIC", "SRMR", "CFI")]
BI$transitions[, c("K_from", "K_to", "ELBO_gain_pct", "BIC_gain_pct",
                   "phi_min", "rmsd_max", "unmatched_ssl", "collision")]
round(BI$persistence$phi, 3)
round(BI$persistence$rmsd, 3)
ssl(BI)
```

In this example the bifactor sweep is much less sharp than the ordinary one.

* Both percentage columns are `NA` for every row. ELBO is highest and BIC
  lowest at the window's lower edge, so neither path contains a positive gain
  and the whole percentage path is `NA` by construction. The Step-1 reading has
  no defined path to consume; a rule that reads those columns must decide for
  itself what an all-`NA` path means, which is exactly the kind of decision the
  package leaves outside.
* `phi_min` is 0.998 or higher in every cell of `$persistence$phi`, and no
  `rmsd_max` cell exceeds 0.061. Everything persists, so persistence
  discriminates nothing either.
* `ssl()` says why. The general column holds 4.29-4.37 and the two backbone
  group columns hold about 2.1-2.25 at every candidate, while the third
  group column reaches only 0.27 at `K = 3` and splits into columns near 0.06
  by `K = 5`. With the general factor already carrying the shared variance, the
  third cluster's specific variance never accumulates into a group column here,
  and the surplus columns that follow are small enough that two of them collide
  at `4 -> 5`.

Read that as a property of this example --- this population, this two-column
backbone, this window, this sample size --- and not as a general ranking. A
bifactor sweep measures the same quantities as an ordinary one, and whether it
separates candidates as well depends on how much group-specific variance is
left once a general factor is in the model. A study that needs both readings
should run both on the same data, backbone, and window rather than assume they
must agree.

## References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical factor
  solutions. *Psychometrika*, 22(1), 53-61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113-128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4-5), 208-229.
  \doi{10.1177/01466216241261704}
* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
