---
title: "Bifactor, higher-order, and testlet structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bifactor, higher-order, and testlet structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

This vignette introduces the structures, fits an anchored bifactor model,
summarizes higher-order and testlet effects, and ends with a two-step
application. Factor-number selection and the bifactor comparison answer
different questions and are kept separate.

## 1. Three related structures

Suppose `J` items fall into `K` clusters.

**Correlated factors.** Each item loads on its cluster factor, and the cluster
factors correlate through $\Phi$.

**Orthogonal bifactor.** Every item loads on a general factor and on one group
factor; the general and group factors are mutually orthogonal.

**Higher-order (HO).** Items load on first-order factors, which in turn load on
a second-order factor. The HO model has an equivalent Schmid--Leiman (SL)
bifactor representation. For item $j$ in cluster $k$,

$$b^{gen}_j = \lambda_j\gamma_k, \qquad
  b^{grp}_j = \lambda_j\sqrt{1-\gamma_k^2}.$$

Thus $|b^{grp}_j|/|b^{gen}_j|$ is constant within each cluster when the SL
constraint holds. A bifactor model with visibly non-constant within-cluster
ratios is richer than that HO representation. Testlet or method factors use
the same general-plus-special-factor loading geometry; the group-factor effect
size below can therefore be read as a testlet effect when that interpretation
fits the design.

Cross-cluster covariance alone cannot distinguish a general factor from
correlated group factors: both produce rank-one cross-cluster blocks. The
distinguishing information is within clusters, so fitted comparisons remain
sensitive to anchors, residual structure, and how strongly the general and
group loading patterns differ.

## 2. Fit an anchored bifactor directly

`vbfa(Y, Q, bifactor = TRUE)` treats the columns of `Q` as group factors and
adds a fully specified general column internally. Consequently `K` denotes
the number of group factors, while the fitted loading matrix has `K + 1`
columns. The factor correlation matrix is fixed at the identity.

The following population is a genuine bifactor: general and group loadings
vary independently within each cluster rather than satisfying SL
proportionality.

```{r bifactor-simulate}
K <- 3
ipf <- 6
J <- K * ipf
cl <- rep(seq_len(K), each = ipf)

set.seed(7)
mla <- matrix(0, J, K + 1)
mla[, 1] <- round(runif(J, .35, .75), 2)
for (k in seq_len(K)) {
  mla[cl == k, k + 1] <- round(runif(ipf, .35, .70), 2)
}
while (max(rowSums(mla^2)) > .9) mla <- round(mla * .97, 3)

sim_bi <- sim_fa(N = 1000, mla = mla, phi = 0, rseed = 1)
Y <- sim_bi$dat
```

Anchor two items per group. For those rows, the own-group entry is one and
the other group entries are fixed at zero; the remaining entries stay
partially exploratory.

```{r bifactor-fit}
Q_direct <- matrix(-1L, J, K)
for (k in seq_len(K)) {
  anchor <- which(cl == k)[1:2]
  Q_direct[anchor, ] <- 0L
  Q_direct[anchor, k] <- 1L
}

fit_bi <- vbfa(Y, Q_direct, bifactor = TRUE)
fit_bi$Phi
round(fit_stats(fit_bi), 3)
```

`fit_bi$Q` contains the augmented design with the general column first. The
short form above is equivalent to constructing that all-one column manually
and fitting an orthogonal model, but it keeps the group-factor count explicit.
A quick recovery check can focus on the general/group split:

```{r bifactor-check}
c(
  congruence_general = abs(sum(fit_bi$Lam[, 1] * sim_bi$MLA[, 1])) /
    sqrt(sum(fit_bi$Lam[, 1]^2) * sum(sim_bi$MLA[, 1]^2)),
  MAE_general = mean(abs(abs(fit_bi$Lam[, 1]) - abs(sim_bi$MLA[, 1])))
)
```

## 3. Higher-order and testlet summaries

`sim_fa(..., gamma = )` generates the SL representation of an HO model. Fit
it as an anchored bifactor, then use `special_effects()` to summarize the
within-group ratios, the implied HO parameters, and group/testlet effect
sizes.

```{r higher-order}
gamma_true <- c(.80, .70, .60)
sim_ho <- sim_fa(
  N = 3000, K = K, ipf = ipf, lam = .8,
  gamma = gamma_true, rseed = 11
)
fit_ho <- vbfa(sim_ho$dat, Q_direct, bifactor = TRUE)

effects <- special_effects(fit_ho)
effects$mean_cv
effects$groups[, c("group", "mean_ratio", "cv", "gamma", "D")]
```

For group $k$, `cv` describes how constant the fitted
$|b^{grp}_j|/|b^{gen}_j|$ ratios are, `gamma` is the corresponding approximate
second-order loading, and `D` is the average squared loading on the selected
group or testlet factor (Zhang & Chen, 2024, Eq. 16). `$items` contains the
item-level ratios and implied first-order loadings.

These summaries are descriptive. Small CVs support an approximate SL/HO
re-expression, but `special_effects()` does not perform a hypothesis test or
choose a structure. When ratios are not close to constant, `gamma` and the
implied first-order loadings are intentionally only summaries of the richer
bifactor fit; `D` remains a direct special-factor effect-size summary.

## 4. A two-step application

When the first-order count and the presence of a separable general factor are
both uncertain, use two steps. The first is the PEFA workflow of Chen and Jin
(2026a); the second is the research comparison developed in Chen and Jin
(2026b), not part of the PEFA gain rule itself.

### Step 1: select the first-order count with ordinary PEFA

Run oblique, diagonal-residual sweeps without a general factor. This selects
the number of cluster-level factors; it does not decide whether their
correlations should later be represented by a general factor. We compare two
backbones that use the same anchors. LPC (least partially confirmatory) gives
each anchor its own-factor `1` and leaves every other entry unspecified. BPC
(bifactor partially confirmatory) additionally fixes the other backbone
entries in each anchor row to zero. The named cuts below are the
companion manuscript's convention: a 20% gain cut is primary and the laxer
10% cut is read as a sensitivity. In this tutorial LPC is the primary Step 1
branch and BPC is its specification check.

```{r step1}
Q0_lpc <- matrix(-1L, J, 2)
for (k in 1:2) {
  anchor <- which(cl == k)[1:2]
  Q0_lpc[anchor, k] <- 1L
}

Q0_bpc <- Q0_lpc
for (k in 1:2) {
  anchor <- which(cl == k)[1:2]
  Q0_bpc[anchor, ] <- 0L
  Q0_bpc[anchor, k] <- 1L
}

sweep_lpc <- pefa(
  Q0_lpc, Y, Kmin = 2, Kmax = 5,
  cuts = c(primary = 20, sensitivity = 10), verbose = FALSE
)
sweep_bpc <- pefa(
  Q0_bpc, Y, Kmin = 2, Kmax = 5,
  cuts = c(primary = 20, sensitivity = 10), verbose = FALSE
)

rbind(
  data.frame(
    backbone = "LPC", role = names(sweep_lpc$settings$cuts),
    cut = unname(sweep_lpc$settings$cuts),
    selected_K = unname(sweep_lpc$selected_K),
    boundary = unname(sweep_lpc$boundary)
  ),
  data.frame(
    backbone = "BPC", role = names(sweep_bpc$settings$cuts),
    cut = unname(sweep_bpc$settings$cuts),
    selected_K = unname(sweep_bpc$selected_K),
    boundary = unname(sweep_bpc$boundary)
  )
)
```

Check that each selected count is not merely a requested-window boundary, and
read the fit trajectories descriptively. Agreement between LPC and BPC is a
specification check, not proof of the count; disagreement must be carried
forward rather than silently resolved in favor of one backbone. The PEFA
vignette shows named-cut and adjacent-count sensitivity summaries. A boundary
result requires a complete wider sweep before proceeding.

### Compare ordinary and bifactor sweeps on the same backbone

The workflow uses the ordinary oblique sweep above to count group factors. To
see why sweep mode still deserves scrutiny, fit the exact bifactor extension
to the same data, LPC backbone, candidate window, cuts, and optimization
controls. In bifactor mode `K` still counts group factors; the fitted loading
matrix has an additional general column.

```{r compare-sweep-modes}
sweep_bifactor <- pefa(
  Q0_lpc, Y, Kmin = 2, Kmax = 5,
  cuts = c(primary = 20, sensitivity = 10),
  bifactor = TRUE, verbose = FALSE
)

rbind(
  data.frame(
    mode = "ordinary oblique", role = names(sweep_lpc$settings$cuts),
    cut = unname(sweep_lpc$settings$cuts),
    selected_K = unname(sweep_lpc$selected_K),
    boundary = unname(sweep_lpc$boundary)
  ),
  data.frame(
    mode = "bifactor", role = names(sweep_bifactor$settings$cuts),
    cut = unname(sweep_bifactor$settings$cuts),
    selected_K = unname(sweep_bifactor$selected_K),
    boundary = unname(sweep_bifactor$boundary)
  )
)

edge_columns <- c(
  "K_from", "K_to", "ELBO_gain", "BIC_gain", "rmsd", "rmsd_max",
  "phi_min", "ari", "pip_rmsd", "collision", "unmatched_max"
)
rbind(
  data.frame(
    mode = "ordinary oblique",
    sweep_lpc$transitions[, edge_columns]
  ),
  data.frame(
    mode = "bifactor",
    sweep_bifactor$transitions[, edge_columns]
  )
)
```

This population contains three group factors. In this run the ordinary sweep
selects `K = 3` at both cuts, whereas the bifactor sweep selects its lower
boundary, `K = 2`, and gains no ELBO by adding the third group column. Thus the
bifactor sweep is the less successful counter in this particular example.

The transition definitions are identical across modes. For the bifactor
sweep, matching removes the labelled general column and compares only the
group loading and PIP blocks, so the stability columns address the same group
structure as the ordinary sweep.

The loading matrices permit a second, wholly descriptive view. The SS loading
(SSL) for a column is its squared length, `colSums(Lam^2)`, equivalently the
diagonal of `crossprod(Lam)`. The bifactor values below include the general
column first, unlike its transition matching, which compares only group
columns.

```{r loading-ss}
loading_ss <- data.frame()
for (mode in c("ordinary oblique", "bifactor")) {
  x <- if (mode == "ordinary oblique") sweep_lpc else sweep_bifactor
  for (K_group in x$sweep$K) {
    Lam <- x$loadings[[as.character(K_group)]]
    SSL <- colSums(Lam^2)
    loading_ss <- rbind(
      loading_ss,
      data.frame(
        mode = mode,
        K_group = K_group,
        SSL = paste(sprintf("%.2f", SSL), collapse = ", ")
      )
    )
  }
}
loading_ss
```

For example, the final two group-column SSL values at bifactor `K = 4` are
about `0.00` and `0.28`. Their column lengths are about `0.00` and
`sqrt(0.28) = 0.53`; that ordered `.53` is a norm, not a Gram eigenvalue.
SS loadings can reveal a weak fitted column, but they are not eigenvalues of
the loading Gram matrix unless the columns are orthogonal in item space. They
therefore do not establish effective rank, choose a factor count, or add a
`pefa()` selection rule. The same per-candidate values are printed by
`summary()` for any sweep, derived from `$loadings` on the fly. This one
simulated comparison illustrates how to assess the relative behavior of the
two sweep modes; it does not establish that one mode is generally more
reliable. That claim requires the planned replicated R6 analysis.

### Step 2: compare matched anchored models by BIC

At each backbone's primary selected count, reconstruct the exact candidate
design used in its Step 1 sweep: retain that sweep's `Q0` and append an
unspecified column for every factor added above the backbone. Use this derived
design unchanged in two direct fits, an oblique model and a bifactor model with
an added general factor. Thus Step 2 cannot silently substitute the known
population design from Section 2.

```{r step2}
K_lpc <- unname(sweep_lpc$selected_K[["primary"]])
K_bpc <- unname(sweep_bpc$selected_K[["primary"]])
stopifnot(
  sweep_lpc$boundary[["primary"]] == "interior",
  sweep_bpc$boundary[["primary"]] == "interior"
)

Q_step2_lpc <- cbind(
  sweep_lpc$Q0,
  matrix(-1L, J, K_lpc - ncol(sweep_lpc$Q0))
)
Q_step2_bpc <- cbind(
  sweep_bpc$Q0,
  matrix(-1L, J, K_bpc - ncol(sweep_bpc$Q0))
)

fit_lpc_oblique <- vbfa(Y, Q_step2_lpc)
fit_lpc_bifactor <- vbfa(Y, Q_step2_lpc, bifactor = TRUE)
fit_bpc_oblique <- vbfa(Y, Q_step2_bpc)
fit_bpc_bifactor <- vbfa(Y, Q_step2_bpc, bifactor = TRUE)

bic_lpc <- c(
  oblique = unname(fit_stats(fit_lpc_oblique)["BIC"]),
  bifactor = unname(fit_stats(fit_lpc_bifactor)["BIC"])
)
bic_bpc <- c(
  oblique = unname(fit_stats(fit_bpc_oblique)["BIC"]),
  bifactor = unname(fit_stats(fit_bpc_bifactor)["BIC"])
)
rbind(
  data.frame(
    backbone = "LPC", selected_K = K_lpc, model = names(bic_lpc),
    BIC = unname(bic_lpc),
    difference_from_oblique = unname(bic_lpc - bic_lpc[["oblique"]])
  ),
  data.frame(
    backbone = "BPC", selected_K = K_bpc, model = names(bic_bpc),
    BIC = unname(bic_bpc),
    difference_from_oblique = unname(bic_bpc - bic_bpc[["oblique"]])
  )
)
```

Lower BIC is preferred. Here both backbones select three factors, but their
backbone-preserving Step 2 margins point in opposite directions (about `-22`
under LPC and `+21` under BPC, where negative favors the bifactor). The example
therefore delivers a depth-sensitivity result, not permission to choose the
more favorable arm.

An alternative is to derive a fully hard design from one selected Step 1 PIP
matrix. Here is the LPC version as a separate sensitivity:

```{r pip-hard-design}
PIP_lpc <- sweep_lpc$pips[[as.character(K_lpc)]]
Q_pip_lpc <- ifelse(PIP_lpc > .5, 1L, 0L)

backbone_columns <- seq_len(ncol(sweep_lpc$Q0))
anchors_ok <- all(
  Q_pip_lpc[, backbone_columns, drop = FALSE][sweep_lpc$Q0 == 1L] == 1L
)
stopifnot(anchors_ok, all(colSums(Q_pip_lpc == 1L) >= 2L))
colSums(Q_pip_lpc == 1L)
```

The strict `PIP > .5` operator differs at exact ties from the package's
`PIP >= .5` convention for descriptive active-loading counts and nominal hard
selection. This hard route is data-adaptive and remains an explicit
sensitivity rather than part of the backbone-preserving comparison;
transition `pip_rmsd` is continuous and uses no PIP threshold. BIC is the
operational Step 2 comparison because the two competitors contain different
loading designs and therefore different prior bookkeeping; ELBO remains
appropriate for comparing same-shaped candidates within a PEFA sweep.

This comparison is evidence, not an automatic structural verdict. Use the
same data, derived `Q`, residual assumptions, and estimation controls in the
two arms within each backbone. If LPC and BPC select different counts, their
Step 2 margins answer questions at different dimensions and must not be
attributed to backbone depth alone. Thin margins, sensitivity to defensible
anchor choices, or conspicuous residual dependence should be reported and
investigated before interpreting a bifactor preference. If the bifactor is
retained, `special_effects()` then describes whether it is close to an HO
representation and quantifies its group/testlet effects.

## References

* Schmid, J., & Leiman, J. M. (1957). The development of hierarchical factor
  solutions. *Psychometrika*, 22(1), 53--61.
* Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship
  between the higher-order factor model and the hierarchical factor model.
  *Psychometrika*, 64(2), 113--128.
* Zhang, Y., & Chen, J. (2024). Accommodating and extending various models for
  special effects within the generalized partially confirmatory factor
  analysis framework. *Applied Psychological Measurement*, 48(4--5),
  208--229. \doi{10.1177/01466216241261704}
* Chen, J., & Jin, Y. (2026a). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
* Chen, J., & Jin, Y. (2026b). *When factor count is uncertain: Stable
  structure before the bifactor decision.* Manuscript in preparation.
