---
title: "PEFA: Factor-Count Evidence from a Fixed Window"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PEFA: Factor-Count Evidence from a Fixed Window}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` fits one partially exploratory factor model for every integer `K` in a
fixed consecutive window and reports what it measured: a candidate table, the
adjacent criterion gains, and the direct structural comparison of every ordered
pair of candidates. It **chooses nothing**. There is no factor count, cut,
threshold, persistence horizon, collision policy, or stopping rule inside the
package, and no argument that would add one.

This vignette walks one sweep end to end. The example is deliberately an
*overextraction* example: the population has three factors, the window runs to
seven, and the point is to see what the three tables look like when the extra
columns are surplus. Bifactor sweeps and the step that follows a count are in
`vignette("bifactor")`.

## One sweep

Three factors correlated at 0.5, six items each with main loadings of 0.7, and
two cross-loadings of 0.3 per factor with alternating signs --- the `sim_fa()`
defaults. The backbone is **anchor-only (AO)**: two anchors per backbone factor
get their intended cell fixed to `1`, and every other cell stays `-1`,
meaning regularized and otherwise unspecified.

```{r sweep}
sim <- sim_fa(N = 500, K = 3, ipf = 6, lam = .7, lac = .3, rseed = 11)
Y <- sim$dat
groups <- rep(1:3, each = 6)

Q0 <- matrix(-1L, ncol(Y), 2)
for (k in 1:2) Q0[which(groups == k)[1:2], k] <- 1L
Q0[1:8, ]

p <- pefa(Q0, Y, Kmin = 2, Kmax = 7, verbose = FALSE)
p
```

The backbone has `K0 = 2` columns, so the window may start no lower than 2.
For each candidate `K`, `pefa()` appends `K - K0` fully regularized columns to
`Q0` and fits that design with `vbfa()`. The first two group columns keep the
backbone's identity by position; the rest are exploratory.

The call omits `v0` and `max_it`, so every candidate is fitted at the
documented `pefa()` defaults: the four-step spike ladder
`c(.01, .005, .002, .001)` and `max_it = 10000`. Those are `pefa()`'s own
defaults, and `vbfa()` has its own.

The object has exactly seven components and class `"pefa"`.

```{r object}
names(p)
class(p)
names(p$settings)
```

`$settings` holds the resolved fitting and fit-statistic controls only --- no
threshold, rule, or provenance record. Candidate matrices are stored in two
K-named lists.

```{r matrices}
names(p$loadings)
names(p$pips)
round(p$loadings[["3"]], 2)
round(p$pips[["3"]][1:6, ], 2)
```

A PIP is the posterior probability that a regularized cell is selected. Fixed
backbone cells are not regularized, so their PIPs are structural rather than
informative.

## The candidate table

```{r sweep-table}
p$sweep
```

One row per `K`, always the same eleven columns. `AIC` and `BIC` are the
hard-selection values at the resolved `tau` and `t` is their effective
parameter count. `RMSEA`, `SRMR`, `CFI`, and `TLI` are descriptive and may be
`NA` when their own definition does not apply --- `RMSEA` and `TLI` need
positive degrees of freedom. A candidate that ran to `max_it` is retained with
`converged = FALSE` and one aggregate warning; every candidate converged here.

Read down the columns. ELBO peaks at `K = 3` (-10936.65) and falls afterwards.
BIC is minimized there too (21709.40) and rises monotonically, and so is AIC
(21519.83) --- at the default spike ladder the three criteria agree on this
data set. The descriptive indices point the same way rather than adding
discrimination: RMSEA is 0.017 at `K = 3` and larger at every later candidate,
SRMR is 0.035 and never lower afterwards, and CFI and TLI peak there at 0.996
and 0.995.

## Adjacent transitions

```{r transitions}
p$transitions
```

One row per adjacent pair. `ELBO_gain_pct` and `BIC_gain_pct` are oriented so
that positive favors `K_to`: the raw gains are `ELBO_to - ELBO_from` and
`BIC_from - BIC_to`, and each is divided by the largest positive gain on its
own path and multiplied by 100. Here the largest positive gain on both paths is
the `2 -> 3` step, so that row is 100 and every later row is negative. If a
path has no positive gain at all, its whole percentage column is `NA`; the raw
criteria stay in `$sweep` and are never duplicated here.

The remaining seven columns are the structural facts for that pair. Backbone
columns are paired by position; each exploratory source column then
independently selects the exploratory target column that minimizes the
sign-aligned squared distance `||a - s b||^2`, where `s = 1` when `a'b >= 0`
and `-1` otherwise. That is Equation 17 of Chen (2023) made invariant to
whole-column reflection. Absolute Tucker congruence breaks exact distance ties
and is otherwise only a reported shape diagnostic; the smaller target index
settles anything still tied.

The assignment is independent, not one-to-one, so two sources may pick the same
target. That reuse is a **collision**, recorded rather than repaired: the
`5 -> 6` and `6 -> 7` rows both carry it here.

## Persistence, read directly

`$persistence` is a list of three upper-triangular matrices over the fitted `K`
values. There is no accessor function: `phi` (holding `phi_min`), `rmsd`
(holding **`rmsd_max`**, the weakest-column value, not the pooled `rmsd` of
`$transitions`), and the logical collision mask are read with `$`.

```{r persistence}
names(p$persistence)
round(p$persistence$phi, 3)
round(p$persistence$rmsd, 3)
p$persistence$collision
```

Every cell above the diagonal is a **direct** comparison of its two endpoint
loading matrices. Nothing is chained through the candidates in between, so the
`3 -> 7` cell is a genuine comparison of the three-factor and seven-factor
solutions rather than a product of four adjacent steps. The diagonal and lower
triangle are typed `NA`.

Read the `K = 3` row across: 0.996 against `K = 4`, 0.990 against `K = 5`,
0.989 against `K = 6`, and then 0.789 against `K = 7`. Every column of the
three-factor solution keeps a near-copy in the four-, five-, and six-factor
solutions, where the weakest-column RMSD stays at 0.039, 0.059, and 0.062. The
seven-factor cell is the exception, at RMSD 0.253.

The `K = 2` row never exceeds 0.052 anywhere, and the two-factor loading matrix
says why.

```{r k2}
round(p$loadings[["2"]][c(1, 7, 8, 13), ], 2)
```

Backbone columns are compared by *position*, and at `K = 2` the second backbone
column is not holding what it holds from `K = 3` on. Items 13-18 have taken it
over, while its own anchors --- items 7 and 8 --- sit at 0.02 and 0.08 and have
been absorbed into the first column alongside cluster 1. Its congruence with the
second column of the three-factor solution is 0.041, and that one weak pair is
the whole of `phi_min` for every `2 -> K` cell. Specifying an anchor cell leaves
it free to be estimated; it does not force the column to be about that cluster
when there are too few columns to go round.

The collision mask is `TRUE` in three cells --- `5 -> 6`, `5 -> 7`, and
`6 -> 7` --- where two or more source columns selected the same target. Those
cells sit at `phi_min` 0.998, 0.284, and 0.294, which is the first sign that a
collision is not by itself bad news about the structure.

## Column strength

`ssl()` is the descriptive companion to those numbers: `colSums(Lam^2)` for
every stored candidate matrix. It screens nothing and enters no correspondence
rule.

```{r ssl}
ssl(p)
```

This is what `phi_min` needs to be read against. `phi_min` is a **weakest-link**
statistic over *every* source column, including one the fit has driven to near
zero. Such a column has no shape, so it matches whichever target has the
smallest norm and contributes an essentially arbitrary congruence; two such
columns typically select the same target and raise `collision`. A low
`phi_min` next to a near-empty column in `ssl()` means "this candidate contains
a column with nothing in it", not "the retained structure moved". The `4 -> 5`
row is exactly that case: its `phi_min` is 0.656, yet its pooled `rmsd` is
0.030 and its `ari` is 1 --- not one item changed its dominant column. What
produced 0.656 is a source column of squared length 0.003 matching a target
column of squared length 0.007, two columns of nothing sitting next to three
columns near 3. The `5 -> 6` and `6 -> 7` collisions come from those same
near-empty columns, which all select the same smallest-norm target.

`unmatched_ssl` is on the same scale. It is the largest sum of squared loadings
among the exploratory target columns that no exploratory source selected ---
the size of the biggest column the smaller candidate did *not* account for.
Across the adjacent rows it reads 3.049, 0.003, 0.007, 0.005, 0.642. At
`2 -> 3` the surplus column carries 3.05, indistinguishable from the three
genuine columns of the three-factor solution (3.08, 3.14, 3.05): the two-factor
candidate really was missing a factor. At `3 -> 4`, `4 -> 5`, and `5 -> 6` the
surplus is under 0.01, because each of those candidates adds a column with
nothing in it and leaves the three real columns where they were.

`6 -> 7` breaks that pattern, and `ssl(p)` shows why. The seven-factor candidate
does not append an empty column: it shatters the third cluster into pieces of
1.853, 0.642, and 0.570.

```{r k7}
round(p$loadings[["7"]][13:18, 3:5], 2)
```

Items 17 and 18 have each been given a column of their own. That is a
genuinely different structure, and it is what drives the `3 -> 7` congruence
down to 0.789 after three cells at 0.989 or better. The `6 -> 7` value of 0.294
is a separate matter: like `4 -> 5`, it is two empty columns being compared.

## What the three readings say here

The criterion path, the persistence row, and `unmatched_ssl` against `ssl()`
point the same way in this example.

* ELBO, AIC, and BIC all turn at `K = 3`, and every later adjacent gain on both
  percentage paths is negative.
* The three-factor solution's columns persist into `K = 4`, `5`, and `6` at
  congruence 0.989 or better, so nothing in that range displaces the structure
  it found --- the extra columns are added beside it.
* The first added column (at `K = 3`) carries a full factor's worth of loading,
  3.05; the columns added at `K = 4`, `5`, and `6` carry less than 0.01 each.

The population had three factors, so the reading is right. **It is still the
reader's inference, not a package verdict.** `pefa()` reported six candidate
rows of eleven numbers, five transition rows, and three triangles; the sentence
"the structure is three-dimensional and the later columns are surplus" was
composed here, in the vignette, out of those numbers. Two of those numbers point
the other way if read alone: `phi_min` is 0.656 at `4 -> 5` and 0.294 at
`6 -> 7`, which looks like structural instability until `ssl()` shows that both
values belong to columns with nothing in them. Nothing in the object arbitrates.

## One analysis-side count rule

A count rule belongs to the analysis. Here is one, in full, as ordinary local
code:

```{r reading}
## Analysis-side reading, not a package rule. Variant: full look-ahead with no
## fallback -- the smallest K_from whose own and every later adjacent ELBO gain
## is under 20% of the largest positive gain on that path, and NA_integer_ when
## no such K exists.
g <- p$transitions$ELBO_gain_pct
below <- vapply(seq_along(g), function(i) all(g[i:length(g)] < 20), logical(1))
if (any(below)) p$transitions$K_from[which(below)[1]] else NA_integer_
```

Stating the variant matters, because the variants disagree. A *capped*
look-ahead rule inspects only a fixed number of following gains and can return
a count where the full-look-ahead rule returns `NA_integer_`; a rule with a
boundary fallback returns an endpoint where this one returns `NA_integer_`. For
`K = 4:6` with scores `c(0, 100, 105)`, a 10% cut, and two sustained gains, the
capped rule delivers `K = 5` and the full-look-ahead rule delivers
`NA_integer_`. An unlabelled reading copied out of a vignette would quietly
become a default; `vbpm` does not have one.

Substituting `BIC_gain_pct` for `ELBO_gain_pct` above returns `3` as well on
this path. Both percentage columns are relative to the largest positive gain
*in this fitted window*, so widening the window can change the denominator and
therefore the reading. Version 0.9.0 has no reassembly API: a wider window
means another `pefa()` call over the whole wider window.

## Display methods

```{r summary}
summary(p)
```

`summary()` returns exactly `window`, `sweep`, `transitions`, `persistence`,
`ssl`, `settings`, and `nonconverged_K`. Its print method suffixes a
persistence value with `*` wherever the parallel collision cell is `TRUE` and
labels the marker; that is display formatting only, and the stored numeric and
logical matrices are untouched.

The plot methods are descriptive trajectories. None of them draws a threshold
line or marks a selected count, because there is nothing to mark.

```{r plots, fig.width=7, fig.height=4.5}
plot(p, type = "objective", criterion = "ELBO")
plot(p, type = "gain", criterion = "BIC", pct = TRUE)
plot(p, type = "fit")
```

## Scope

What `pefa()` returns is measurement. It carries no validated threshold: the
20% cut above is an illustration, and no persistence or congruence value in
this package has been calibrated against an error rate. It performs no
automatic window extension, writes no checkpoint, saves nothing to disk, and
offers no resume or recovery path --- a failed or malformed candidate aborts
the whole call with an error naming its `K`, and a nonconverged but
well-formed candidate is retained and reported. It makes no cross-version
promise about a saved object; the package version and session used for an
analysis belong in that analysis's own run record.

The reading shown above is a property of this simulated example --- 500
observations, main loadings of 0.7, three clean clusters, and an anchored
backbone. Read it as a demonstration of how the three tables behave under
overextraction, not as evidence about how they behave at weaker signal.

## References

* Chen, J. (2023). Fully and partially exploratory factor analysis with
  bi-level Bayesian regularization. *Behavior Research Methods*, 55(4),
  2125-2142. \doi{10.3758/s13428-022-01884-7}
* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
