---
title: "Factor-number sweeps with pefa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor-number sweeps with pefa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` fits a bracketed window of partially exploratory factor models. Its
paper-defined use is an ordinary oblique factor sweep with diagonal residuals.
The result is a compact comparison object, not another fitted probability
model: it retains scalar diagnostics and loading/PIP matrices, but not the
complete candidate fits.

## A compact sweep

Start with a small LPC (least partially confirmatory) backbone, then leave
enough room above the plausible count for the sustained-drop rule to observe
what happens after the elbow. In an LPC backbone each anchor carries its `1`,
but its entries on the other backbone factors remain unspecified (`-1`); LPC
therefore imposes no fixed-zero cross-loading assumptions.

```{r sweep}
sim <- sim_fa(N = 500, K = 4, ipf = 5, cpf = 0, lam = .7, phi = .3,
              rseed = 123)
Y <- sim$dat

Q0 <- matrix(-1L, ncol(Y), 2)
Q0[1:2, 1] <- 1L
Q0[6:7, 2] <- 1L

p <- pefa(
  Q0, Y, Kmin = 2, Kmax = 6,
  cuts = c(primary = 10, sensitivity = 20),
  verbose = FALSE
)
```

The object has two scalar tables with different indexing units, while the
larger candidate matrices are kept separately:

```{r two-tables}
p$sweep
p$transitions
names(p$loadings)
```

- `$sweep` has one row per candidate. Its scalar columns contain ELBO, AIC,
  BIC, absolute fit indices, timing, iteration, and convergence information.
- `$transitions` has one row per adjacent step. It stores raw ELBO and
  larger-is-better BIC gains, five descriptive stability metrics, a logical
  column-match collision flag, and the size of the unmatched loading structure.
- `$loadings` and `$pips` are lists whose names are exactly
  `as.character(p$sweep$K)`. Each element is a matrix at its natural size, so
  no padding is needed as `K` changes. In the bifactor research extension each
  matrix has `K + 1` columns, with the general factor first.

Named vectors give the ELBO-gain choices and their positions in the requested
window. Their names and order follow `cuts`; `primary` is a role, not a fixed
percentage.

```{r named-selections}
cuts <- p$settings$cuts
data.frame(
  cut = unname(cuts),
  selected_K = unname(p$selected_K[names(cuts)]),
  boundary = unname(p$boundary[names(cuts)]),
  row.names = names(cuts)
)
```

The complete top-level contract is deliberately short:

```{r object-contract}
names(p)
names(p$settings)
c(Kmin = min(p$sweep$K), Kmax = max(p$sweep$K), K0 = ncol(p$Q0))
```

`Q0` records the backbone. The lean `$settings` list records only the controls
needed to interpret the selections, candidate statistics, matrix dimensions,
and transitions: `cuts`, `sustain`, `bifactor`, `general`, `tau`,
`stability_eps`, and `rank_adjust`. The requested window is already determined
by `$sweep$K`, and the backbone size by `ncol(p$Q0)`, so neither is duplicated.
This is still an object of class `pefa`, not a public generic sweep framework.

## The gain rule

For a larger-is-better criterion $c(K)$, the adjacent gain is
$g(K)=c(K)-c(K-1)$. The gain rule finds the first candidate followed by a
complete run of `sustain` gains strictly below

$$\frac{\delta}{100}\max_K g(K).$$

With the default `sustain = 2`, two following sub-threshold gains must be
available. The look-ahead is not shortened at the right edge; a window without
enough post-elbow headroom therefore ends at its upper boundary.

```{r elbow-toy}
elbo <- c(-5400, -5100, -4900, -4895, -4892)
select_K_elbow(2:6, elbo, delta = 10, sustain = 2)
```

Here the gains are 300, 200, 5, and 3: after the two large steps, the complete
run of two small gains selects `K = 4`. `pefa()` applies this rule to ELBO for
every named cut. The normalizer and threshold need no separate selection table
because they are exactly reproducible from the transition gains:

```{r thresholds}
gain_max <- max(p$transitions$ELBO_gain)
p$settings$cuts / 100 * gain_max
```

AIC and BIC remain useful descriptive trajectories, but `pefa()` does not
automatically select their optima or elbows. Raw optima, when wanted, are
ordinary operations on the retained paths:

```{r raw-optima}
c(
  ELBO = p$sweep$K[which.max(p$sweep$ELBO)],
  BIC = p$sweep$K[which.min(p$sweep$BIC)]
)
```

For a BIC-gain sensitivity analysis, make the smaller-is-better criterion
larger-is-better explicitly:

```{r bic-sensitivity}
select_K_elbow(p$sweep$K, -p$sweep$BIC, delta = 10,
               sustain = p$settings$sustain)
```

The normalization is relative to one fixed criterion path. Consequently the
gain decision is unchanged by a positive affine transformation of that same
path. This does not make results invariant to changing the data, items,
preprocessing, priors, `tau`, optimization settings, or candidate window.

## Reading candidates and refitting a selected model

Use the character form of `K` when extracting a candidate matrix; do not assume
that `K` is a list position.

```{r candidate-matrices}
K_selected <- p$selected_K[["primary"]]
loading_selected <- p$loadings[[as.character(K_selected)]]
pip_selected <- p$pips[[as.character(K_selected)]]

dim(loading_selected)
dim(pip_selected)
```

The retained matrices also let you inspect why a neighboring candidate may be
structurally weak. Here are the maximum absolute loading and the number of
loadings with `PIP >= .5` in every column of the `K = 5` and `K = 6`
candidates:

```{r inspect-extra-columns}
Lam5 <- p$loadings[["5"]]
PIP5 <- p$pips[["5"]]
Lam6 <- p$loadings[["6"]]
PIP6 <- p$pips[["6"]]

rbind(
  data.frame(
    K = 5, column = seq_len(ncol(Lam5)),
    max_abs_loading = round(apply(abs(Lam5), 2, max), 3),
    active_PIPs = colSums(PIP5 >= .5)
  ),
  data.frame(
    K = 6, column = seq_len(ncol(Lam6)),
    max_abs_loading = round(apply(abs(Lam6), 2, max), 3),
    active_PIPs = colSums(PIP6 >= .5)
  )
)
```

At `K = 5`, column 4 is a singleton: it has one active PIP while the four
substantive columns have five each. At `K = 6` that singleton persists and the
sixth column is essentially empty, with no active PIP and a very small maximum
loading. The matrices expose both features instead of hiding them inside one
aggregate transition number. This `.5` count is a candidate-level diagnostic,
not the continuous PIP stability metric defined below and not an automatic
rule.

The unmatched maximum alone does not establish that diagnosis. After locating
the unmatched fourth column by comparing the `K = 4` and `K = 5` matrices,
pull its loading and PIP entries directly:

```{r inspect-singleton}
data.frame(
  item = seq_len(nrow(p$loadings[["5"]])),
  loading = round(p$loadings[["5"]][, 4], 3),
  PIP = round(p$pips[["5"]][, 4], 3)
)
```

One loading is about `.61`, the second largest is only about `.054`, and only
the first has `PIP >= .5`. That is the singleton evidence; the transition
maximum is only the signal to perform this matrix-level check.

These matrices support structural inspection and post-hoc comparison. They
are not partial `vbfa` objects. If downstream work needs factor correlations,
residual parameters, scores, optimizer state, or another fit component,
explicitly refit the chosen model using the same data and estimation controls:

```{r explicit-refit, eval=FALSE}
K_selected <- p$selected_K[["primary"]]
Q_selected <- if (K_selected == ncol(Q0)) {
  Q0
} else {
  cbind(Q0, matrix(-1L, nrow(Q0), K_selected - ncol(Q0)))
}
fit_selected <- vbfa(Y, Q_selected)
```

`summary(p)` omits the matrix payloads but derives a per-candidate `ssl`
component -- the column sums of squared loadings, `colSums(Lam^2)` -- and
prints one row per candidate, which makes weak or emptied columns visible at
a glance. `print(p)` reports each named cut, selection, and boundary
compactly:

```{r summary-print}
summary(p)
print(p)
```

## Fit trajectories and parameter counting

RMSEA, SRMR, CFI, and TLI are descriptive trajectories. They help assess
whether the selected model is adequate and whether fit has levelled off, but
`pefa()` does not turn conventional fit-index cutoffs into eligibility gates.

```{r plots, fig.width=7, fig.height=4.5}
plot(p, type = "objective")
plot(p, type = "gain")
plot(p, type = "fit")
```

The gain plot defaults to ELBO. Use `criterion = "bic"` for a descriptive BIC
gain trajectory; it has no automatic threshold or selection marker. Gains
belong to transitions, so their points are located at the factor count reached
by each step. The fit plot deliberately shows trajectories without fixed
cutoff lines.

Candidate information criteria use the nominal hard-selected parameter count
by default. The optional Jacobian-rank sensitivity can be requested while the
candidate fits still exist:

```{r rank-adjusted-sweep, eval=FALSE}
p_rank <- pefa(Q0, Y, Kmin = 2, Kmax = 6,
               rank_adjust = TRUE, rank_max_J = 100,
               verbose = FALSE)
```

This optional calculation requires `numDeriv` and is guarded by `rank_max_J`;
the default sweep does not invoke numerical derivatives. Record the setting
when comparing BIC values across analyses.

## Boundary results require a new bracket

A boundary is a statement about the requested window, not proof that its edge
is the right dimension. For example, a window ending at the simulated count
does not contain the two following steps needed by the default rule:

```{r boundary}
p_narrow <- pefa(Q0, Y, Kmin = 2, Kmax = 4, verbose = FALSE)
data.frame(
  selected_K = p_narrow$selected_K[["primary"]],
  boundary = p_narrow$boundary[["primary"]]
)
```

An `upper` result asks for a complete rerun with a larger `Kmax`; a `lower`
result asks for a smaller `Kmin` when the backbone permits it. Because the
backbone has `ncol(Q0)` specified columns, `Kmin` cannot be smaller than that
count; a lower result there asks you to reduce or revise `Q0`. A one-candidate
window has boundary `single`.

Every selection also requires a complete, converged, finite ELBO path. An
unsuccessful interior candidate invalidates the selections, which become typed
`NA` with boundary `none`; the rule does not bridge the gap or continue on a
later fragment. Version 0.8.3 has no checkpoint, resume, or sweep-extension
API, so changing either boundary requires a complete rerun.

## Adjacent-count stability is descriptive

For consecutive converged candidates, `$transitions` asks how well the smaller
loading solution reappears in the larger one:

```{r adjacent-stability}
p$transitions[, c(
  "K_from", "K_to", "rmsd", "rmsd_max", "phi_min", "ari",
  "pip_rmsd", "collision", "unmatched_max"
)]
```

For a bifactor edge, the labelled general column is removed first so that the
same comparison operates on group columns in both modes. The comparison then
orients each loading column to have a nonnegative mean. The
`K0 = ncol(Q0)` backbone columns remain fixed by position. Beyond the backbone,
a column is retained when its squared L2 norm is at least
`stability_eps` (default `.1`). Following Chen (2023, Eq. 17), each retained
smaller-fit non-backbone column independently selects the minimum-SSE retained
column of the larger fit; exact ties choose the lowest larger-fit column.
Larger-fit columns are not removed after a match, so repeated selections are
possible. This is deterministic nearest-column matching, not a rotation or a
global assignment search.

The five stability metrics are:

- `rmsd`, over all entries in the fixed and matched common columns;
- `rmsd_max`, the largest column-wise RMSD, which exposes a changing factor
  that the aggregate RMSD can dilute;
- `phi_min`, the minimum **signed** Tucker congruence after the common
  orientation;
- `ari`, comparing dominant-absolute-loading item partitions using all
  retained columns, including unmatched retained columns in the larger fit;
- `pip_rmsd`, the continuous RMSD between PIPs after applying the same column
  match, restricted to cells that are selectable at both endpoints.

The logical `collision` flag is `TRUE` when two or more retained non-backbone
columns in the smaller fit independently select the same larger-fit column.
It exposes a many-to-one reproduction rather than silently forcing a distinct
match; `FALSE` only says that no match index was reused, not that the edge is
otherwise stable. It is `NA` when loading matching is unavailable. A malformed
PIP endpoint alone does not erase it because collision is loading-based.

Read signed `phi_min` beside both loading RMSDs. A bipolar or contrast column
with zero or nearly zero mean has a weak orientation: equivalent reflected
columns can then produce `phi_min` near `-1`.

For `pip_rmsd`, a backbone cell is eligible only when its `Q0` entry is `-1`
at both endpoints; every cell in a matched added exploratory column is
eligible. Fixed backbone entries (`0` or `1`) and unmatched columns are
excluded, so they cannot dilute the result. Zero means identical PIP values
over the eligible matched cells, without dichotomizing them at `.5`; no
eligible cells or a malformed PIP endpoint gives typed `NA`. The loading-based
metrics remain available when only the PIP comparison is malformed.

`unmatched_max` is separate from the five stability metrics. It is the
largest absolute loading outside the matched non-backbone core, including
larger-fit columns screened out by `stability_eps`; it is zero when nothing is
unmatched and `NA` when a complete match is impossible. It is a continuous
size diagnostic, not a `.3` salient-loading rule and not proof that one column
is uniquely the added factor.

`stability_eps` is a squared-column-norm screen, not a loading cutoff, a PIP
threshold, or an ELBO gain cut. Values around `.1`--`.2` can be used as a
sensitivity range, but are not validated structural thresholds. All these
fields are descriptive and none enters automatic factor-number selection.
Stability values are `NA` for failed endpoints, incomplete matches, and
malformed endpoint matrices as applicable.

These summaries are a package extension, not part of Chen and Jin's gain rule.
The paper's separate robustness analysis instead compares solutions obtained
from alternative backbones. When backbone choice is uncertain, repeat the
complete sweep from another substantively defensible backbone and compare the
selected counts and loading patterns. Candidate comparisons are meaningful
only when the data and all other estimation settings are held fixed.

`pefa()` does not sweep local-dependence models. Fit a prespecified residual
structure directly with `vbfa(..., ld = TRUE)` when that model is required.

## References

* Chen, J. (2023). Fully and partially exploratory factor analysis with
  bi-level Bayesian regularization. *Behavior Research Methods*, 55(4),
  2125--2142. \doi{10.3758/s13428-022-01884-7}
* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
