---
title: "Factor-number sweeps with pefa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor-number sweeps with pefa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` fits a bracketed window of partially exploratory factor models. Its
paper-defined use is an ordinary oblique factor sweep with diagonal residuals.
The result is a compact comparison object, not another fitted probability
model: it retains scalar diagnostics and loading/PIP matrices, but not the
complete candidate fits.

## A compact sweep

Start with a small confirmatory backbone, then leave enough room above the
plausible count for the sustained-drop rule to observe what happens after the
elbow.

```{r sweep}
sim <- sim_fa(N = 500, K = 4, ipf = 5, cpf = 0, lam = .7, phi = .3,
              rseed = 123)
Y <- sim$dat

Q0 <- matrix(-1L, ncol(Y), 2)
Q0[1:2, ] <- 0L; Q0[1:2, 1] <- 1L
Q0[6:7, ] <- 0L; Q0[6:7, 2] <- 1L

p <- pefa(
  Q0, Y, Kmin = 2, Kmax = 6,
  cuts = c(primary = 10, sensitivity = 20),
  verbose = FALSE
)
```

The object has two tables with different indexing units:

```{r two-tables}
scalar_columns <- setdiff(names(p$sweep), c("loading", "pip"))
p$sweep[scalar_columns]
p$transitions
```

- `$sweep` has one row per candidate. Its scalar columns contain ELBO, AIC,
  BIC, absolute fit indices, timing, iteration, and convergence information.
  Its `loading` and `pip` list-columns contain the corresponding matrices;
  list-columns are necessary because their column counts change with `K`. In
  the bifactor research extension each matrix has `K + 1` columns, with the
  general factor first.
- `$transitions` has one row per adjacent step. It stores raw ELBO and
  larger-is-better BIC gains, six descriptive stability metrics, and the size
  of the unmatched loading structure.

Named vectors give the ELBO-gain choices and their positions in the requested
window. Their names and order follow `cuts`; `primary` is a role, not a fixed
percentage.

```{r named-selections}
data.frame(
  cut = unname(p$cuts),
  selected_K = unname(p$selected_K[names(p$cuts)]),
  boundary = unname(p$boundary[names(p$cuts)]),
  row.names = names(p$cuts)
)
```

The remaining compact metadata records `Q0`, the named window, cuts, model and
objective type, and the estimation/comparison controls needed to interpret
these tables. This is still an object of class `pefa`, not a public generic
sweep framework.

## The gain rule

For a larger-is-better criterion $c(K)$, the adjacent gain is
$g(K)=c(K)-c(K-1)$. The gain rule finds the first candidate followed by a
complete run of `sustain` gains strictly below

$$\frac{\delta}{100}\max_K g(K).$$

With the default `sustain = 2`, two following sub-threshold gains must be
available. The look-ahead is not shortened at the right edge; a window without
enough post-elbow headroom therefore ends at its upper boundary.

```{r elbow-toy}
elbo <- c(-5400, -5100, -4900, -4895, -4892)
select_K_elbow(2:6, elbo, delta = 10, sustain = 2)
```

Here the gains are 300, 200, 5, and 3: after the two large steps, the complete
run of two small gains selects `K = 4`. `pefa()` applies this rule to ELBO for
every named cut. The normalizer and threshold need no separate selection table
because they are exactly reproducible from the transition gains:

```{r thresholds}
gain_max <- max(p$transitions$ELBO_gain)
p$cuts / 100 * gain_max
```

AIC and BIC remain useful descriptive trajectories, but `pefa()` does not
automatically select their optima or elbows. Raw optima, when wanted, are
ordinary operations on the retained paths:

```{r raw-optima}
c(
  ELBO = p$sweep$K[which.max(p$sweep$ELBO)],
  BIC = p$sweep$K[which.min(p$sweep$BIC)]
)
```

For a BIC-gain sensitivity analysis, make the smaller-is-better criterion
larger-is-better explicitly:

```{r bic-sensitivity}
select_K_elbow(p$sweep$K, -p$sweep$BIC, delta = 10,
               sustain = p$sustain)
```

The normalization is relative to one fixed criterion path. Consequently the
gain decision is unchanged by a positive affine transformation of that same
path. This does not make results invariant to changing the data, items,
preprocessing, priors, `tau`, optimization settings, or candidate window.

## Reading candidates and refitting a selected model

Match on `K` when extracting a candidate matrix; do not assume that `K` is a
list position.

```{r candidate-matrices}
K_selected <- p$selected_K[["primary"]]
i <- match(K_selected, p$sweep$K)
loading_selected <- p$sweep$loading[[i]]
pip_selected <- p$sweep$pip[[i]]

dim(loading_selected)
dim(pip_selected)
```

These matrices support structural inspection and post-hoc comparison. They
are not partial `vbfa` objects. If downstream work needs factor correlations,
residual parameters, scores, optimizer state, or another fit component,
explicitly refit the chosen model using the same data and estimation controls:

```{r explicit-refit, eval=FALSE}
K_selected <- p$selected_K[["primary"]]
Q_selected <- if (K_selected == ncol(Q0)) {
  Q0
} else {
  cbind(Q0, matrix(-1L, nrow(Q0), K_selected - ncol(Q0)))
}
fit_selected <- vbfa(Y, Q_selected)
```

There is deliberately no `$fits`, `$selection`, or `selected_fit()` interface.
`summary(p)` omits the matrix payloads, while `print(p)` reports each named
cut, selection, and boundary compactly.

## Fit trajectories and parameter counting

RMSEA, SRMR, CFI, and TLI are descriptive trajectories. They help assess
whether the selected model is adequate and whether fit has levelled off, but
`pefa()` does not turn conventional fit-index cutoffs into eligibility gates.

```{r plots, fig.width=7, fig.height=4.5}
plot(p, type = "objective")
plot(p, type = "gain")
plot(p, type = "fit")
```

The gain plot defaults to ELBO. Use `criterion = "bic"` for a descriptive BIC
gain trajectory; it has no automatic threshold or selection marker. Gains
belong to transitions, so their points are located at the factor count reached
by each step. The fit plot deliberately shows trajectories without fixed
cutoff lines.

Candidate information criteria use the nominal hard-selected parameter count
by default. The optional Jacobian-rank sensitivity can be requested while the
candidate fits still exist:

```{r rank-adjusted-sweep, eval=FALSE}
p_rank <- pefa(Q0, Y, Kmin = 2, Kmax = 6,
               rank_adjust = TRUE, rank_max_J = 100,
               verbose = FALSE)
```

This optional calculation requires `numDeriv` and is guarded by `rank_max_J`;
the default sweep does not invoke numerical derivatives. Record the setting
when comparing BIC values across analyses.

## Boundary results require a new bracket

A boundary is a statement about the requested window, not proof that its edge
is the right dimension. For example, a window ending at the simulated count
does not contain the two following steps needed by the default rule:

```{r boundary}
p_narrow <- pefa(Q0, Y, Kmin = 2, Kmax = 4, verbose = FALSE)
data.frame(
  selected_K = p_narrow$selected_K[["primary"]],
  boundary = p_narrow$boundary[["primary"]]
)
```

An `upper` result asks for a complete rerun with a larger `Kmax`; a `lower`
result asks for a smaller `Kmin` when the backbone permits it. Because the
backbone has `ncol(Q0)` specified columns, `Kmin` cannot be smaller than that
count; a lower result there asks you to reduce or revise `Q0`. A one-candidate
window has boundary `single`.

Every selection also requires a complete, converged, finite ELBO path. An
unsuccessful interior candidate invalidates the selections, which become typed
`NA` with boundary `none`; the rule does not bridge the gap or continue on a
later fragment. Version 0.8.2 has no checkpoint, resume, or sweep-extension
API, so changing either boundary requires a complete rerun.

## Adjacent-count stability is descriptive

For consecutive converged oblique fits, `$transitions` asks how well the
smaller loading solution reappears in the larger one:

```{r adjacent-stability}
p$transitions[, c(
  "K_from", "K_to", "rmsd", "rmsd_max", "phi_min", "ari",
  "pip_sum", "pip_product", "max_unmatched_loading"
)]
```

The comparison first orients each loading column to have a nonnegative mean.
The `K0 = ncol(Q0)` backbone columns remain fixed by position. Beyond the
backbone, a column is retained when its squared L2 norm is at least
`stability_eps` (default `.1`), and retained smaller-fit columns are matched in
order to the minimum-SSE unused retained columns of the larger fit. Exact ties
choose the lowest available larger-fit column. This is a deterministic
one-to-one heuristic, not a rotation or globally optimal assignment search.

The six stability metrics are:

- `rmsd`, over all entries in the fixed and matched common columns;
- `rmsd_max`, the largest column-wise RMSD, which exposes a changing factor
  that the aggregate RMSD can dilute;
- `phi_min`, the minimum **signed** Tucker congruence after the common
  orientation;
- `ari`, comparing dominant-absolute-loading item partitions using all
  retained columns, including unmatched retained columns in the larger fit;
- `pip_sum`, the proportion of matched cells whose `PIP >= .5` status reverses
  in either direction; and
- `pip_product`, the product of the two directional reversal proportions.

Read signed `phi_min` beside both RMSDs. A bipolar or contrast column with
zero or nearly zero mean has a weak orientation: equivalent reflected columns
can then produce `phi_min` near `-1`. For PIPs, a positive `pip_sum` with zero
`pip_product` means that reversals occurred in only one direction, not that the
solutions were perfectly stable. The bounds are
`0 <= pip_product <= pip_sum^2 / 4 <= .25`.

The PIP summaries use all matched cells. Fixed design entries can therefore
dilute them, sometimes differently across Q designs; do not compare their
magnitudes substantively across differently constrained backbones.
`max_unmatched_loading` is separate from the six stability metrics. It is the
largest absolute loading outside the matched non-backbone core, including
larger-fit columns screened out by `stability_eps`; it is zero when nothing is
unmatched and `NA` when a complete match is impossible. It is a continuous
size diagnostic, not a `.3` salient-loading rule and not proof that one column
is uniquely the added factor.

`stability_eps` is a squared-column-norm screen, not a loading cutoff, a PIP
threshold, or an ELBO gain cut. Values around `.1`--`.2` can be used as a
sensitivity range, but are not validated structural thresholds. All these
fields are descriptive and none enters automatic factor-number selection.
Stability values are `NA` for bifactor edges, failed endpoints, incomplete
matches, and malformed endpoint matrices as applicable.

These summaries are a package extension, not part of Chen and Jin's gain rule.
The paper's separate robustness analysis instead compares solutions obtained
from alternative backbones. When backbone choice is uncertain, repeat the
complete sweep from another substantively defensible backbone and compare the
selected counts and loading patterns. Candidate comparisons are meaningful
only when the data and all other estimation settings are held fixed.

`pefa()` does not sweep local-dependence models. Fit a prespecified residual
structure directly with `vbfa(..., ld = TRUE)` when that model is required.

## References

* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
