---
title: "Factor-number sweeps with pefa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor-number sweeps with pefa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` evaluates a factor-number *window*. The returned `pefa` object is also
a `vbpm_sweep`: it contains the candidate `vbpm_fit` objects and their
comparison table, but is not itself a fitted probability model. "Path" is
reserved for within-fit continuation sequences such as `v0` or `xi0`.

```{r sweep}
sim <- sim_fa(N = 500, K = 4, ipf = 5, cpf = 0, lam = .7, phi = .3, rseed = 123)
Y <- sim$dat
Q0 <- matrix(-1L, ncol(Y), 2)      # K0 = 2 backbone
Q0[1:2, ] <- 0L; Q0[1:2, 1] <- 1L
Q0[6:7, ] <- 0L; Q0[6:7, 2] <- 1L
p <- pefa(Q0, Y, Kmin = 2, Kmax = 6, verbose = FALSE)
summary(p)
selected_fit(p)
```

The data were simulated with a true K = 4, and the sweep recovers it: all five
candidates converge, and the gain rule stops at K = 4 in the interior of the
2:6 window. No `v0` or `max_it` was passed to `pefa()` here -- the call runs
with `vbfa()`'s default warm-started path, `v0 = c(.01, .005, .002, .001)`,
which is the estimator's recommended mode, so the sweep needed no tuning at
all.

The summary's comparison table lines up the evidence that the gain rule acts
on next to the fit indices that a reader would sanity-check it against. ELBO
climbs sharply from K = 2 to K = 3 to K = 4, then *drops* at K = 5 and K = 6 --
adding factors beyond the true number no longer improves the objective, it
makes the fit slightly worse. The `ELBO_gain%` column expresses each marginal
step as a percentage of the largest gain seen anywhere in the window (here the
K = 2 -> 3 step, fixed at 100): K = 3 -> 4 keeps 83.7% of that gain, while
K = 4 -> 5 and K = 5 -> 6 are negative. BIC gain tells the same story (87.9%,
then -2.0%, then 0.0%). RMSEA falls from .118 to .082 to .014 and then flattens
at .012; CFI rises from .629 to .823 to .995 and then flattens at .996. Every
criterion -- objective, BIC, and the fit indices -- agrees that the structure
is exhausted at K = 4 and that K = 5 and K = 6 add nothing but two more free
factors chasing noise.

The summary reports raw objective, objective-gain, raw BIC, and BIC-gain
selections (plus raw ELBO/ELBO-gain when the objective is the ELBO). The
primary rule is always `objective_gain`; a boundary result is a prompt to
extend the candidate window in that direction. The rest of this vignette
unpacks that rule, shows it on plots and on two deliberately awkward
simulations, and demonstrates checkpointing a long sweep.

## The gain rule

`pefa()` selects K with `select_K_elbow()`, exported so it can be applied to
any criterion curve, not only a `pefa` sweep. The idea: walk the window from
`Kmin` to `Kmax` and look at the *marginal* gain of each added factor (the
step-to-step change in the criterion, with the criterion oriented so that
larger is better). A factor is worth keeping while its marginal gain is at
least `delta` percent of the largest gain seen anywhere in the window;
selection stops at the first *sustained* run of `sustain` consecutive
sub-threshold steps (or, if the run never sustains, at `Kmax`). Because the
threshold is a fraction of the window's own largest gain, the rule is
scale-free: it does not care about sample size, item count, or the units of
the criterion.

A toy curve makes the mechanics concrete. Suppose ELBO keeps climbing sharply
through K = 4, then only creeps:

```{r elbow-toy}
elbo <- c(-5400, -5100, -4900, -4895, -4892)
select_K_elbow(2:6, elbo)
```

The gains are `r diff(elbo)`: two large jumps, then two small ones. The
largest gain is 300 (K = 3 -> 4), so with the default `delta = 10` the
threshold is 30; both remaining steps (5 and 3) fall under it, `sustain = 2`
is satisfied immediately, and the rule stops at K = 4.

The same function reads directly off a sweep's own comparison table --
`select_K_elbow()` is exactly what `pefa()` calls internally on `Objective`:

```{r elbow-sweep}
select_K_elbow(p$sweep$K, p$sweep$Objective)
```

## Visualizing the sweep

`plot.pefa()` has three panel types, selected with `type`.

`type = "objective"` puts the oriented objective (ELBO here; higher is
better) on the left axis and `BIC` (lower is better) on its own right axis --
the two live on different scales, so forcing them onto one shared axis would
flatten both curves. A dotted vertical line marks the selected K.

```{r plot-objective, fig.width=7, fig.height=4.5}
plot(p, type = "objective")
```

`type = "gain"` plots the marginal gains as a percentage of the largest gain
in the window, together with the `delta` threshold line and the selected-K
line. Gains are undefined at `Kmin` (there is no step into the first
candidate), so both curves start one step into the window even though the
x-axis spans the whole thing.

```{r plot-gain, fig.width=7, fig.height=4.5}
plot(p, type = "gain")
```

`type = "fit"` draws two panels side by side: absolute misfit (RMSEA and
SRMR, with the conventional .06/.08 cutoffs shown as dotted lines) and
incremental fit (CFI and TLI, with the conventional .95 cutoff). Both panels
mark the selected K.

```{r plot-fit, fig.width=9, fig.height=4.5}
plot(p, type = "fit")
```

## Boundary selections: a prompt to extend, not an answer

If the window is too narrow, the gain rule can only pick the largest K it was
offered -- that is a boundary selection, not evidence that K is correct. Reuse
the same `Y` from the lead sweep (true K = 4), but this time restrict the
window to K = 2:4:

```{r boundary-narrow}
p_narrow <- pefa(Q0, Y, Kmin = 2, Kmax = 4, verbose = FALSE)
p_narrow$selected_K
summary(p_narrow)$boundary
```

`summary()$boundary` is `"upper"`: within 2:4 both marginal gains stay above
the `delta` threshold, so the rule never finds a sustained sub-threshold run
and simply stops at the top of the window it was given -- exactly what a
window that stops too early looks like, even though K = 4 happens to be the
right answer here.

The lead sweep `p` already shows what widening the window resolves: run on
the same data with `Kmax = 6` instead of 4, the selection lands in the
interior at K = 4 (`summary(p)$boundary` is `"interior"`), because two more
candidates on either side confirm nothing better is being left out. The rule
for using `boundary`: a `"lower"` or `"upper"` result is never a settled
answer about dimensionality by itself, only an instruction to widen the
window in that direction and re-run until the selection lands in the interior
(or a principled reason exists to stop, e.g. `Kmin`/`Kmax` are already at a
substantive limit).

## Minor factors: reading dust as dust

A harder case is data that truly *does* have extra structure, but structure so
weak it barely deserves the name "factor". `sim_fa(..., minor = c("weak",
"moderate"))` plants two orthogonal minor factors that load weakly (or
moderately) on every item, following the minor-factor definitions of
Auerswald and Moshagen (2019) -- these are not noise, they are real
population parameters, just tiny ones. The auxiliary sweeps below fix
`v0 = 0.001` (skipping the warm-started default's earlier stages) so the
vignette builds quickly; the lead sweep above is the one run in `vbfa()`'s
recommended default mode.

```{r minor-sim}
simM <- sim_fa(N = 400, K = 3, ipf = 6, lam = .7, lac = .3,
               minor = c("weak", "moderate"), rseed = 11)
Ym <- simM$dat
Jm <- ncol(Ym)
Q0m <- matrix(-1L, Jm, 2)
for (k in 1:2) {
  a <- which(rep(1:3, each = 6) == k)[1:2]
  Q0m[a, ] <- 0L; Q0m[a, k] <- 1L
}
pm <- pefa(Q0m, Ym, Kmin = 2, Kmax = 6, v0 = 0.001, max_it = 500,
           verbose = FALSE)
pm$selected_K
pm$sweep[, c("K", "Objective", "Objective_gain_pct", "BIC",
             "RMSEA", "SRMR", "CFI", "TLI")]
```

The gain rule stops at K = 3 -- the three substantive factors -- even though
the population has two more, very real, orthogonal minor factors on top.
Look at `SRMR`: it keeps falling all the way to K = 6, because two more
free factors always let the model absorb a little more residual covariance,
however trivial. A criterion read on its raw scale (lowest SRMR, or highest
CFI/TLI) would keep climbing the window and never stop cleanly. The
scale-free gain threshold does not: past K = 3 the marginal improvement is a
small fraction of the K = 2 -> 3 jump, so it reads as dust rather than
structure. This is the gain rule's core selling point -- it is built to tell
substantive dimensions from minor ones, not just to find *some* local
improvement.

## Checkpointing a sweep: the provenance manifest

Each candidate in a window is a full `vbfa()` fit, so a wide sweep can be
slow enough to want resuming across sessions. `save_path` writes a CSV
checkpoint plus a sidecar `<save_path>.manifest`:

```{r checkpoint-run}
cp <- tempfile(fileext = ".csv")
p1 <- pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.001, max_it = 500,
           save_path = cp, verbose = FALSE)
file.exists(cp)
file.exists(paste0(cp, ".manifest"))
```

Calling `pefa()` again with the *same* data, design, window, and settings
finds the checkpoint and reuses every row instead of refitting:

```{r checkpoint-resume}
p2 <- pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.001, max_it = 500,
           save_path = cp, verbose = TRUE)
identical(p1$selected_K, p2$selected_K)
all.equal(summary(p1)$comparison$Objective, summary(p2)$comparison$Objective)
```

The `(from checkpoint)` lines above show every K being served from the CSV;
`vbfa()` is deterministic, so the reused rows agree with a fresh fit to
floating-point precision.

Now change one setting that affects comparability -- say `v0` -- and try to
resume from the same checkpoint:

```{r checkpoint-refuse}
bad <- try(
  pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.01, max_it = 500,
       save_path = cp, verbose = FALSE),
  silent = TRUE
)
cat(conditionMessage(attr(bad, "condition")))
```

```{r checkpoint-cleanup, include=FALSE}
unlink(c(cp, paste0(cp, ".manifest")))
```

The manifest records the data, design, window, and every tuning setting that
affects comparability (including anything passed through `...`, such as
`v0`), and names exactly what differs. Resuming by matching `K` alone would
silently pool candidates fit under different regularization -- rows that look
comparable in a table but are not -- so the checkpoint refuses instead.
Delete the checkpoint (or pick a new `save_path`) to start over deliberately.

## Objective-type discipline

`pefa()` does not assume the objective is always the ELBO. For ordinary
(diagonal) `vbfa` sweeps it is the ELBO; for LD-aware sweeps (`ld = TRUE`) it
is the terminal VECM objective instead, and the ELBO-specific
columns and selections (`elbo`, `elbo_gain`) are `NA` because the residual
precision there is point-updated rather than given a full variational
distribution. `objective_gain` remains the primary rule in both cases, just
computed on whichever objective the candidates actually share. What `pefa()`
will not do is mix the two silently: every candidate in a window is fit with
the same `ld` setting, but if a sweep is ever assembled from candidates whose
`objective_type` differs, `pefa()` refuses with an explicit error rather than
comparing ELBO to a VECM objective as though they were the same scale.

## References

* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
* Auerswald, M., & Moshagen, M. (2019). How to determine the number of
  factors to retain in exploratory factor analysis: A comparison of
  extraction methods under realistic conditions. *Psychological Methods*,
  24(4), 468-491.
