---
title: "Factor-number sweeps with pefa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor-number sweeps with pefa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` fits a bracketed window of partially exploratory factor models. Its
paper-defined use is an ordinary oblique factor sweep with diagonal residuals.
The result is a sweep, not another fitted probability model: candidate fits
live in `$fits`, and candidate-, transition-, and rule-level quantities live in
three separate tables.

## A compact sweep

Start with a small confirmatory backbone, then leave enough room above the
plausible count for the sustained-drop rule to observe what happens after the
elbow.

```{r sweep}
sim <- sim_fa(N = 500, K = 4, ipf = 5, cpf = 0, lam = .7, phi = .3,
              rseed = 123)
Y <- sim$dat

Q0 <- matrix(-1L, ncol(Y), 2)
Q0[1:2, ] <- 0L; Q0[1:2, 1] <- 1L
Q0[6:7, ] <- 0L; Q0[6:7, 2] <- 1L

p <- pefa(
  Q0, Y, Kmin = 2, Kmax = 6,
  cuts = c(primary = 10, sensitivity = 20),
  verbose = FALSE
)
```

The object has three tables with different indexing units:

```{r three-tables}
p$sweep
p$transitions
p$selection
```

- `$sweep` has one row per candidate. ELBO, AIC, BIC, and the absolute fit
  indices describe that fitted model.
- `$transitions` has one row per adjacent step. It stores raw ELBO and
  larger-is-better BIC gains, plus small adjacent-count stability summaries.
- `$selection` has one row per rule. Each gain rule records the named cut, its
  normalization denominator (`gain_max`), and the resulting absolute
  `threshold`.

The primary top-level result is the ELBO gain selection at the cut named
`primary`. Every candidate fit is retained, and the accessor retrieves that
fit without storing another copy:

```{r selected-fit}
p$selected_K
selected_fit(p)
```

## Raw and gain selections

For a larger-is-better criterion $c(K)$, the adjacent gain is
$g(K)=c(K)-c(K-1)$. The gain rule finds the first candidate followed by a
complete run of `sustain` gains strictly below

$$\frac{\delta}{100}\max_K g(K).$$

With the default `sustain = 2`, two following sub-threshold gains must be
available. The look-ahead is not shortened at the right edge; a window without
enough post-elbow headroom therefore ends at its upper boundary.

```{r elbow-toy}
elbo <- c(-5400, -5100, -4900, -4895, -4892)
select_K_elbow(2:6, elbo, delta = 10, sustain = 2)
```

Here the gains are 300, 200, 5, and 3: after the two large steps, the complete
run of two small gains selects `K = 4`. `pefa()` applies
the same scanner to ELBO and to `-BIC`; AIC remains available in `$sweep` as a
descriptive fit statistic but has no automated selection rule.

Raw optima and gain choices answer different questions and are kept together
for comparison. Multiple named cuts provide a compact sensitivity analysis:

```{r selection-rules}
subset(p$selection, criterion %in% c("elbo", "bic"))
```

The normalization is relative to one fixed criterion path. Consequently the
gain decision is unchanged by a positive affine transformation of that same
path. This does not make results invariant to changing the data, items,
preprocessing, priors, `tau`, optimization settings, or candidate window.

## Reading fit and plots

RMSEA, SRMR, CFI, and TLI are descriptive trajectories. They help assess
whether the selected model is adequate and whether fit has levelled off, but
`pefa()` does not turn conventional fit-index cutoffs into eligibility gates.

```{r plots, fig.width=7, fig.height=4.5}
plot(p, type = "objective")
plot(p, type = "gain")
plot(p, type = "fit")
```

The gain plot defaults to ELBO; use `criterion = "bic"` to display BIC gains.
Gains belong to transitions, so their points are located at the factor count
reached by each step. The fit plot deliberately shows trajectories without
fixed cutoff lines.

## Boundary results require a new bracket

A boundary is a statement about the requested window, not proof that its edge
is the right dimension. For example, a window ending at the simulated count
does not contain the two following steps needed by the default rule:

```{r boundary}
p_narrow <- pefa(Q0, Y, Kmin = 2, Kmax = 4, verbose = FALSE)
subset(
  p_narrow$selection,
  criterion == "elbo" & form == "gain" & cut_name == "primary"
)
```

An `upper` result asks for a complete rerun with a larger `Kmax`; a `lower`
result asks for a smaller `Kmin` when the backbone permits it. Because the
backbone has `ncol(Q0)` specified columns, `Kmin` cannot be smaller than that
count. Rerun the complete requested window after changing either boundary.

Every selection also requires a complete, converged, finite criterion path.
An unsuccessful interior candidate invalidates that criterion's selection;
the rule does not bridge the gap or continue on a later fragment.

## Adjacent-count stability is descriptive

For consecutive converged oblique fits, `$transitions` also asks how well the
smaller loading solution reappears in the larger one:

```{r adjacent-stability}
p$transitions[, c(
  "K_from", "K_to", "phi_min", "phi_mean", "ari", "rmsd", "max_loading"
)]
```

`phi_min` and `phi_mean` summarize absolute Tucker congruence after a simple
best-unused-column alignment; `ari` compares dominant-factor item assignments;
`rmsd` is computed after sign alignment; and `max_loading` describes the one
unmatched column in the larger solution. There is no built-in stability cutoff.
These summaries are a package extension, not part of Chen and Jin's gain rule.
The paper's separate robustness analysis instead compares solutions obtained
from alternative backbones.

When backbone choice is uncertain, repeat the complete sweep from another
substantively defensible backbone and compare the selected counts and loading
patterns. Candidate comparisons are meaningful only when the data and all
other estimation settings are held fixed.

`pefa()` does not sweep local-dependence models. Fit a prespecified residual
structure directly with `vbfa(..., ld = TRUE)` when that model is required.

## References

* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
