---
title: "PEFA Fit Assessment and Factor-Count Evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PEFA Fit Assessment and Factor-Count Evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` fits a fixed window of partially exploratory factor models and returns
an evidence object. It does not select a factor count, apply a gain cut, or
declare a boundary. This vignette follows the PEFA fit-and-count framework of
Chen and Jin (2026): ordinary oblique candidates, post-selection fit evidence,
and several transparent readings of the same criterion paths. Structural
persistence and the later bifactor decision are covered in
`vignette("bifactor")`.

## One ordinary oblique sweep

Simulate four correlated factors and launch the sweep from a two-column
anchor-only (AO) backbone. In AO, an anchor's intended loading is specified by
`1`, while its cross-backbone cells remain unspecified (`-1`).

```{r sweep}
sim <- sim_fa(N = 500, K = 4, ipf = 5, cpf = 0, lam = .7, phi = .3,
              rseed = 123)
Y <- scale(sim$dat, center = TRUE, scale = TRUE)

Q0 <- matrix(-1L, ncol(Y), 2)
Q0[1:2, 1] <- 1L
Q0[6:7, 2] <- 1L

p <- pefa(
  Q0, Y, Kmin = 2, Kmax = 7,
  v0 = .001, max_it = 10000, convChk = FALSE, tolVal = 1e-4,
  tau = .50, stability_eps = .10,
  rank_adjust = TRUE, rank_max_J = 100,
  verbose = FALSE
)
```

The analyzed matrix is standardized once explicitly so that its means, sample
SDs, and analysis-input hash can be archived; `vbfa()`'s internal
standardization then makes no substantive change. The package provenance hash
for its engine-standardized matrix has a distinct canonical preimage, however,
so an archive should retain and verify both rather than require the two byte
hashes to be equal.

The evidence object separates candidate evidence from adjacent criterion
evidence. Loading and PIP matrices remain available for audit or an explicit
refit, but no decision-dependent field is stored.

```{r evidence-object}
p$sweep
p$transitions
names(p$loadings)
names(p$pips)
names(p$settings)
names(p$provenance)
```

- `$sweep` has one row per candidate, including ELBO, nominal/hard/soft
  parameter counts, hard/soft AIC/BIC, hard absolute-fit indices, convergence,
  and component-specific status/reason fields.
- `$transitions` contains raw and percent-of-maximum gains for ELBO and
  hard/soft AIC/BIC. Every AIC/BIC gain is oriented as `IC_from - IC_to`, so
  larger is better for every displayed gain.
- `$loadings` and `$pips` are named by the actual candidate count.

The percentage gain is a display and decision aid, not an additional fit
criterion. It is relative to the largest positive gain in this fitted window;
changing the window can therefore change both its denominator and a later
count reading.

## Several count readings from the same evidence

`select_K_elbow()` is a criterion-generic calculator. Its convenient defaults
are `delta = 20` and `sustain = 1`; `pefa()` never calls it automatically. The
following decision profiles deliberately use different settings.

```{r count-profiles}
K_path <- p$sweep$K

read_score_path <- function(x, criterion) {
  K <- as.integer(x$sweep$K)
  if (!identical(K, seq.int(min(K), max(K))))
    stop("The count rule requires a complete consecutive K path.")
  score <- switch(
    criterion,
    elbo = x$sweep$ELBO,
    aic = -x$sweep$AIC,
    bic = -x$sweep$BIC,
    aic_s = -x$sweep$AIC_S,
    bic_s = -x$sweep$BIC_S
  )
  component_ok <- switch(
    criterion,
    elbo = x$sweep$elbo_status == "ok",
    aic = x$sweep$stats_status == "ok",
    bic = x$sweep$stats_status == "ok",
    aic_s = x$sweep$stats_status == "ok",
    bic_s = x$sweep$stats_status == "ok"
  )
  if (!all(x$sweep$fit_status == "ok" & component_ok & is.finite(score)))
    stop("The requested criterion path is not usable as a complete path.")
  score
}

score_elbo <- read_score_path(p, "elbo")
score_aic <- read_score_path(p, "aic")
score_bic <- read_score_path(p, "bic")
score_bic_s <- read_score_path(p, "bic_s")

count_readings <- data.frame(
  profile_id = c(
    "pefa_fit_v1_style_elbo",
    "convenience_elbo",
    "convenience_aic",
    "convenience_bic",
    "convenience_soft_bic",
    "raw_elbo_maximum",
    "raw_aic_minimum",
    "raw_bic_minimum",
    "raw_soft_bic_minimum"
  ),
  profile_source_id = "vbpm_pefa_vignette",
  profile_source_version = "0.9.0",
  profile_source_sha256 = NA_character_,
  count_rule_semantics_id = c(
    rep("vbpm_0.9_gain_full_lookahead_na", 5),
    rep("raw_smallest_optimum", 4)
  ),
  candidate_path_policy_id = "complete_consecutive_all_ok",
  criterion = c(
    "ELBO", "ELBO", "hard AIC", "hard BIC", "soft BIC",
    "ELBO", "hard AIC", "hard BIC", "soft BIC"
  ),
  count_rule = c(rep("gain", 5), rep("raw_extremum", 4)),
  delta = c(10, 20, 20, 20, 20, NA, NA, NA, NA),
  sustain = c(2, 1, 1, 1, 1, NA, NA, NA, NA),
  selected_K = c(
    select_K_elbow(K_path, score_elbo, delta = 10, sustain = 2),
    select_K_elbow(K_path, score_elbo),
    select_K_elbow(K_path, score_aic, delta = 20, sustain = 1),
    select_K_elbow(K_path, score_bic, delta = 20, sustain = 1),
    select_K_elbow(K_path, score_bic_s, delta = 20, sustain = 1),
    K_path[which.max(score_elbo)],
    K_path[which.max(score_aic)],
    K_path[which.max(score_bic)],
    K_path[which.max(score_bic_s)]
  )
)
count_readings
```

The first row uses the PEFA fit paper v1's 10%/two-gain settings under the
0.9.0 helper semantics. It is not an exact reproduction of v1 boundary cases:
0.9.0 requires complete look-ahead and returns `NA_integer_` rather than a
boundary count when no complete strict crossing exists. The second row uses the
helper defaults. The next three hold those settings fixed while changing the
hard/soft criterion, and the final four abandon the gain rule entirely. These
profiles can agree, but agreement is an empirical property of this path rather
than a package guarantee. When they differ, report the different counts and
the evidence responsible for the difference; do not silently choose the one
that gives the preferred downstream model.

The source hash is `NA` only because this installed vignette is not a sealed
research registry. A paper registry must record the exact source version and
hash together with both semantics identifiers.

The arguments can be changed without refitting because every reading consumes
the retained path:

```{r change-a-decision}
data.frame(
  delta = c(5, 10, 20, 30),
  sustain_1 = vapply(
    c(5, 10, 20, 30),
    function(d) select_K_elbow(K_path, p$sweep$ELBO,
                               delta = d, sustain = 1),
    integer(1)
  ),
  sustain_2 = vapply(
    c(5, 10, 20, 30),
    function(d) select_K_elbow(K_path, p$sweep$ELBO,
                               delta = d, sustain = 2),
    integer(1)
  )
)
```

This table is a sensitivity display, not permission to select `delta` after
seeing which count is most attractive. For a confirmatory use, state the
criterion, cut, and sustain value before interpreting the count. For an
exploratory use, show the relevant profile range and label any preferred
reading as judgemental. Every reading above first requires a complete
consecutive path, `fit_status == "ok"`, the relevant ELBO/statistics status, and
finite scores; it never deletes a failed candidate and bridges the resulting K
gap.

### If the PEFA fit paper changes

The paper and this package are versioned separately. A later paper revision may
apply another retained criterion, cut, or sustain value as a newly named profile
over the same usable path. A different window, simulation design, fitted control,
or criterion absent from the object creates new evidence and requires the
corresponding rerun rather than a profile-only change. If the revision
changes the algorithm itself---for example, how no positive gain, incomplete
look-ahead, or no crossing is handled---give it a new `count_rule` name rather
than silently redefining `gain`. Also freeze the hard/soft and rank-count
conventions: changing rank adjustment changes hard AIC/BIC and requires those
paths and their simulations to be regenerated. The v1 engine also omitted
nonconverged candidates before selection; retaining that behavior versus
requiring a complete path is another rule-semantic decision, not an incidental
implementation detail. Preserve a legacy path/output fixture before migrating.
In particular, the paper's saved K = 4:6 design cannot return K = 5 with
`sustain = 2` under complete look-ahead; that reading requires K = 7 and a new
simulation rather than post-hoc rescoring of the three-candidate path.
Freeze `tau` as well, because it changes the hard-selected pattern and therefore
the hard parameter count and information criteria.
Also record the optimizer path: this vignette passes scalar `v0 = .001`
explicitly, whereas the package convenience default is a continuation sequence.
Changing `v0`, `max_it`, `convChk`, or `tolVal` creates new fitted evidence.
The structural-persistence
workflow remains in `vignette("bifactor")` and need not change merely because a
PEFA-fit count profile changes.

## Fit trajectories and hard/soft post-selection evidence

Absolute fit trajectories describe whether candidate misfit has levelled off;
they do not create automatic eligibility thresholds.

```{r plots, fig.width=7, fig.height=4.5}
plot(p, type = "objective", criterion = "elbo")
plot(p, type = "gain", criterion = "elbo", pct = TRUE)
plot(p, type = "fit")
```

To obtain a complete fitted object, refit an explicitly identified candidate.
Here the raw hard-BIC minimum is used only to make the example deterministic.

```{r explicit-refit}
K_refit <- count_readings$selected_K[
  count_readings$profile_id == "raw_bic_minimum"
]
Q_refit <- cbind(
  Q0,
  matrix(-1L, nrow(Q0), K_refit - ncol(Q0))
)
fit_refit <- vbfa(
  Y, Q_refit, v0 = .001, max_it = 10000,
  convChk = FALSE, tolVal = 1e-4
)
round(fit_stats(
  fit_refit, Y = Y, Q = fit_refit$Q, tau = .50,
  rank_adjust = TRUE, rank_max_J = 100
), 3)
```

`fit_stats()` distinguishes the nominal, hard-selected, and soft/effective
parameter counts and their corresponding information criteria. Record the
parameter-count convention whenever AIC or BIC is compared across analyses.
If Jacobian-rank adjustment is required, request and record it in the sweep and
the refit rather than mixing conventions.

## Window position is evidence, not a package boundary verdict

A count at an edge, an `NA` returned because no strict crossing exists, or too
little look-ahead for the requested `sustain` is information about the fitted
window. It is not proof that an endpoint is the correct count.

```{r headroom}
gain_rows <- count_readings$count_rule == "gain"
data.frame(
  profile_id = count_readings$profile_id[gain_rows],
  selected_K = count_readings$selected_K[gain_rows],
  at_lower_edge = count_readings$selected_K[gain_rows] == min(K_path),
  at_upper_edge = count_readings$selected_K[gain_rows] == max(K_path),
  following_candidates = max(K_path) - count_readings$selected_K[gain_rows],
  requested_sustain = count_readings$sustain[gain_rows]
)
```

If the prespecified analysis requires more headroom, change the bracket and
obtain a new complete evidence object over the expanded window. In vbpm 0.9.0
this means another `pefa()` call. A future managed extension may reuse verified
compatible candidate fits only when exactly one bound moves strictly outward
without creating a count below `max(1L, ncol(p$Q0))`, and only when the
schema/engine/data/Q0/settings/evidence/lineage fingerprints agree. It must leave
`p` byte-identical and return a complete child with the parent evidence hash and
the exact reused/new K values. Do not paste in one isolated candidate or
reinterpret a missing crossing as a selected endpoint. The package rebuilds
the full evidence; because a percentage gain is window-relative, the assessor
then recomputes every profile reading against the child.

## References

* Chen, J. (2023). Fully and partially exploratory factor analysis with
  bi-level Bayesian regularization. *Behavior Research Methods*, 55(4),
  2125-2142. \doi{10.3758/s13428-022-01884-7}
* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
