---
title: "Factor-number sweeps with pefa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor-number sweeps with pefa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` evaluates a factor-number *window*. The returned `pefa` object is also
a `vbpm_sweep`: it contains the candidate `vbpm_fit` objects and their
comparison table, but is not itself a fitted probability model. "Path" is
reserved for within-fit continuation sequences such as `v0` or `xi0`.

```{r sweep}
s <- sim_fa(N = 250, K = 3, ipf = 5, rseed = 19)
Y <- s$dat
Q0 <- matrix(-1L, ncol(Y), 2)
Q0[1:2, ] <- 0L; Q0[1:2, 1] <- 1L
Q0[6:7, ] <- 0L; Q0[6:7, 2] <- 1L

p <- pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.001, max_it = 500,
          verbose = FALSE)
summary(p)
selected_fit(p)
```

The summary reports raw objective, objective-gain, raw BIC, and BIC-gain
selections (plus raw ELBO/ELBO-gain when the objective is the ELBO). The
primary rule is always `objective_gain`; a boundary result is a prompt to
extend the candidate window in that direction. The rest of this vignette
unpacks that rule, shows it on plots and on two deliberately awkward
simulations, and demonstrates checkpointing a long sweep.

## The gain rule

`pefa()` selects K with `select_K_elbow()`, exported so it can be applied to
any criterion curve, not only a `pefa` sweep. The idea: walk the window from
`Kmin` to `Kmax` and look at the *marginal* gain of each added factor (the
step-to-step change in the criterion, with the criterion oriented so that
larger is better). A factor is worth keeping while its marginal gain is at
least `delta` percent of the largest gain seen anywhere in the window;
selection stops at the first *sustained* run of `sustain` consecutive
sub-threshold steps (or, if the run never sustains, at `Kmax`). Because the
threshold is a fraction of the window's own largest gain, the rule is
scale-free: it does not care about sample size, item count, or the units of
the criterion.

A toy curve makes the mechanics concrete. Suppose ELBO keeps climbing sharply
through K = 4, then only creeps:

```{r elbow-toy}
elbo <- c(-5400, -5100, -4900, -4895, -4892)
select_K_elbow(2:6, elbo)
```

The gains are `r diff(elbo)`: two large jumps, then two small ones. The
largest gain is 300 (K = 3 -> 4), so with the default `delta = 10` the
threshold is 30; both remaining steps (5 and 3) fall under it, `sustain = 2`
is satisfied immediately, and the rule stops at K = 4.

The same function reads directly off a sweep's own comparison table --
`select_K_elbow()` is exactly what `pefa()` calls internally on `Objective`:

```{r elbow-sweep}
select_K_elbow(p$sweep$K, p$sweep$Objective)
```

## Visualizing the sweep

`plot.pefa()` has three panels: the oriented objective and `-BIC` curves, the
marginal-gain curves against the `delta` threshold line, and the fit-index
curves.

```{r plots, fig.width=6, fig.height=4}
plot(p, type = "objective")
plot(p, type = "gain")
plot(p, type = "fit")
```

## Boundary selections: a prompt to extend, not an answer

If the window is too narrow, the gain rule can only pick the largest K it was
offered -- that is a boundary selection, not evidence that K is correct. Take
data simulated with a true K = 4, and sweep only K = 2:4:

```{r boundary-sim}
sim4 <- sim_fa(N = 400, K = 4, ipf = 5, lam = .7, lac = .3, rseed = 7)
Y4 <- sim4$dat
J4 <- ncol(Y4)
Q0_4 <- matrix(-1L, J4, 2)
for (k in 1:2) {
  a <- which(rep(1:4, each = 5) == k)[1:2]
  Q0_4[a, ] <- 0L; Q0_4[a, k] <- 1L
}
```

```{r boundary-narrow}
p_narrow <- pefa(Q0_4, Y4, Kmin = 2, Kmax = 4, v0 = 0.001, max_it = 500,
                 verbose = FALSE)
p_narrow$selected_K
summary(p_narrow)$boundary
```

`summary()$boundary` is `"upper"`: the sweep picked the top of the window it
was given, which is exactly what a window that stops too early looks like.
Extend `Kmax` and check again:

```{r boundary-wide}
p_wide <- pefa(Q0_4, Y4, Kmin = 2, Kmax = 6, v0 = 0.001, max_it = 500,
               verbose = FALSE)
p_wide$selected_K
summary(p_wide)$boundary
```

Now the selection is `"interior"` at the true K = 4: two more candidates on
either side confirm nothing better is being left out. The rule for using
`boundary`: a `"lower"` or `"upper"` result is never a settled answer about
dimensionality by itself, only an instruction to widen the window in that
direction and re-run until the selection lands in the interior (or a
principled reason exists to stop, e.g. `Kmin`/`Kmax` are already at a
substantive limit).

## Minor factors: reading dust as dust

A harder case is data that truly *does* have extra structure, but structure so
weak it barely deserves the name "factor". `sim_fa(..., minor = c("weak",
"moderate"))` plants two orthogonal minor factors that load weakly (or
moderately) on every item, following the minor-factor definitions of
Auerswald and Moshagen (2019) -- these are not noise, they are real
population parameters, just tiny ones.

```{r minor-sim}
simM <- sim_fa(N = 400, K = 3, ipf = 6, lam = .7, lac = .3,
               minor = c("weak", "moderate"), rseed = 11)
Ym <- simM$dat
Jm <- ncol(Ym)
Q0m <- matrix(-1L, Jm, 2)
for (k in 1:2) {
  a <- which(rep(1:3, each = 6) == k)[1:2]
  Q0m[a, ] <- 0L; Q0m[a, k] <- 1L
}
pm <- pefa(Q0m, Ym, Kmin = 2, Kmax = 6, v0 = 0.001, max_it = 500,
           verbose = FALSE)
pm$selected_K
pm$sweep[, c("K", "Objective", "Objective_gain_pct", "BIC",
             "RMSEA", "SRMR", "CFI", "TLI")]
```

The gain rule stops at K = 3 -- the three substantive factors -- even though
the population has two more, very real, orthogonal minor factors on top.
Look at `SRMR`: it keeps falling all the way to K = 6, because two more
free factors always let the model absorb a little more residual covariance,
however trivial. A criterion read on its raw scale (lowest SRMR, or highest
CFI/TLI) would keep climbing the window and never stop cleanly. The
scale-free gain threshold does not: past K = 3 the marginal improvement is a
small fraction of the K = 2 -> 3 jump, so it reads as dust rather than
structure. This is the gain rule's core selling point -- it is built to tell
substantive dimensions from minor ones, not just to find *some* local
improvement.

## Checkpointing a sweep: the provenance manifest

Each candidate in a window is a full `vbfa()` fit, so a wide sweep can be
slow enough to want resuming across sessions. `save_path` writes a CSV
checkpoint plus a sidecar `<save_path>.manifest`:

```{r checkpoint-run}
cp <- tempfile(fileext = ".csv")
p1 <- pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.001, max_it = 500,
           save_path = cp, verbose = FALSE)
file.exists(cp)
file.exists(paste0(cp, ".manifest"))
```

Calling `pefa()` again with the *same* data, design, window, and settings
finds the checkpoint and reuses every row instead of refitting:

```{r checkpoint-resume}
p2 <- pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.001, max_it = 500,
           save_path = cp, verbose = TRUE)
identical(p1$selected_K, p2$selected_K)
all.equal(summary(p1)$comparison$Objective, summary(p2)$comparison$Objective)
```

The `(from checkpoint)` lines above show every K being served from the CSV;
`vbfa()` is deterministic, so the reused rows agree with a fresh fit to
floating-point precision.

Now change one setting that affects comparability -- say `v0` -- and try to
resume from the same checkpoint:

```{r checkpoint-refuse}
bad <- try(
  pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.01, max_it = 500,
       save_path = cp, verbose = FALSE),
  silent = TRUE
)
cat(conditionMessage(attr(bad, "condition")))
```

```{r checkpoint-cleanup, include=FALSE}
unlink(c(cp, paste0(cp, ".manifest")))
```

The manifest records the data, design, window, and every tuning setting that
affects comparability (including anything passed through `...`, such as
`v0`), and names exactly what differs. Resuming by matching `K` alone would
silently pool candidates fit under different regularization -- rows that look
comparable in a table but are not -- so the checkpoint refuses instead.
Delete the checkpoint (or pick a new `save_path`) to start over deliberately.

## Objective-type discipline

`pefa()` does not assume the objective is always the ELBO. For ordinary
(diagonal) `vbfa` sweeps it is the ELBO; for LD-aware sweeps (`ld = TRUE`) it
is the manuscript V4 terminal VECM objective instead, and the ELBO-specific
columns and selections (`elbo`, `elbo_gain`) are `NA` because the residual
precision there is point-updated rather than given a full variational
distribution. `objective_gain` remains the primary rule in both cases, just
computed on whichever objective the candidates actually share. What `pefa()`
will not do is mix the two silently: every candidate in a window is fit with
the same `ld` setting, but if a sweep is ever assembled from candidates whose
`objective_type` differs, `pefa()` refuses with an explicit error rather than
comparing ELBO to a VECM objective as though they were the same scale.

## References

* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection: Fit assessment and factor-number selection in
  partially exploratory factor analysis. *arXiv preprint* arXiv:2607.07159.
* Auerswald, M., & Moshagen, M. (2019). How to determine the number of
  factors to retain in exploratory factor analysis: A comparison of
  extraction methods under realistic conditions. *Psychological Methods*,
  24(4), 468-491.
