---
title: "Factor-number sweeps with pefa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor-number sweeps with pefa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`pefa()` evaluates a factor-number *window*. The returned `pefa` object is also
a `vbpm_sweep`: it contains the candidate `vbpm_fit` objects and their
comparison table, but is not itself a fitted probability model. “Path” is
reserved for within-fit continuation sequences such as `v0` or `xi0`.

```{r sweep}
s <- sim_fa(N = 250, K = 3, ipf = 5, rseed = 19)
Y <- s$dat
Q0 <- matrix(-1L, ncol(Y), 2)
Q0[1:2, ] <- 0L; Q0[1:2, 1] <- 1L
Q0[6:7, ] <- 0L; Q0[6:7, 2] <- 1L

p <- pefa(Q0, Y, Kmin = 2, Kmax = 4, v0 = 0.001,
          verbose = FALSE)
summary(p)
selected_fit(p)
```

The summary reports raw objective, objective-gain, raw ELBO, ELBO-gain, raw
BIC, and BIC-gain selections. For ordinary `vbfa` the objective is the ELBO.
For LD-aware sweeps it is the manuscript V4 VECM objective, so the ELBO fields
are intentionally `NA`. The primary rule is always `objective_gain`; a
boundary result is a prompt to extend the candidate window in that direction.

```{r plots, eval=FALSE}
plot(p, type = "objective")
plot(p, type = "gain")
plot(p, type = "fit")
```
