---
title: "MIMIC models with vbmimic: regularizing measurement and structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MIMIC models with vbmimic: regularizing measurement and structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

A MIMIC (multiple-indicators multiple-causes) model adds a structural part to
the factor model: observed covariates $X$ predict the factors, which in turn
produce the item responses $Y$. `vbmimic()` implements the extended MIMIC model
of Jin & Chen (2025, *Multivariate Behavioral Research*), whose contribution is
to regularize **both** parts: spike-and-slab priors select unspecified entries
of the measurement design `Q_A` (items on factors, `J x K`) *and* of the
structural design `Q_B` (factors on covariates, `K x P`). Both matrices use the
same `-1/0/1` codes as `vbfa()`.

With the default settings, `vbmimic()` reproduces the published estimator
exactly.

## Simulating MIMIC data

`sim_lvm()` (absorbed from the LAWBL package) generates the data. Supply the
structural coefficients as a matrix — a sparse design where each factor is
predicted by its own covariates is both realistic and estimable:

```{r simulate}
B <- matrix(0, 3, 9)
for (k in 1:3) B[k, (k * 3 - 2):(k * 3)] <- .3   # 3 covariates per factor

sim <- sim_lvm(N = 500, K = 3, J = 18, P = 9, b = B, phx = 0, rseed = 1)
Y <- sim$dat[, 1:18]        # items first ...
X <- sim$dat[, 19:27]       # ... covariates last
```

(A scalar `b` would make *every* covariate predict *every* factor, which
quickly leaves no disturbance variance — `sim_lvm()` stops with an informative
error if the design is infeasible.)

## Anchor at least one part

One modelling caution, learned the empirical way: with **both** `Q_A` and
`Q_B` fully exploratory the model converges cleanly but the solution is
rotationally ambiguous — factors can merge or swap. Anchoring either part
resolves it. Here we anchor two items per factor in the measurement part and
leave the whole structural part to the data:

```{r design}
Q_A <- matrix(-1L, 18, 3)
for (k in 1:3) {
  a <- which(rep(1:3, each = 6) == k)[1:2]
  Q_A[a, ] <- 0L
  Q_A[a, k] <- 1L
}
Q_B <- matrix(-1L, 3, 9)     # structural selection is the question
```

## Fit and read

```{r fit}
fit <- vbmimic(Y, X, Q_A, Q_B)
fit
```

The structural question — *which covariates predict which factors?* — is
answered by `B` and its posterior inclusion probabilities `pi_B`:

```{r structural}
round(fit$B, 2)
round(fit$pi_B, 2)
```

Compare against the generating design (up to factor relabeling, which is
inherent to exploratory fits):

```{r recovery}
est_blocks  <- apply(abs(fit$B) > .15, 1, function(r) which(r))
true_blocks <- apply(B != 0, 1, function(r) which(r))
est_blocks
true_blocks
```

The measurement side reads exactly as in `vbfa()`:

```{r measurement}
round(fit$A, 2)[1:6, ]
round(fit$Phi, 2)            # factor correlations, from the disturbances
```

## Options worth knowing

* **`v0` path.** As in `vbfa()`, a decreasing vector gives a warm-started
  regularization path over both parts; the scalar default `0.001` is the
  published fixed spike.
* **`standardize`.** The default `FALSE` is the published estimator's
  behaviour. `vbfa()` by contrast always standardizes internally; set
  `standardize = TRUE` for the analogous behaviour.
* **Determinism.** Like `vbfa()`, the estimator consumes no random numbers —
  no seed argument, bit-identical reruns.
* **No ELBO yet.** The published algorithm converges on a residual criterion,
  not the bound; `fit$ELBO` is `NA`. `pefa()` therefore refuses
  `vbmimic` fits rather than reporting wrong statistics. An implemented ELBO
  is on the roadmap.

## References

* Jin, Y., & Chen, J. (2025). Regularized variational Bayesian approximations
  for variable selection in extended multiple-indicators multiple-causes
  models. *Multivariate Behavioral Research*.
* Ročková, V., & George, E. I. (2018). The spike-and-slab LASSO. *Journal of
  the American Statistical Association*, 113(521), 431–444.
