---
title: "Getting started with vbpm: partially confirmatory factor analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with vbpm: partially confirmatory factor analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(vbpm)
```

`vbpm` estimates psychometric measurement models by regularized mean-field
variational Bayes. Its models are *partially confirmatory*: instead of choosing
between a confirmatory analysis (every loading prespecified) and an exploratory
one (nothing prespecified), you specify what you know and let continuous
spike-and-slab priors select the rest from the data.

This vignette walks the factor-analysis side of the package: designing a `Q`
matrix, fitting with `vbfa()`, reading the fit, computing fit statistics with
`vb_fit()`, selecting the number of factors with `pefa()`, and two special
modes (bifactor and local dependence). It ends with an empirical example and
with a recipe for turning *graded* prior knowledge into a `Q` matrix.

## The design matrix Q

Everything starts with a `J x K` integer matrix with three codes:

| code | meaning |
|------|---------|
| `1`  | **specified** (anchored): the loading is estimated freely, with no spike — you assert this item measures this factor |
| `0`  | **fixed zero**: the loading is constrained to zero — you assert it does not |
| `-1` | **unspecified**: the data decide, via a spike-and-slab prior; the fit reports a posterior inclusion probability (PIP) |

A fully confirmatory model has no `-1`; a fully exploratory one is all `-1`
(plus enough anchors to identify the factors). Everything between is the
partially confirmatory continuum.

## Simulate, design, fit

`sim_fa()` generates factor-analytic data from a loading *pattern*: `K` factors,
`ipf` items per factor, primary loadings `lam`, cross-loadings `lac`.

```{r fit}
sim <- sim_fa(N = 500, K = 3, ipf = 6, lam = .7, lac = .3, rseed = 1)
Y <- sim$dat

## two anchors per factor; everything else left to the data
Q <- matrix(-1L, ncol(Y), 3)
for (k in 1:3) {
  a <- which(rep(1:3, each = 6) == k)[1:2]
  Q[a, ] <- 0L
  Q[a, k] <- 1L
}

fit <- vbfa(Y, Q)
fit
```

Typing the fit's name gives the compact summary above — the object carries an
S3 class (`vbpm_fit`), but it is still an ordinary list underneath and every
component is public API (see `?vbpm_fit`):

```{r access}
round(fit$Lam[1:6, ], 2)   # posterior mean loadings
round(fit$pi[1:6, ], 2)    # PIPs of the unspecified entries
round(fit$Phi, 2)          # factor correlations (oblique by default)
```

A loading is conventionally treated as *active* when its PIP is at least .5.
Compare recovered structure against the truth:

```{r recovery}
active <- (Q == 1) | (Q == -1 & fit$pi >= .5)
table(truth = sim$MLA != 0, active = active)
```

Two properties worth knowing. The estimator is **deterministic** — fixed
initialization, no random numbers — so there is no seed to set and two runs are
bit-identical. And the default `v0` is a decreasing *vector*, a warm-started
regularization path; a scalar `v0 = 0.001` reproduces the fixed-spike estimator
of Jin & Chen (2025) exactly. When a hard model fails to converge, prefer a
gentler, longer path over raising `max_it`:

```{r path, eval = FALSE}
fit_gentle <- vbfa(Y, Q, v0 = exp(seq(log(0.05), log(0.001), length.out = 10)))
```

## Fit statistics

`vb_fit()` computes SEM-like statistics for a fit, with hard selection
(PIP >= `tau`) as the default:

```{r vbfit}
round(vb_fit(fit, Y, Q), 3)
```

`t_nom`/`t`/`t_S` are parameter counts (nominal, rank-adjusted, soft), the rest
are the usual absolute and information indices. The function reads `orthogonal`
and `ld` from the fit object itself, so a bifactor fit needs no extra argument
— and supplying a contradictory one is an error, not a silent miscount.

## How many factors? pefa() and the gain rule

`pefa()` sweeps a window of factor counts, holding your partially confirmatory
backbone fixed and padding with exploratory columns, then selects K with a
scale-free *gain rule*: a factor is worth keeping while its marginal ELBO gain
is at least `delta` percent of the largest gain in the window (with a
sustained-drop guard against transient dips).

```{r pefa}
Q0 <- Q[, 1:2, drop = FALSE]   # pretend we only know two factors exist
r  <- pefa(Q0, Y, Kmin = 2, Kmax = 4, verbose = FALSE)
r$selected_K
r$sweep[, c("K", "ELBO", "BIC", "CFI", "flag")]
```

The rule itself is exported, so you can apply it to any criterion curve:

```{r elbow}
select_K_elbow(K = r$sweep$K, score = r$sweep$ELBO)
```

## Bifactor models

`orthogonal = TRUE` fixes the factor correlation at the identity. With a design
whose first column is a general factor (all items load) plus group columns,
this is an identified orthogonal bifactor model:

```{r bifactor}
Qb <- cbind(1L, Q)
fb <- vbfa(Y, Qb, orthogonal = TRUE)
fb$Phi[1:3, 1:3]           # identity by construction
round(vb_fit(fb, Y, Qb), 3)
```

## Local dependence

Correlated residuals (e.g. testlets, shared stems) are handled by a *graphical*
spike-and-slab prior on the residual precision, solved by QUIC. Turn it on with
`ld = TRUE`; by default the search is fully exploratory, or restrict it with a
`J x J` design `Qe` using the same `-1/0/1` codes.

```{r ld}
simLD <- sim_fa(N = 500, K = 3, ipf = 6, lam = .7, lac = .3, ecr = .3,
                rseed = 2)
fLD <- vbfa(simLD$dat, Q, ld = TRUE, max_it = 300, tolVal = 1e-3)

## the largest recovered residual edges, vs the planted pairs
Poff <- abs(fLD$Psi); Poff[lower.tri(Poff, diag = TRUE)] <- 0
which(Poff >= sort(Poff, decreasing = TRUE)[3], arr.ind = TRUE)
simLD$ofd_ind
```

Two current limitations, both documented in `NEWS`: the ELBO is `NA` under
local dependence, and `vb_fit()` refuses LD fits rather than computing
statistics on the wrong implied covariance.

## An empirical example: NLSY 1997

`nlsy27` ships with the package: 3,458 respondents, 27 mixed-type items, and an
initial three-factor design with two to three anchors per factor.

```{r nlsy}
data(nlsy27)
## vbfa() does not support NA yet -> complete cases; the 17 polytomous items
## are treated as continuous for this illustration
Yn <- as.matrix(nlsy27$dat[stats::complete.cases(nlsy27$dat), ])
dim(Yn)
fn <- vbfa(Yn, nlsy27$Q)
fn
round(fn$Lam, 2)
```

## From graded knowledge to a Q matrix

Prior knowledge is often *graded* rather than crisp — say, a membership score
`S[j, k]` in `[0, 1]` for item `j` on factor `k`, from expert ratings, text
similarity, or a clustering consensus. A practical thresholding rule turns such
scores into a partially confirmatory `Q`:

* **anchor** (`1`) where the evidence is strong: `S > .80`;
* **leave to the data** (`-1`) where it is equivocal: `.20 <= S <= .80`;
* **exclude** (`0`) where it is absent: `S < .20`.

```{r softq}
## a toy graded membership matrix for the simulated items
set.seed(3)
S <- matrix(runif(18 * 3, 0, .2), 18, 3)          # baseline noise
S[cbind(1:18, rep(1:3, each = 6))] <- runif(18, .55, .95)  # true memberships

Qsoft <- matrix(0L, 18, 3)
Qsoft[S >= .20 & S <= .80] <- -1L
Qsoft[S >= .80] <- 1L
table(Qsoft)

fsoft <- vbfa(Y, Qsoft)
fsoft$flag
```

This costs no new machinery — it is simply a disciplined way to build `Q` — and
it uses the graded information honestly: strong evidence anchors, equivocal
evidence is *tested* rather than assumed, absence excludes. (A deeper version,
in which the graded scores enter the prior itself as per-element inclusion
probabilities, is on the package roadmap.)

## References

* Chen, J., Guo, Z., Zhang, L., & Pan, J. (2021). A partially confirmatory
  approach to scale development with the Bayesian Lasso. *Psychological
  Methods*, 26(2), 210–235.
* Chen, J. (2023). Fully and partially exploratory factor analysis with
  bi-level Bayesian regularization. *Behavior Research Methods*, 55(4),
  2125–2142.
* Jin, Y., & Chen, J. (2025). Regularized variational approximation for
  partially confirmatory factor analysis. *Structural Equation Modeling*,
  32(3), 437–449.
* Chen, J., & Jin, Y. (2026). Recovering latent structures after variational
  Bayesian variable selection. *arXiv:2607.07159*.
